% ============================================================
%  ch01_jeszcze_jedan_pogrzeb.tex  —  Chapter 1
%  Jeszcze jeden pogrzeb (pp. 5–30)
% ============================================================
%
% ── STYLE RULES FOR THIS FILE ───────────────────────────────
%
%  These rules must be followed consistently throughout.
%  Drift from these rules has occurred before — check carefully.
%
%  HEADINGS:
%  - Use \section{} for ALL passage headings without exception.
%  - \subsection{} and \subsection*{} are NOT used in this file.
%  - Section titles follow the pattern:
%      Polish title / English title   (e.g. \section{Obchód / The Rounds})
%    or just a name if self-evident   (e.g. \section{Groszek})
%  - \section{} renders in blue bold with a rule beneath
%    (defined in olowiane_dzieci_analiza.tex).
%
%  POLISH SOURCE TEXT:
%  - Always use \poltext{...} — never \begin{quote}...\end{quote}
%    and never plain \textit{...} or any other wrapper.
%  - \poltext{} renders the text in red italic (keywordred).
%
%  ENGLISH TRANSLATION:
%  - Always use \poltrans{...} immediately after \poltext{}.
%  - \poltrans{} renders in blue (polishblue) with a rule beneath.
%  - No label precedes the translation — colour contrast suffices.
%
%  ANALYSIS NOTES:
%  - Follow \poltrans{} with free prose commentary.
%  - Bold Polish terms with \textbf{}, gloss in \textit{}.
%  - Standard pattern: \textbf{term} --- \textit{gloss} --- analysis.
%  - Use \medskip before thematic summary paragraphs.
%  - Grammar or vocabulary notes may be prefixed with:
%    \noindent\textit{Grammar note:} or \noindent\textit{Vocabulary note:}
%
%  PASSAGE SEPARATORS:
%  - Use % ──────────────────────────────────────────────────
%    (single % with em-dashes) between passages within a section.
%    Do NOT use the %% --- style.
%
%  UPDATE PROTOCOL:
%  - The LaTeX file is updated after each passage is confirmed
%    by the user — not in batches.
%  - Always append new content by str_replace on the final line.
%  - Verify line count after each update with wc -l.
%
% ── WHERE THE READING CURRENTLY STANDS ─────────────────────
%
%  Last passage analysed (session ending 2026-03-06):
%  "To nie Gorol" — Hanka corrects Groszek; Gorol has moved
%  away; the men pod sklepem are indistinguishable.
%  See \section{To nie Gorol / That's Not Gorol} at end of file.
%
%  Chapter 1 is in progress. The cemetery obchód continues.
%
% ============================================================


\chapter{Jeszcze jeden pogrzeb}
\markboth{Jeszcze jeden pogrzeb}{}

% ── Chapter title ────────────────────────────────────────────
\section{Tytuł rozdziału / Chapter Title}

\poltext{Jeszcze jeden pogrzeb}
\poltrans{Yet Another Funeral / One More Funeral}

\textbf{Jeszcze} — \textit{still, yet, another}. Implies continuation and 
accumulation, even weariness. \textit{Jeszcze raz} — once more; 
\textit{jeszcze jeden} — yet one more, with the implication that there 
have already been others.

\textbf{jeden} — \textit{one} (masculine nominative, agreeing with \textit{pogrzeb}).

\textbf{pogrzeb} — \textit{funeral} (masculine noun, nominative). 
Root: \textit{grzebać} — to bury, to dig. The earthiness of the root word 
is audible in it.

The devastating efficiency of this title is in \textit{jeszcze}. Not ``a funeral'' 
or even ``another funeral'' — but \textit{yet one more}, the way someone says 
it who has lost count.

% ── First paragraph ──────────────────────────────────────────
\section{Pierwsze akapity / Opening Paragraphs}

\poltext{Umarł Fiszer, kolega z podstawówki. Postać ważna, gdyż jego dziadek, 
stary Fiszer, miał kiosk. Można było tam dostać spod lady komiksy z serii: 
Kapitan Kloss, Kapitan Żbik albo Podziemny Front --- dla nas absolutne rarytasy, 
a także czerwone Carmeny i niebieskie Caro, towar jeszcze bardziej pożądany, 
zważywszy, że były to pierwsze w życiu papierosy palone przez nas w krzakach. 
Już w piątej klasie.}
\poltrans{Fiszer died --- a classmate from primary school. An important figure, 
because his grandfather, old Fiszer, had a kiosk. You could get there under the 
counter comics from the series: Kapitan Kloss, Kapitan Żbik, or Podziemny Front 
--- for us absolute rarities, and also red Carmens and blue Caros, goods even more 
desirable considering that these were the first cigarettes smoked by us in the 
bushes. Already in fifth grade.}

\textbf{Umarł Fiszer} — verb first, subject second. Polish allows this inversion 
freely; the death lands before we know who died.

\textbf{kolega z podstawówki} — \textit{podstawówka} is the colloquial, affectionate 
diminutive of \textit{szkoła podstawowa} (primary school). The suffix \textit{-ka} 
sets the nostalgic, intimate tone immediately.

\textbf{gdyż} — \textit{because} (literary register). Signals deliberate craft. 
See Grammar Foundations §\,Conjunctions for the full \textit{bo / ponieważ / gdyż} 
register spectrum.

\textbf{Można było tam dostać} — \textit{one could get there}. Impersonal 
\textit{można} (one can) + \textit{było} (past auxiliary) + 
\textit{dostać} (to obtain, perfective infinitive).

\textbf{spod lady} — \textit{from under the counter}. \textit{Spod} is a 
compound preposition (from under), taking the genitive. Under-the-counter goods 
in communist-era Poland were a cultural institution.

\textbf{rarytasy} — \textit{rarities, treats}. Plural of \textit{rarytas} — 
an old-fashioned, savoured word.

\textbf{czerwone Carmeny i niebieskie Caro} — Polish cigarette brands identified 
entirely by their packaging colour. \textit{Czerwone} and \textit{niebieskie} are 
plural adjectives in agreement with the brand names.

\textbf{zważywszy, że} — \textit{considering that, given that}. A formal 
participial construction applied with dry irony to children smoking in bushes.

\textbf{palone przez nas w krzakach} — \textit{smoked by us in the bushes}. 
Passive participial construction. \textit{Krzaki} (bushes) — a word that 
needs no translation once heard.

\textbf{Już w piątej klasie.} — \textit{Already in fifth grade.} The fragment 
stands alone. \textit{Już} marks arrival at a milestone — fifth grade, roughly 
age eleven. The author's eyebrow is raised at his own childhood.

% ────────────────────────────────────────────────────────────

\poltext{Fiszer umarł nagle, niespodziewanie i kłopotliwie.}
\poltrans{Fiszer died suddenly, unexpectedly, and inconveniently.}

Three adverbs in escalating syllable count, each adding weight before 
the final, jarring word.

\textbf{nagle} — \textit{suddenly}. Abrupt, no warning.

\textbf{niespodziewanie} — \textit{unexpectedly}. From \textit{spodziewać się} 
(to expect); \textit{nie-} prefix negates it, exactly as English \textit{un-expected}.

\textbf{kłopotliwie} — \textit{inconveniently, troublesomely}. From 
\textit{kłopot} (trouble, bother, hassle). Applied to a man's death, 
this is either bleakly comic or quietly brutal. The author could have 
written \textit{tragicznie} (tragically). He chose \textit{kłopotliwie}. 
That choice announces the voice of the book.

% ────────────────────────────────────────────────────────────

\poltext{Zrozpaczona wdowa musiała transportować jego zwłoki z południa Francji.}
\poltrans{The grief-stricken widow had to transport his remains from the south of France.}

\textbf{Zrozpaczona} — \textit{grief-stricken}. From \textit{rozpacz} (despair). 
The \textit{z-} prefix indicates a state fully arrived at. A strong, unambiguous word.

\textbf{musiała} — \textit{had to, was obliged to}. Past tense of \textit{musieć}. 
Not ``chose to'' or ``arranged to'' but \textit{had to} — the logistics were inescapable.

\textbf{transportować} — borrowed from international vocabulary. The clinical, 
bureaucratic register against \textit{zrozpaczona wdowa} is quietly jarring: 
grief on one side, freight paperwork on the other.

\textbf{zwłoki} — \textit{mortal remains, corpse}. Exists only in the plural 
in Polish, like English ``remains.'' More formal than \textit{trup} (body), 
but unflinching.

\textbf{z południa Francji} — \textit{from the south of France}. 
\textit{Południe} means both \textit{south} and \textit{noon} — 
sharing a root with the sun's position. Genitive follows \textit{z} (from).

The contrast with \textit{kłopotliwie} is now fully clear: the 
``inconvenience'' mentioned so dryly has a very human cost, borne entirely 
by the widow.

% ────────────────────────────────────────────────────────────

\poltext{Zdecydowaliśmy, że dla uczczenia jego pamięci trzeba się napić, 
i z takim postanowieniem udaliśmy się na pogrzeb.}
\poltrans{We decided that to honour his memory one must have a drink, and 
with that resolution we set off for the funeral.}

\textbf{Zdecydowaliśmy} — \textit{we decided}. Perfective past, first person 
plural. A decision fully made.

\textbf{dla uczczenia jego pamięci} — \textit{for the honouring of his memory}. 
Formal, almost ceremonial. \textit{Uczczenie} is a verbal noun from 
\textit{uczcić} (to honour, to commemorate). The genitive chain 
\textit{uczczenia jego pamięci} illustrates how Polish stacks nouns in the genitive.

\textbf{trzeba się napić} — \textit{one must have a drink}. 
\textit{Trzeba} (one must) reappears from the epigraph material. 
\textit{Napić się}: the reflexive \textit{się} with \textit{napić} means 
to have a drink, to drink one's fill — a distinction between an activity 
(\textit{pić}, to drink) and a completed, satisfying act (\textit{napić się}).

\textbf{z takim postanowieniem} — \textit{with such a resolution}. 
Instrumental after \textit{z} (with). \textit{Postanowienie} is a wonderfully 
weighty word for what is essentially ``let's get drunk at the funeral.''

\textbf{udaliśmy się} — \textit{we set off, we betook ourselves}. Slightly 
elevated register for going somewhere. The formality is faintly comic here.

The sentence wraps a decision to drink in the language of solemn civic duty. 
The gap between the gravity of the language and the actual resolution is where 
all the humour lives.

% ────────────────────────────────────────────────────────────

\poltext{Na cmentarzu w Dąbrówce Małej przy świeżo wykopanym grobie czekała 
na kondukt grupka żałobników.}
\poltrans{At the cemetery in Dąbrówka Mała, by the freshly dug grave, a small 
group of mourners waited for the funeral procession.}

\textbf{Na cmentarzu} — \textit{at the cemetery}. \textit{Na} with the locative 
for location. \textit{Cmentarz} — a word worth knowing immediately in Poland, 
where cemeteries are extraordinary cultural spaces.

\textbf{w Dąbrówce Małej} — \textit{in Dąbrówka Mała}. A small district of 
Katowice in Silesia. The place name declines: \textit{Dąbrówka Mała} 
(nominative) → \textit{Dąbrówce Małej} (locative after \textit{w}). 
Both words change their endings in agreement — noun and adjective declining 
together, exactly as in Latin.

\textbf{przy świeżo wykopanym grobie} — \textit{by the freshly dug grave}. 
\textit{Przy} takes the locative. \textit{Świeżo} (adverb: freshly); 
\textit{wykopanym} (past passive participle, locative: dug). Note the echo 
of \textit{Groby są świeże} from the epigraph — the author has made good 
on his own promise.

\textbf{grupka żałobników} — \textit{a small group of mourners}. 
\textit{Grupka} — diminutive of \textit{grupa}, the \textit{-ka} suffix 
quietly measuring how few people are present. \textit{Żałobników} is the 
genitive plural of \textit{żałobnik} (mourner), required after \textit{grupka}.

The sentence moves the camera slowly inward: location → specific location 
→ the grave → the waiting → the handful of people.

% ────────────────────────────────────────────────────────────

\poltext{Za nimi, przy głównej alejce, ustawili się dyskretnie grabarze.}
\poltrans{Behind them, by the main path, the gravediggers had discreetly 
taken their positions.}

\textbf{Za nimi} — \textit{behind them}. \textit{Za} with the instrumental 
for position. \textit{Nimi} is the instrumental form of \textit{oni/one} (they).

\textbf{przy głównej alejce} — \textit{by the main path}. 
\textit{Alejka} is a diminutive of \textit{aleja} — a little lane between graves.

\textbf{ustawili się} — \textit{positioned themselves}. 
\textit{Ustawić się}: reflexive perfective — they have arranged themselves 
and are settled.

\textbf{dyskretnie} — \textit{discreetly}. From French/Latin \textit{discret}. 
Its position between verb and subject gives it quiet emphasis.

\textbf{grabarze} — \textit{gravediggers}. From \textit{grób} (grave) — 
the same root recurring throughout the chapter. Placed last in the sentence: 
the most concrete image saved for the end.

The discreet gravediggers waiting in the wings carry an unspoken 
practicality that echoes the earlier \textit{kłopotliwie}. Everyone 
is performing their assigned role.

% ────────────────────────────────────────────────────────────

\poltext{Wczesną jesienią, ciepłą i kolorową, pasującą bardziej do spaceru 
z psem lub grzybobrania niż do śmierci, na cmentarzu parafialnym kościoła 
pod wezwaniem Świętego Antoniego, patrona między innymi od spraw beznadziejnych, 
po prawie czterdziestu latach spotkałem szkolne koleżanki i kolegów z podstawówki.}
\poltrans{In the early autumn, warm and colourful, more suited to a walk with 
a dog or mushroom picking than to death, at the parish cemetery of the church 
under the patronage of Saint Anthony, patron among other things of hopeless 
causes, after nearly forty years I met my schoolgirls and schoolboys from 
primary school.}

\textbf{Wczesną jesienią} — \textit{in early autumn}. Instrumental case used 
for time periods. Both words decline together.

\textbf{pasującą bardziej do\ldots niż do} — \textit{more suited to\ldots than to}. 
\textit{Bardziej\ldots niż} is the standard Polish comparative construction. 
\textit{Pasować do} (to suit, to fit) always takes the genitive.

\textbf{grzybobrania} — \textit{mushroom picking}. A compound: 
\textit{grzyb} (mushroom) + \textit{branie} (taking/picking). 
Mushroom foraging is a deeply serious Polish cultural institution, and its 
appearance here as the opposite of death is both funny and characteristically Polish.

\textbf{pod wezwaniem Świętego Antoniego} — \textit{under the patronage of 
Saint Anthony}. Literally ``under the invocation of.'' The fixed Polish phrase 
for a church's dedicatory patron.

\textbf{patrona między innymi od spraw beznadziejnych} — \textit{patron among 
other things of hopeless causes}. Saint Anthony is the patron of lost things 
and hopeless cases. \textit{Między innymi} (among other things) is used 
constantly in Polish. \textit{Beznadziejnych}: from \textit{nadzieja} (hope), 
negated by \textit{bez-} (without).

\textbf{po prawie czterdziestu latach} — \textit{after nearly forty years}. 
\textit{Po} with the locative for ``after.'' \textit{Czterdziestu} is the 
genitive plural of \textit{czterdzieści} (forty), required after \textit{prawie} (almost).

\textbf{spotkałem szkolne koleżanki i kolegów} — \textit{I met schoolgirls 
and schoolboys}. First appearance of a first-person singular verb in the 
narrative. \textit{Spotkałem}: perfective past, masculine singular. Note that 
Polish distinguishes \textit{koleżanka} (female classmate) and \textit{kolega} 
(male classmate) — both are needed to include everyone.

The sentence delays the main verb \textit{spotkałem} through layer after layer 
of atmospheric detail. The weather is wrong for a funeral. The church's own 
patron saint specialises in hopeless causes. By the time we arrive at the actual 
event, it has been framed as something faintly absurd, faintly doomed, and entirely human.

% ────────────────────────────────────────────────────────────

\poltext{Oprócz garstki obecnych wspomnijmy też całą resztę: nieobecnych lub martwych.}
\poltrans{Besides the handful of those present, let us also remember all the rest: 
the absent or the dead.}

\poltext{Najpierw my, obecni.}
\poltrans{First, us --- the ones who are here.}

\textbf{Oprócz} — \textit{besides, apart from}. Takes the genitive.

\textbf{garstki} — genitive singular of \textit{garstka} — a handful, literally 
the amount that fits in a cupped hand. A second diminutive quietly measuring 
how few remain.

\textbf{wspomnijmy} — \textit{let us remember}. First person plural imperative 
of \textit{wspomnieć}. The \textit{-my} ending is the inclusive ``let us'' form. 
The author pulls the reader into the obligation.

\textbf{całą resztę} — \textit{all the rest}. Accusative after \textit{wspomnijmy}.

\textbf{nieobecnych lub martwych} — \textit{the absent or the dead}. Genitive 
plural in apposition to \textit{resztę}. The colon before this phrase makes it 
land as a definition: the rest = the absent or the dead. Only two categories remain.

\textbf{Najpierw my, obecni.} — \textit{First, us --- the present ones.} 
No verb. \textit{My} (we/us) stated explicitly is emphatic in Polish, where 
the subject is normally encoded in the verb ending alone. Saying \textit{my} 
means: \textit{us, specifically, as distinct from them.}

% ────────────────────────────────────────────────────────────

\poltext{Zbliżyliśmy się do pozostałej grupki stojącej przy grobie.}
\poltrans{We drew closer to the remaining little group standing by the grave.}

\textbf{Zbliżyliśmy się} — \textit{we drew closer}. From \textit{zbliżyć się}, 
the perfective reflexive of \textit{bliski} (close). A gentle, hesitant motion 
— not \textit{podeszliśmy} (we walked up to) but a gradual closing of distance.

\textbf{do pozostałej grupki} — \textit{to the remaining little group}. 
\textit{Pozostałej}: genitive of \textit{pozostały} (remaining, left over) 
— a quietly melancholy adjective. \textit{Grupki}: the same diminutive 
as before, insisting again on their fewness.

\textbf{stojącej przy grobie} — \textit{standing by the grave}. 
Present active participle, genitive singular agreeing with \textit{grupki}.

After the rhetorical force of \textit{Najpierw my, obecni}, the narrative 
simply resumes. The mundane physical action after the weightier reflection 
is characteristic of this author's rhythm.

% ── Groszek's entrance ───────────────────────────────────────
\section{Groszek}

\poltext{--- Coś nie za dużo nas --- powiedział Groszek, odkręcając nakrętkę.}
\poltrans{`Not too many of us, is there,' said Groszek, unscrewing the cap.}

\textbf{Coś nie za dużo nas} — \textit{colloquial and understated}. \textit{Coś} 
here is not ``something'' but a softening particle — a shrug built into the grammar. 
\textit{Nie za dużo}: not too many. The whole phrase is casual, almost throwaway, 
for an observation about how many of their childhood friends are still alive.

\textbf{Groszek} — literally \textit{little pea} or \textit{little coin} 
(diminutive of \textit{grosz}, a small coin; or \textit{groch}, pea). 
A childhood nickname that has clearly followed someone into late middle age.

\textbf{odkręcając nakrętkę} — \textit{unscrewing the cap}. 
Present adverbial participle: simultaneous action. \textit{Odkręcać} 
(to unscrew, to twist open) and \textit{nakrętka} (cap, lid, screw-top) 
share the root \textit{kręcić} (to turn, to twist). He is already opening 
the bottle while speaking.

% ────────────────────────────────────────────────────────────

\poltext{--- Zostaw tę flaszkę!}
\poltrans{`Put that bottle down!'}

\textbf{Zostaw} — \textit{leave it, put it down}. Imperative singular of 
\textit{zostawić}. Sharp, direct, unambiguous.

\textbf{tę} — accusative singular feminine of the demonstrative \textit{ta} 
(this/that). Polish demonstratives decline fully; \textit{tę} is a form 
worth recognising on sight.

\textbf{flaszkę} — accusative singular of \textit{flaszka}. A colloquial, 
slightly affectionate word for a bottle — more informal than \textit{butelka}. 
The diminutive flavour makes the outrage slightly comic.

% ────────────────────────────────────────────────────────────

\poltext{Groszek, z wykształcenia inżynier, z przekonania śląski autonomista 
i w mediach społecznościowych radykalny zwolennik teorii o wyższości silników 
diesla nad elektrycznymi, posłuchał chóralnej sugestii. Na razie.}
\poltrans{Groszek, an engineer by training, a Silesian autonomist by conviction, 
and on social media a radical proponent of the theory of the superiority of 
diesel engines over electric ones, heeded the collective suggestion. For now.}

\textbf{z wykształcenia\ldots z przekonania} — \textit{by education\ldots by conviction}. 
A fixed Polish construction for introducing someone's professional identity 
and worldview simultaneously. Clean and elegant.

\textbf{śląski autonomista} — \textit{Silesian autonomist}. Silesian regionalism 
and greater autonomy from Warsaw is a genuine, lively political position. 
This detail tells you immediately who Groszek is and where he comes from.

\textbf{w mediach społecznościowych} — \textit{on social media}. 
\textit{Media społecznościowe} is the standard Polish term, a calque of 
``social media.'' Locative plural after \textit{w}.

\textbf{radykalny zwolennik teorii o wyższości\ldots nad\ldots} — 
\textit{a radical proponent of the theory of the superiority of\ldots over\ldots} 
\textit{O wyższości}: from \textit{wysoki} (high/superior); \textit{nad} 
with the instrumental expresses ``over, above.'' The specificity of the 
diesel-over-electric conviction is perfect character comedy: not just an 
opinion, but a \textit{theory}, held \textit{radically}, prosecuted on social media.

\textbf{posłuchał chóralnej sugestii} — \textit{heeded the collective suggestion}. 
\textit{Posłuchać} (to listen to, to obey) takes the genitive. 
\textit{Chóralnej}: \textit{choral, unanimous, in chorus} — from \textit{chór} 
(choir). Everyone told him to put the bottle down at once; the word chosen 
is musical, theatrical.

\textbf{Na razie.} — \textit{For now.} Two words doing enormous work. 
He complied — but only provisionally. \textit{Na razie} is one of the most 
useful phrases in Polish: ``for the time being.'' Here it functions as both 
narrative aside and character revelation. Groszek has been introduced, 
assessed, and shown to be entirely himself in one sentence.

% ── Social anxiety and the bottle ───────────────────────────
\section{Chyba nie powinno się pić}

\poltext{Chyba nie powinno się pić na cmentarzu, przynajmniej w tym kraju 
i w tym obrządku. Nawet w dobrej intencji.}
\poltrans{One probably shouldn't drink at a cemetery, at least in this country 
and in this rite. Even with good intentions.}

\textbf{Chyba} --- \textit{probably, I suppose}. Softens the observation into 
a reasonable social reflection rather than a scolding. See Grammar Foundations.

\textbf{nie powinno się pić} --- \textit{one shouldn't drink}. \textbf{Powinno} 
is the neuter form of \textit{powinien} (should) --- neuter because there is no 
named subject. \textbf{Się} makes it impersonal: the mild reproach is distributed 
across everyone present, including the speaker.

\textbf{przynajmniej} --- \textit{at least}. High-frequency hedging word.

\textbf{obrządek} --- \textit{rite, religious observance} --- specifically 
religious and formal. The speaker acknowledges the rule might differ elsewhere: 
other countries, other funerary customs. A notably cosmopolitan hedge.

\textbf{Nawet w dobrej intencji.} --- \textit{Even with good intentions.} 
A verbless fragment. \textbf{Nawet} (\textit{even}) is a high-frequency particle. 
Even granting honourable intent, it still probably shouldn't happen. Probably.

% ── Warsaw ──────────────────────────────────────────────────
\section{Słoik z Warszawy}

\poltext{Poza tym, co powiedzą nasi dawno niewidziani znajomi, jak nas 
zobaczą\ldots? Pewno nic. Zestarzeliśmy się. To będzie największą atrakcją.}
\poltrans{Besides, what will our long-unseen acquaintances say when they 
see us\ldots? Probably nothing. We've grown old. That will be the greatest attraction.}

\textbf{Poza tym} --- \textit{besides, apart from that} --- conversational 
connector, slightly more colloquial than \textit{oprócz}.

\textbf{dawno niewidziani znajomi} --- \textit{long-unseen acquaintances}. 
\textbf{Dawno niewidziani}: \textit{dawno} (long ago) modifying the past 
passive participle \textit{niewidziany}. \textbf{Znajomi} --- warmer than 
\textit{obcy} (strangers), cooler than \textit{przyjaciele} (friends).

\textbf{Pewno nic.} --- \textit{Probably nothing.} Colloquial variant of 
\textit{pewnie}. One devastating word in answer to the question just posed.

\textbf{Zestarzeliśmy się.} --- \textit{We've grown old.} The prefix \textbf{ze-} 
perfectivises \textit{starzeć się}. Stated flatly, without self-pity.

\textbf{największą atrakcją} --- superlative instrumental, as predicates after 
\textit{być} take the instrumental. The spectacle of ageing classmates will 
be the main event --- not the funeral itself.

% ────────────────────────────────────────────────────────────

\poltext{Jako „słoik" przybyły z Warszawy byłem obiektem szczególnego zainteresowania.}
\poltrans{As a `jar' who had come from Warsaw, I was an object of particular interest.}

\textbf{słoik} --- Warsaw slang for someone who has moved to the capital from 
the provinces, stereotypically arriving each Monday with a jar of home-cooked 
food. Applied to himself with ironic self-awareness: he has returned to Silesia 
as the outsider.

\textbf{obiektem szczególnego zainteresowania} --- \textit{an object of particular 
interest}: instrumental after \textit{być}, genitive chain following.

% ────────────────────────────────────────────────────────────

\poltext{Nic tak nie pobudza do, nazwijmy to, wyrażania zdania odrębnego jak Warszawa.}
\poltrans{Nothing provokes, let us call it, the expression of a dissenting 
opinion quite like Warsaw.}

\textbf{nazwijmy to} --- \textit{let us call it} --- first person plural imperative, 
a parenthetical flagging that what follows is a deliberate euphemism.

\textbf{wyrażania zdania odrębnego} --- \textit{the expression of a dissenting opinion}. 
\textbf{Odrębny} --- \textit{separate, distinct, dissenting} --- formal, almost 
legalistic. The whole phrase wraps strong regional resentment in mock-bureaucratic diction.

% ────────────────────────────────────────────────────────────

\poltext{Łączy całą Polskę w wyrażaniu jednolitego, solidarnego, wspólnego 
ponad podziałami uczucia nienawiści.}
\poltrans{It unites all of Poland in the expression of a uniform, solidary, 
shared-across-divisions feeling of hatred.}

\textbf{ponad podziałami} --- \textit{above/across divisions} --- lifted 
directly from Polish political reconciliation rhetoric, deployed with lethal 
irony. The sentence is built in the exact syntactic shape of a patriotic 
proclamation and delivers \textit{hatred} as the payload.

% ── The ceremony approaches ──────────────────────────────────
\section{Zaraz się zacznie}

\poltext{Zaraz się zacznie. Nadchodzi moja młodsza siostra z koleżanką i kolegą.}
\poltrans{It's about to start. My younger sister is approaching with a female 
friend and a male friend.}

\textbf{Zaraz} --- one of the essential Polish time words: \textit{in a moment, 
right away, any second now}. After three sentences of sardonic Warsaw commentary, 
the narrative snaps back to the immediate present.

\textbf{Nadchodzi} --- imperfective present of \textit{nadchodzić}: the prefix 
\textit{nad-} carries a sense of arrival from a distance.

% ────────────────────────────────────────────────────────────

\poltext{Wcześniej zdążyłem dowiedzieć się, że koleżanka jest świeżo po 
rozwodzie, a kolegę zobaczę o kulach.}
\poltrans{Earlier I had managed to find out that the female friend is freshly 
divorced, and that I would see the male friend on crutches.}

\textbf{zdążyłem} --- perfective past of \textit{zdążyć}: doing something 
just in time, managing to accomplish something within a window of opportunity.

\textbf{świeżo po rozwodzie} --- \textit{freshly divorced}. \textbf{Świeżo}: 
the same root as \textit{świeżo wykopanym grobie} from the opening --- 
applied here to a divorce rather than a grave, with no less bluntness.

\textbf{o kulach} --- \textit{on crutches} --- \textit{o} with the locative 
in this fixed expression.

% ── Dialogue: greetings ──────────────────────────────────────
\section{Powitania / Greetings}

\poltext{--- Cześć, siostra --- przywitałem się z jedyną z tej grupy osobą, 
z którą miałem kontakt przez te lata.}
\poltrans{`Hi, sis,' I greeted the only person from this group with whom 
I had been in contact over these years.}

\textbf{przez te lata} --- \textit{over these years} --- \textit{przez} with 
the accusative for duration. The narration quietly reveals that of everyone 
here, his sister is the only one he stayed in contact with.

% ────────────────────────────────────────────────────────────

\poltext{--- Cześć, brat.}
\poltrans{`Hi, bro.'}

She mirrors his greeting exactly. Two words --- and the warmth lands after 
everything that preceded: the funeral, old age, crutches, divorce, Warsaw 
resentment, a bottle at a graveside. The author knows when to stop.

% ────────────────────────────────────────────────────────────

\poltext{--- Cześć, Daisy. Cześć, Hanka --- Groszek też się przywitał.}
\poltrans{`Hi, Daisy. Hi, Hanka,' --- Groszek greeted them too.}

\textbf{Daisy} sits unglossed beside \textit{Hanka} --- a standard Polish 
diminutive --- without comment. Whether by birth, adoption, or some story 
not yet told, someone in this childhood group from the Katowice suburbs 
is called Daisy. Groszek, meanwhile, is performing social normalcy. 
\textit{Na razie} is still in effect.

% ── Middle age ───────────────────────────────────────────────
\section{Mielizna średniego wieku / The Sandbank of Middle Age}

\poltext{Jeżeli najważniejsze jest pierwsze wrażenie, a --- pocieszmy się --- 
nie jest, to powinniśmy być w szoku, oceniając nasz aktualny wygląd.}
\poltrans{If first impressions are what matter most --- and, let us console 
ourselves, they are not --- then we should be in shock, assessing our current appearance.}

\textbf{pocieszmy się} --- \textit{let us console ourselves} --- first person 
plural imperative, a parenthetical dropped into the conditional. The consolation 
is bleak: first impressions don't matter. Small mercies.

\textbf{aktualny} --- a Polish false friend: does not mean \textit{actual} 
in the English sense but \textit{current, as things stand now}.

% ────────────────────────────────────────────────────────────

\poltext{Czas zatarł podział na trochę starszych i trochę młodszych, wyrzucając 
nas wszystkich na tak kiedyś odległą mieliznę średniego wieku.}
\poltrans{Time has erased the distinction between the slightly older and the 
slightly younger, throwing us all onto the sandbank of middle age, once so distant.}

\textbf{zatarł} --- perfective past of \textit{zatrzeć} (to erase, to obliterate) 
--- a total erasure, not a diminishing.

\textbf{wyrzucając nas wszystkich} --- \textit{throwing us all} --- present 
adverbial participle: not guiding but \textit{throwing}, as a wave throws 
something onto a beach.

\textbf{mielizna} --- \textit{sandbank, shoal} --- a nautical word for shallow 
water where boats run aground. Not a destination chosen but a place where you 
strand. \textbf{Tak kiedyś odległą}: the thing that seemed impossibly far 
away turned out to be right here.

% ────────────────────────────────────────────────────────────

\poltext{Dalej dotarł tylko Fiszer.}
\poltrans{Only Fiszer made it further.}

\textbf{Dalej} --- \textit{further, beyond} --- past the sandbank, which is 
to say, into death. Subject last, the heaviest word at the end. The sandbank 
metaphor inverted in four words: those stranded on the shallows are the 
fortunate ones. Perhaps the finest sentence in the chapter.

% ────────────────────────────────────────────────────────────

\poltext{Gdzie są te niegdysiejsze loki, smukłości i gładkości?!}
\poltrans{Where are those erstwhile curls, slender figures, and smooth complexions?!}

\textbf{niegdysiejsze} --- \textit{erstwhile, of former times} --- archaic and 
literary, from \textit{niegdyś} (once, in days gone by). Applied to curls and 
smooth skin at a funeral: simultaneously grandiose and comic.

The ?! combines question and cry. After the quiet devastation of 
\textit{Dalej dotarł tylko Fiszer}, the author suddenly throws his hands up.

% ────────────────────────────────────────────────────────────

\poltext{Dziewczynki i chłopcy zmienili się najpierw w panie i panów, by potem 
przeistoczyć się w\ldots kogoś takiego.}
\poltrans{The little girls and boys turned first into ladies and gentlemen, 
only to then transform into\ldots someone like this.}

\textbf{przeistoczyć się} --- stronger than \textit{zmienić się}: the prefix 
\textbf{prze-} signals a thorough, complete transformation, passing through 
into something else entirely.

\textbf{w\ldots kogoś takiego} --- the ellipsis and vague demonstrative do 
everything. The first transformation is named with dignity; the second trails 
into a gesture at the actual bodies at the actual funeral. The author cannot 
find a word for what they have become.

% ────────────────────────────────────────────────────────────

\poltext{Tylko po co tyle fałdek, zakoli i podbródków?}
\poltrans{But why so many folds, bald patches, and double chins?}

\textbf{Tylko po co} --- \textit{but why, what's the point of} --- colloquial 
and resigned, more weary than \textit{dlaczego}.

\textbf{zakoli} --- genitive plural of \textit{zakole}: primarily a river bend, 
used colloquially for receding hairlines. The geographical metaphor is quietly apt.

The diminutives (\textit{fałdek}, \textit{zakoli}, \textit{podbródków}) give 
the inventory rueful tenderness rather than pure horror.

% ────────────────────────────────────────────────────────────

\poltext{Protez, licówek, okularów?}
\poltrans{Of dentures, veneers, glasses?}

The list continues and accelerates: from soft diminutives into harder material. 
Dental hardware, cosmetic repairs, corrective lenses. The body is not just 
ageing but being actively maintained, patched, propped up.

% ────────────────────────────────────────────────────────────

\poltext{Jeżeli oni pomyśleli o mnie to, co ja o nich, to są z nich świnie!}
\poltrans{If they thought about me what I thought about them, then they are swine!}

\textbf{to są z nich świnie} --- \textit{then they are swine} --- colloquial 
construction, literally \textit{there are swine from them}. The logic is 
airtight and entirely self-implicating: the \textit{świnie} in question is 
inescapably also the author.

% ────────────────────────────────────────────────────────────

\poltext{Po chwili oczywiście nam przeszło, ale pierwsze wrażenie\ldots}
\poltrans{After a moment it of course passed, but the first impression\ldots}

\textbf{chwila} --- one of the most common time words in Polish: 
\textit{za chwilę} (in a moment), \textit{przez chwilę} (for a moment), 
\textit{po chwili} (after a moment).

\textbf{nam przeszło} --- impersonal construction, the dative \textbf{nam} 
indicating who experienced the passing. Not \textit{we stopped feeling it} 
but \textit{it passed for us}.

The ellipsis refuses to complete the thought because the thought completes 
itself. The three dots are funnier than any conclusion could be.

% ── Recognition ─────────────────────────────────────────────
\section{Rozpoznanie / Recognition}

\poltext{Hanka? Waldi? To naprawdę wy? A ta starsza pani to kto? Nie, pomyłka, 
ta pani to nie od nas --- z innego pogrzebu.}
\poltrans{Hanka? Waldi? Is it really you? And who is that older lady? No, my 
mistake, that lady isn't one of ours --- she's from a different funeral.}

\textbf{ta pani to nie od nas} --- \textit{that lady isn't one of ours} --- 
\textbf{od nas} literally \textit{from us}, meaning \textit{one of our group}.

\textbf{z innego pogrzebu} --- \textit{from a different funeral} --- the 
chapter title (\textit{Jeszcze jeden pogrzeb}) lands quietly here. They 
have aged so much that a classmate is momentarily indistinguishable from 
the elderly mourners at the next funeral over.

% ────────────────────────────────────────────────────────────

\poltext{--- No, ciebie to już dawno nie widziałam. Bardzo dawno\ldots --- 
Hanka przyglądała mi się badawczo.}
\poltrans{`Well, I haven't seen you in a long time. A very long time\ldots' 
--- Hanka was scrutinising me searchingly.}

\textbf{No} --- untranslatable particle: somewhere between \textit{well}, 
\textit{so}, and \textit{right then}. Ubiquitous in Polish conversation.

\textbf{widziałam} --- feminine ending confirms it is Hanka speaking.

\textbf{przyglądała mi się badawczo} --- \textit{was scrutinising me searchingly}. 
\textbf{Przyglądać się}: imperfective, a sustained deliberate examination. 
\textbf{Badawczo} --- from \textbf{badać} (to examine, to investigate) --- 
slightly clinical, investigative. The scrutiny is mutual: the author has been 
inventorying everyone's deterioration and is now on the receiving end of exactly 
the same examination.

% ────────────────────────────────────────────────────────────

\poltext{--- Cześć. Chciałaś może powiedzieć, że nic się nie zmieniłem?}
\poltrans{`Hi. Did you perhaps mean to say that I haven't changed at all?'}

\textbf{Chciałaś może powiedzieć} --- \textbf{chciałaś}: past feminine of 
\textit{chcieć}, confirming he addresses Hanka. \textbf{Może} softens the 
question into a tease rather than a demand.

\textbf{nic się nie zmieniłem} --- double negation (\textit{nic\ldots nie}) 
as Polish requires. He is fishing for a compliment, and he knows it. After 
Hanka's long \textit{badawczo} scrutiny he pre-empts whatever she was about 
to say by supplying the most flattering possible interpretation. The gap 
between what he hopes she might say and what her gaze actually found is 
the whole comedy of the moment.

% ────────────────────────────────────────────────────────────

\poltext{--- Nie. Nie chciałam.}
\poltrans{`No. I didn't mean to say that.'}

Two words, then two words. Hanka doesn't elaborate, doesn't soften it,
doesn't laugh. She just answers the question he asked.

\textbf{Nie chciałam} --- feminine past of \textit{chcieć} (to want).
The \textit{-am} ending confirms unambiguously that Hanka is speaking,
and makes the denial entirely personal and direct. Not \textit{nie,
pomyłka} (no, my mistake) or \textit{no, trochę się zmieniłeś} (well,
you've changed a little) --- just the flat, complete negation of the
thing he hoped she'd say.

The structure mirrors his line exactly: he asked \textit{chciałaś może
powiedzieć} --- she answers \textit{nie chciałam}. He handed her the
script; she handed it back. The comedy is in the ruthlessness of the
economy.

% ────────────────────────────────────────────────────────────

\poltext{--- To teraz będziesz widywać mnie częściej. Wybieram się do 
ciebie po szafy i kuchnię. Urządzam się --- oznajmiłem Hance, która była 
lokalną bizneswoman w branży meblowej.}
\poltrans{`So now you'll be seeing me more often. I'm heading to you for 
wardrobes and a kitchen. I'm setting myself up,' I informed Hanka, who 
was a local businesswoman in the furniture trade.}

\textbf{To} --- the conversational pivot particle, not the pronoun
\textit{to} (that). Equivalent to \textit{so then} or \textit{well
then} --- he takes Hanka's brutal honesty and immediately turns it into
an opportunity.

\textbf{będziesz widywać} --- future tense with imperfective infinitive
\textit{widywać}. Not \textit{widzieć} (to see, a single instance) but
\textit{widywać} --- the habitual/frequentative imperfective, implying
regular, recurring sightings. He is not saying \textit{you'll see me};
he is saying \textit{you'll be seeing me, regularly, as a pattern.}

\textbf{częściej} --- comparative of \textit{często} (often):
\textit{more often}. Confirms the habitual sense of \textit{widywać}.

\textbf{Wybieram się} --- \textit{I'm heading, I'm making my way}. From
\textit{wybierać się} --- to be setting out, to be planning a trip.
Slightly effortful or deliberate; not just \textit{idę} (I'm going) but
\textit{I'm making the trip to you}.

\textbf{po szafy i kuchnię} --- \textit{for wardrobes and a kitchen}.
This is the \textit{po} of purpose: the preposition meaning \textit{in
order to fetch/get}. You go \textit{po chleb} to the bakery,
\textit{po dzieci} to the school. Here: to collect wardrobes and a
kitchen. The items are named with breezy confidence, as though Hanka's
shop is simply his personal furniture depot.

\textbf{Urządzam się} --- \textit{I'm setting up / furnishing / sorting
myself out}. From \textit{urządzić się} --- to get oneself installed,
to set up house. The reflexive \textit{się} makes it personal and
ongoing.

\textbf{oznajmiłem} --- \textit{I informed, I announced}. From
\textit{oznajmić} --- a formal, declarative verb. Not
\textit{powiedziałem} (I said) but something closer to \textit{I
notified, I made known}. Applied to furniture shopping plans announced
at a funeral, it has the same comic register as \textit{z takim
postanowieniem udaliśmy się na pogrzeb} earlier: language of civic
proclamation wrapped around mundane errands.

\textbf{w branży meblowej} --- \textit{in the furniture trade/sector}.
\textit{Branża} (branch, sector, industry) is a thoroughly modern
business word. The narrator explains in passing exactly why he pivoted
from \textit{Nie chciałam} to furniture shopping in under one second.

% ── Kolejny kolega / Another schoolmate ─────────────────────
\section{Kolejny kolega / Another Schoolmate}

\poltext{Witamy się z kolejnym kolegą ze szkoły. To szef Daisy w 
przychodni stomatologicznej. Chodzili razem do klasy.}
\poltrans{We greet another schoolmate. This is Daisy's boss at a dental 
clinic. They were in the same class together.}

\textbf{Witamy się} --- \textit{we greet each other} --- first person
plural reflexive, the same \textit{my} (we) as throughout: the author
and his old classmates as a collective.

\textbf{kolejnym} --- \textit{another, the next one} --- from
\textit{kolej} (turn, sequence): yet one more in an arriving series.
The funeral keeps filling up.

\textbf{przychodnia stomatologiczna} --- \textit{dental clinic}.
\textit{Przychodnia} is an outpatient clinic or medical practice.
\textit{Stomatologiczna} --- adjectival form of \textit{stomatologia}
(dentistry), the formal Polish word, from Greek roots.

\textbf{Chodzili razem do klasy} --- \textit{they used to go to class
together}. \textbf{Chodzić do klasy} --- literally \textit{to walk to
class} --- is the standard expression for attending school together.
The imperfective past \textit{chodzili} gives it its habitual,
long-ago quality: not a single event but an entire shared childhood,
compressed into one verb.

% ────────────────────────────────────────────────────────────

\poltext{--- Cześć, Waldi. Co ci się stało w nogę?}
\poltrans{`Hi, Waldi. What happened to your leg?'}

\textbf{Co ci się stało} --- \textit{what happened to you}. The dative
\textbf{ci} (\textit{to you}) makes it personal: not \textit{what
happened} in the abstract but \textit{what befell you}. Polish uses
this dative-of-the-affected-person naturally where English simply says
``what happened to your leg'' --- but the Polish encodes a shade of
sympathy or concern that English elides.

\textbf{w nogę} --- accusative of \textit{noga} after \textit{w} ---
the accusative with \textit{w} expresses direction into something,
motion toward a target. Not \textit{w nodze} (locative: \textit{in the
leg}, where something is located) but \textit{w nogę} --- something
went \textit{into} the leg, struck it. The construction is idiomatic
for injuries: \textit{dostał w nogę} --- \textit{he got it in the leg}.

The question tells us everything. After all the inventory of folds and
receding hairlines and dental hardware, here is something more acute:
one of them is visibly injured. The question is asked directly, without
preamble, the way old friends ask things.

% ────────────────────────────────────────────────────────────

\poltext{--- Chłop z kryką? Nie za wcześnie? --- dodaje Groszek.}
\poltrans{`A bloke with a crutch? Isn't it a bit early for that?' --- 
Groszek adds.}

\textbf{Chłop} --- \textit{bloke, fellow, man}. Colloquial and earthy
--- not the neutral \textit{mężczyzna} but something closer to
\textit{guy, chap}. Here it has a gruff, joshing quality: \textit{a
grown man, on a crutch?}

\textbf{kryką} --- instrumental of \textit{kryka} --- \textit{crutch},
specifically an underarm crutch. The bluntness of the phrasing ---
subject-plus-instrument, no verb --- reads like a double-take.

\textbf{Nie za wcześnie?} --- \textit{Isn't it a bit early? Too soon?}
\textbf{Za} is the \textit{too} of excess: \textit{za dużo} (too
much), \textit{za późno} (too late), \textit{za wcześnie} (too early).
The affectionate cruelty is precise: you're not old enough yet to need
that. The implication being, of course, that the day will come for all
of them.

\textbf{Groszek} --- literally \textit{little pea}: a childhood
nickname, carried forward unchanged. The diminutive suffix
\textit{-ek} is everywhere in Polish nicknames; \textit{groch} (pea)
becomes \textit{groszek}. That this person is still called Little Pea
at a funeral in middle age is quietly perfect.

% ── Serwis pogwarancyjny / Post-warranty Servicing ──────────
\section{Serwis pogwarancyjny / Post-warranty Servicing}

\poltext{--- Endoproteza. W tym wieku jest już potrzebny serwis 
pogwarancyjny. Niedawno była operacja, jeszcze trochę pochodzę o kulach.}
\poltrans{`A joint replacement. At this age you already need post-warranty 
servicing. There was an operation recently, I'll be walking on crutches a 
little while longer.'}

\textbf{Endoproteza} --- \textit{endoprosthesis, joint replacement}: an
artificial joint, typically hip or knee. Dropped alone, without a verb,
as a complete explanation. One word, clinical, total.

\textbf{W tym wieku} --- \textit{at this age} --- echoing Hanka's
\textit{w twoim wieku} from moments earlier. The phrase is becoming a
quiet refrain throughout this gathering.

\textbf{serwis pogwarancyjny} --- \textit{post-warranty servicing}: the
maintenance required once the manufacturer's guarantee has expired.
Entirely a term from consumer electronics and appliances. Waldi has
reframed his own body as a piece of machinery that has outlived its
warranty period and now requires paid maintenance. \textbf{Już} adds
the familiar rueful note: the milestone has been reached, perhaps
sooner than expected.

\textbf{Niedawno była operacja} --- \textit{there was an operation
recently}. Impersonal construction: not \textit{I had an operation} but
\textit{an operation was} --- the event stated as a fact, Waldi at a
slight grammatical remove from his own surgery.

\textbf{jeszcze trochę pochodzę o kulach} --- \textit{I'll walk on
crutches a little while longer}. \textbf{Pochodzę} --- future with the
prefix \textit{po-} giving a sense of duration covering a period:
\textit{I'll be going about} for a while. \textbf{O kulach} ---
\textit{on crutches}: instrumental of \textit{kule}. \textbf{Jeszcze
trochę} --- \textit{just a bit more, a little while yet} --- patient
and matter-of-fact.

The \textit{serwis pogwarancyjny} metaphor is the centrepiece. It
extends naturally from the whole chapter's inventory of bodily
deterioration --- folds, veneers, glasses, dentures --- but Waldi
crystallises it into a single perfectly chosen phrase. What makes it
land is that he says it without complaint, almost cheerfully: he has
accepted the terms of the post-warranty period and is simply reporting
on current service status.

% ────────────────────────────────────────────────────────────

\poltext{--- Do wesela się zagoi --- dodaje optymistycznie Daisy.}
\poltrans{`It'll heal by the wedding' --- Daisy adds optimistically.}

\textbf{Do wesela się zagoi} --- a set Polish folk reassurance:
\textit{it'll heal by the wedding}, meaning \textit{don't worry, you'll
be fine in time for what matters}. Mothers say it to children who have
scraped their knees. Applied to a man in his sixties with a hip
replacement, standing at a funeral, it is either the most cheerful
possible response --- or, depending on your angle, quietly devastating.
At this age, \textit{do wesela} might mean a grandchild's wedding,
still years away.

\textbf{zagoi się} --- from \textit{zagoić się}, to heal over. The
\textit{za-} prefix signals completion: fully healed, closed over.
Future perfective.

\textbf{dodaje optymistycznie Daisy} --- the adverb
\textbf{optymistycznie} is placed by the narrator with a certain
dryness: he notes the optimism as a quality of the remark, not
necessarily endorsing it. Three different responses to the same crutch:
Groszek mocks (\textit{nie za wcześnie?}), Waldi explains with rueful
wit (\textit{serwis pogwarancyjny}), Daisy closes with breezy folk
reassurance. Three temperaments, a handful of lines, a graveside.

% ────────────────────────────────────────────────────────────

\poltext{--- No, najpierw to musiałby się rozwieść. Polecam, ja tam 
bardzo dobrze na tym wyszłam --- w ten sposób podsumowuje Hanka ostatnie 
rewelacje w swoim życiu.}
\poltrans{`Well, first he'd have to get divorced. I recommend it --- I 
came out of it very well' --- that's how Hanka summarises the latest 
revelations in her life.}

\textbf{najpierw to musiałby się rozwieść} --- \textit{first he'd have
to get divorced}. Hanka has understood the furniture shopping not as a
business opportunity but as the announcement of a new living situation,
and has drawn the correct inference. \textbf{Musiałby} --- conditional
of \textit{musieć}: \textit{he would have to} --- the condition stated
calmly, as a logical preliminary. \textbf{Rozwieść się} ---
\textit{to get divorced}, perfective reflexive.

\textbf{Polecam} --- \textit{I recommend it.} First person singular,
present, no hedging. Not \textit{it wasn't so bad} or \textit{it has
its advantages} --- a clean, direct endorsement. The same word you
would use recommending a restaurant.

\textbf{ja tam bardzo dobrze na tym wyszłam} --- \textit{I came out of
it very well}. \textbf{Tam} here is not spatial but a colloquial
particle adding casual personal emphasis: \textit{for my part, as for
me}. \textbf{Wyjść na czymś} --- \textit{to come out of something} in
terms of result: to fare well or badly. \textbf{Bardzo dobrze} ---
Hanka is not merely surviving her divorce; she came out ahead. The
furniture business, presumably, is part of the evidence.

\textbf{ostatnie rewelacje w swoim życiu} --- \textit{the latest
revelations in her life}. \textbf{Rewelacje} carries a flavour of
\textit{bombshells, sensational news} --- the word you'd use for gossip
or surprising disclosures. Applied to one's own divorce, stated
cheerfully at a funeral, it has a wonderfully self-aware theatricality.
The narrator labels it all a \textit{summary}, watching Hanka perform
her own life story with total composure.

In two sentences Hanka has: correctly diagnosed the author's domestic
situation, volunteered her own experience as precedent, rendered a
verdict (\textit{polecam}), reported her outcome (\textit{bardzo
dobrze}), and had it all dryly labelled by the narrator. Not a word
wasted.

% ────────────────────────────────────────────────────────────

\poltext{--- Chodźcie do Fiszera. Potem sobie pogadacie --- postanawiam.}
\poltrans{`Go to Fiszer. You can chat afterwards' --- I decide.}

\textbf{Chodźcie do Fiszera} --- \textit{go to Fiszer / go to Fiszer's
grave}. \textbf{Chodźcie} is the second person plural imperative of
\textit{chodzić}: \textit{go, come on, move}. The group has been
standing around exchanging divorces and hip replacement diagnoses; now
the author redirects them. \textit{Do Fiszera}: genitive, the direction
toward Fiszer. Not \textit{to the grave}, not \textit{to the ceremony}
--- just \textit{to Fiszer}, as if he's waiting nearby and they've kept
him long enough.

\textbf{Potem sobie pogadacie} --- \textit{you can chat afterwards}.
\textbf{Sobie} is the reflexive dative used as a casual particle ---
it softens and domesticates the verb, implying something done at
leisure, for one's own pleasure: \textit{have yourselves a chat}.
\textbf{Pogadać} is the perfective of the colloquial \textit{gadać}
(to chat, to natter) --- not the formal \textit{rozmawiać}.

\textbf{postanawiam} --- \textit{I decide, I resolve}. Present tense,
first person singular. Not \textit{I said} or \textit{I suggested} but
\textit{I decide} --- the same weighty \textit{postanowienie}
(resolution) from earlier in the chapter, now in verb form.

All the reunion warmth has been real, but \textit{chodźcie do Fiszera}
brings everything back. They are, after all, at a funeral. First things
first.

% ── Nieobecni / The Absent ───────────────────────────────────
\section{Nieobecni / The Absent}

\poltext{--- A nieobecni? Sławek nie przyjdzie. Jest znowu gdzieś na 
końcu świata --- wyjaśnia siostra.}
\poltrans{`And those who aren't here? Sławek won't come. He's somewhere 
at the end of the world again' --- his sister explains.}

\textbf{A nieobecni?} --- \textit{And those who aren't here?}
\textbf{Nieobecni} --- \textit{the absent ones}, plural adjective used
as a noun, from \textit{obecny} (present) with the negating
\textit{nie-}. Someone takes a mental roll call. The funeral is also
an accounting of who didn't make it.

\textbf{Jest znowu gdzieś na końcu świata} --- \textit{he's somewhere
at the end of the world again}. \textbf{Znowu} --- \textit{again} ---
the word that carries the whole character sketch. This is not new
behaviour; Sławomir is habitually, reliably somewhere unreachable.
\textbf{Gdzieś} --- \textit{somewhere} --- the vagueness deliberate.
\textbf{Na końcu świata} --- \textit{at the end of the world} --- a
fixed Polish expression for an extremely remote place, equivalent to
\textit{the back of beyond}.

\textbf{wyjaśnia siostra} --- \textit{his sister explains}. She has
clearly explained Sławek's whereabouts before --- perhaps many times.
The verb \textit{wyjaśniać} implies there was something requiring
clarification; she supplies the answer to the question the group was
already asking.

% ────────────────────────────────────────────────────────────

\poltext{Sławomir był artystą konceptualnym (jak mawia Daisy, 
współczesna sztuka nie jest od tego, by ją rozumieć), wykładowcą 
i podróżnikiem, postacią nieprzeciętną.}
\poltrans{Sławomir was a conceptual artist (as Daisy likes to say, 
contemporary art isn't there to be understood), a lecturer and 
traveller, an exceptional figure.}

\textbf{Sławomir} --- the full name, now that we're in the narrator's
expository voice rather than dialogue. He was \textit{Sławek} to the
group; here he gets the dignity of his proper name.

\textbf{był artystą konceptualnym} --- \textit{was a conceptual
artist}. Instrumental after \textit{być} (to be) in the past:
\textbf{artystą}, instrumental of \textit{artysta}. Polish requires
this throughout --- you don't \textit{be} a thing in the nominative
but \textit{be} it in the instrumental, as though identity is something
you inhabit rather than something you simply are.

\textbf{jak mawia Daisy} --- \textit{as Daisy likes to say}.
\textbf{Mawia} is a frequentative form of \textit{mówić} --- not
\textit{says} on this occasion but \textit{is wont to say, habitually
says}. It marks the parenthesis as a well-worn Daisy opinion, trotted
out reliably.

\textbf{współczesna sztuka nie jest od tego, by ją rozumieć} ---
\textit{contemporary art isn't there to be understood}. \textbf{Od
tego, by} --- \textit{for that purpose, in order to} --- a
construction meaning something exists (or doesn't exist) for a given
purpose. Daisy's parenthetical is both a defence of Sławomir's work
and a perfectly self-contained piece of comic deflation.

\textbf{podróżnik} --- a wonderfully old-fashioned, romantic word for
traveller, explorer: not a \textit{turysta} (tourist) but someone for
whom travel is a vocation or identity.

\textbf{postacią nieprzeciętną} --- \textit{an exceptional figure}.
\textbf{Nieprzeciętny} --- literally \textit{not average}, from
\textit{przeciętny} (average, mediocre). The summary epithet, placed
last with the weight of a conclusion.

The parenthetical is exquisite: rather than explaining conceptual art,
the narrator outsources the explanation to Daisy --- who doesn't
explain it either, but dismisses the need for explanation altogether.
Sławomir's work is described, defended, and gently mocked in eleven
words, none of which actually tell us anything about it.

% ────────────────────────────────────────────────────────────

\poltext{Mało małomiasteczkową.}
\poltrans{Not very small-town.}

\textbf{Mało} --- \textit{not very, scarcely} --- used as a negating
qualifier: \textit{hardly, not much of a}. It doesn't flatly negate
but diminishes.

\textbf{małomiasteczkową} --- instrumental feminine of
\textbf{małomiasteczkowy}: \textit{small-town, provincial}. A compound
adjective built from \textbf{małe miasto} (small town). It agrees
instrumentally with \textit{postacią} from the previous sentence.

The sentence functions as a dry afterthought: having assembled
Sławomir's credentials --- artist, lecturer, traveller, exceptional ---
the narrator adds this fragment as if \textit{małomiasteczkowy} were
the natural opposite of everything just listed, and confirming he
wasn't that is the final, clinching point. There is also a faint
self-awareness: these are people \textit{from} a small town, gathered
at a funeral in it. Sławomir escaped.

% ────────────────────────────────────────────────────────────

\poltext{Miał być, ale ostatecznie nie dojechał.}
\poltrans{He was supposed to come, but in the end he didn't make it.}

\textbf{Miał być} --- \textit{he was supposed to be there, he was
meant to come}. \textbf{Mieć} + infinitive is the Polish construction
for obligation, expectation, or plan: \textit{miał przyjść} --- he was
supposed to come; \textit{miał być} --- he was supposed to be there.
It implies an arrangement that existed, a slot reserved for him.

\textbf{ostatecznie nie dojechał} --- \textit{in the end he didn't
make it}. \textbf{Ostatecznie} --- \textit{ultimately, in the end} ---
carries a slight air of resignation, of a situation that played out to
its conclusion. \textbf{Dojechać} --- perfective of \textit{dojeżdżać},
to arrive by vehicle, to complete a journey. The \textit{do-} prefix
is directional completion: to travel \textit{all the way to} a
destination. Not \textit{nie przyszedł} (didn't come) but \textit{nie
dojechał} --- the gap between setting out and arriving, the journey
that didn't finish.

% ────────────────────────────────────────────────────────────

\poltext{Zabrakło też Mirka, filozofa amatora i diamentowego 
konsultanta Amwaya.}
\poltrans{Mirek was also absent --- amateur philosopher and diamond 
consultant for Amway.}

\textbf{Zabrakło} --- impersonal perfective of \textit{brakować} (to
lack, to be missing). Not \textit{Mirek nie przyszedł} (Mirek didn't
come) but \textit{there was a lacking of Mirek} --- the person rendered
as an absence, a gap in the room. \textbf{Mirka}: genitive, as required
by the impersonal \textit{zabrakło}.

\textbf{filozofa amatora} --- \textit{amateur philosopher}. Both in
the genitive, in apposition to \textit{Mirka}. \textbf{Amator} in
Polish carries less condescension than in English; it means someone
who does something from love (\textit{amor}) rather than
professionally.

\textbf{diamentowego konsultanta Amwaya} --- \textit{diamond consultant
for Amway}. \textbf{Diamentowy} --- \textit{diamond} --- is an actual
rank in Amway's multi-level marketing hierarchy, awarded to top-tier
distributors. The collision of \textit{filozofa amatora} and
\textit{diamentowego konsultanta Amwaya} in a single breath is the
entire joke: Mirek has two identities, and the narrator presents them
with perfect deadpan equivalence. Neither is privileged over the other.

% ────────────────────────────────────────────────────────────

\poltext{Był nieobecnym usprawiedliwionym, na stałe mieszkał we 
Włoszech. Łączył się w bólu na Facebooku.}
\poltrans{He was an excused absence --- he lived permanently in Italy. 
He joined in the grief on Facebook.}

\textbf{nieobecnym usprawiedliwionym} --- \textit{an excused absence}.
In Polish schools, absences are formally categorised as
\textbf{usprawiedliwione} (excused, with a valid reason) or
\textbf{nieusprawiedliwione} (unexcused). The narrator applies the
bureaucratic vocabulary of a primary school attendance book to a grown
man's failure to attend his old classmate's funeral. Since Mirek lives
in Italy, his absence is officially \textit{usprawiedliwione}.

\textbf{na stałe} --- \textit{permanently, for good}. Not
\textit{currently} or \textit{for now} but settled, fixed.

\textbf{Łączył się w bólu na Facebooku} --- \textit{he joined in the
grief on Facebook}. \textbf{Łączyć się w bólu} --- \textit{to unite
in pain/grief} --- is the formal Polish condolence phrase used in
funeral notices and obituaries: \textit{I share in your grief, I unite
with you in sorrow}. The ancient, solemn phrase has been redirected
through a social media platform. The narrator records it without
editorial comment. None is needed.

% ── Umarła klasa / The Dead Class ───────────────────────────
\section{Umarła klasa / The Dead Class}

\poltext{A martwi? Po uroczystości Groszek, czyniąc cmentarne honory 
domu, wyciągnął wreszcie okolicznościową flaszkę i zaproponował, by 
odwiedzić resztę naszych na kwaterach, tę „umarłą klasę".}
\poltrans{And the dead? After the ceremony, Groszek, playing the 
graveyard host, finally produced the occasion's bottle and proposed 
that we visit the rest of our people in their plots --- that `dead class'.}

\textbf{A martwi?} --- \textit{And the dead?} Two words, and the
question turns the whole register on its head. We've had the absent
living --- Sławomir at the end of the world, Mirek on Facebook from
Italy. Now the narrator pivots to the other category of non-attenders:
those in the ground.

\textbf{czyniąc cmentarne honory domu} --- \textit{playing the
graveyard host}, literally \textit{doing the cemetery honours of the
house}. \textbf{Czyniąc} --- present adverbial participle of
\textit{czynić} (to do, to make) --- slightly elevated, literary.
\textbf{Honory domu} --- \textit{the honours of the house} --- a fixed
phrase for the duties of a host, transplanted wholesale to a cemetery,
so that Groszek becomes the host of this outdoor gathering among the
graves.

\textbf{wyciągnął wreszcie okolicznościową flaszkę} --- \textit{finally
produced the occasion's bottle}. \textbf{Wyciągnął} --- \textit{pulled
out, produced} --- the perfective suggesting the bottle has been
waiting, held in reserve. \textbf{Wreszcie} --- \textit{finally, at
last}. \textbf{Okolicznościową} --- \textit{for the occasion,
commemorative}: an adjective applied to things made or brought for a
specific event. This is the fulfilment of \textit{zdecydowaliśmy, że
dla uczczenia jego pamięci trzeba się napić} from the very opening of
the chapter.

\textbf{na kwaterach} --- \textit{in their plots}. \textbf{Kwatera}
means a plot or section in a cemetery, but also a billet or lodging
(military quarters, accommodation). Visiting the dead in their
\textit{kwatery} as one might visit friends in their lodgings.

\textbf{tę „umarłą klasę"} --- \textit{that `dead class'}. A direct
reference to Tadeusz Kantor's legendary 1975 theatre piece
\textit{Umarła klasa}, one of the defining works of Polish
avant-garde theatre, in which elderly figures return to their school
desks and encounter the ghosts of their childhood selves. The narrator
names it quietly, without explanation, trusting the reader. The
classmates gathering at Fiszer's grave, the absent ones already under
it --- they are Kantor's image made literal.

% ────────────────────────────────────────────────────────────

\poltext{Oczywiście w tym momencie zmyślam. Groszek by tak nie 
powiedział. Groszek był po politechnice.}
\poltrans{Obviously at this point I'm making things up. Groszek 
wouldn't have said it like that. Groszek had been to polytechnic.}

\textbf{zmyślam} --- present tense of \textit{zmyślać} (to make up,
to fabricate, to invent) --- not apologetic, not alarmed, just stating
a fact. The \textit{oczywiście} (obviously) is the masterstroke: of
course he's inventing. Did you think he wasn't?

\textbf{Groszek by tak nie powiedział} --- \textit{Groszek wouldn't
have said it like that}. The narrator has caught himself putting
\textit{cmentarne honory domu} and \textit{tę umarłą klasę} into
Groszek's mouth --- language that is clearly the author's, literary
and allusive --- and corrects the record.

\textbf{Groszek był po politechnice} --- \textit{Groszek had been to
polytechnic}. The construction \textit{być po [instytucji]} means to
have attended and graduated: \textit{po wojsku} (having done military
service), \textit{po studiach} (having finished university). The
polytechnic --- a technical university, engineering-focused --- is
offered as the complete explanation. Engineers do not say \textit{tę
umarłą klasę}.

The joke runs in two directions: gently at Groszek's expense (engineers
don't talk like that) and at the author's own (I was showing off and I
know it). The self-correction is funnier than the original flourish,
and the original flourish was already very good.

% ────────────────────────────────────────────────────────────

\poltext{Są tu prawie wszyscy nasi klasowi umarli! Nie chce się 
wierzyć! Ze łzami w oczach pociągnęliśmy z gwinta i ruszyliśmy 
w obchód.}
\poltrans{Almost all our class dead are here! It's hard to believe! 
With tears in our eyes we took a pull from the bottle and set off 
on our rounds.}

\textbf{Są tu prawie wszyscy nasi klasowi umarli!} --- what Groszek
actually said. Plain, direct, with an exclamation mark.
\textbf{Klasowi umarli} --- \textit{class dead}, the dead of the class
--- a noun phrase of remarkable bluntness. Not \textit{ci, których
straciliśmy} (those we have lost) but simply \textit{umarli} --- the
dead ones, \textit{ours}. \textbf{Prawie wszyscy} --- \textit{almost
all} --- the \textit{prawie} doing quiet, terrible work: not all yet,
but nearly.

\textbf{Nie chce się wierzyć!} --- \textit{It's hard to believe! One
doesn't want to believe it!} The impersonal \textit{chce się}
construction: not \textit{I don't want to believe it} but \textit{it
doesn't want to be believed}, \textit{belief resists}.

\textbf{pociągnęliśmy z gwinta} --- \textit{we took a pull from the
bottle}. \textbf{Pociągnąć z gwinta} is a vivid colloquial expression:
\textbf{gwint} literally means a screw-thread --- the neck of the
bottle where the cap screws on. To drink \textit{z gwinta} is to drink
straight from the bottle, no glasses.

\textbf{ruszyliśmy w obchód} --- \textit{we set off on our rounds}.
\textbf{Obchód} --- a round, a circuit, an inspection tour: the word
used for a guard's patrol, a doctor's ward round, a ceremonial
procession. Here: a tour of the graves of their dead classmates, bottle
in hand, tears in eyes, Groszek leading the way.

The chapter has been moving toward this moment from its first sentence.
The resolution to drink is honoured; Groszek's plain words --- not the
narrator's literary flourishes --- name what they are doing and why.

% ── Obchód / The Rounds ──────────────────────────────────────
\section{Obchód / The Rounds}

\poltext{Zostawiliśmy świeży grób Fiszera z wieńcami, kwiatami.}
\poltrans{We left Fiszer's fresh grave with its wreaths, its flowers.}

\textbf{Zostawiliśmy} --- \textit{we left, we left behind}. Perfective
past of \textit{zostawiać} --- to leave something in place, to leave
it as it is. Not \textit{opuściliśmy} (we departed from) but
\textit{zostawiliśmy} --- they left it there, behind them, as they
moved on.

\textbf{świeży grób} --- \textit{the fresh grave}. \textbf{Świeży}
--- \textit{fresh} --- the same word for fresh bread, fresh air,
fresh paint. Applied to a grave it is entirely ordinary Polish usage,
but the word still lands with its full weight: the earth newly turned,
not yet settled.

\textbf{z wieńcami, kwiatami} --- \textit{with its wreaths, its
flowers}. The comma between them, no \textit{i} (and), gives the list
an almost visual quality --- the eye moving across the grave's surface,
noting one thing, then another.

They set off on their rounds --- and the first thing they do is leave
Fiszer behind. The living move on; the fresh grave stays.

% ────────────────────────────────────────────────────────────

\poltext{Z przyjmującą kondolencje zapłakaną i śmiertelnie zmęczoną żoną.}
\poltrans{With the tear-stained and mortally exhausted wife receiving 
condolences.}

The sentence continues the list begun with \textit{z wieńcami,
kwiatami} --- the grave left behind, and now: the widow.

\textbf{Z przyjmującą kondolencje} --- \textit{with the one receiving
condolences}. \textbf{Przyjmującą} is the present active participle of
\textit{przyjmować} (to receive, to accept) in the instrumental
feminine. She is not described as standing, or waiting, or grieving
--- but as \textit{receiving condolences}: defined entirely by the
ritual function she is currently performing.

\textbf{zapłakaną} --- \textit{tear-stained, having been crying}.
Instrumental feminine participle from \textit{zapłakać się} --- to
have cried, to be red-eyed with weeping. The \textit{za-} prefix marks
the completed state: she has been crying and it shows.

\textbf{śmiertelnie zmęczoną} --- \textit{mortally exhausted}.
\textbf{Śmiertelnie} --- \textit{mortally, deathly} --- the intensifier
drawn from death itself, entirely idiomatic in Polish
(\textit{śmiertelnie znudzony} --- bored to death). Here it sits with
particular weight: she is deathly tired, at a death, having transported
her husband's body from the south of France. The logistics noted so
dryly at the chapter's opening --- \textit{musiała transportować jego
zwłoki} --- have arrived at their human cost.

% ────────────────────────────────────────────────────────────

\poltext{Niewprawnym zygzakiem poruszaliśmy się za Groszkiem pomiędzy 
grobami po wąskich cmentarnych ścieżkach, szukając klasowych 
nieboszczyków.}
\poltrans{In an unsteady zigzag we moved behind Groszek between the 
graves along the narrow cemetery paths, looking for the class dead.}

\textbf{Niewprawnym zygzakiem} --- \textit{in an unsteady zigzag}.
The instrumental of manner. \textbf{Niewprawny} --- \textit{unskilled,
unpractised, clumsy} --- from \textit{wprawny} (skilled) with the
negating \textit{nie-}. The bottle from Groszek's pocket has done its
early work. The group is weaving.

\textbf{poruszaliśmy się} --- \textit{we moved, we made our way}.
Imperfective past, the reflexive \textit{się} with \textit{poruszać}
giving the sense of sustained, ongoing movement.

\textbf{za Groszkiem} --- \textit{behind Groszek}. Instrumental after
\textit{za}. Groszek leads; the others follow. He is, after all, the
one doing the \textit{honory domu}.

\textbf{po wąskich cmentarnych ścieżkach} --- \textit{along the narrow
cemetery paths}. \textbf{Po} with the locative expresses movement
along a surface: \textit{po ulicy} (along the street), \textit{po
podłodze} (across the floor). The paths are narrow; the zigzag is
wide; the combination explains itself.

\textbf{szukając klasowych nieboszczyków} --- \textit{looking for the
class dead}. \textbf{Nieboszczyk} --- \textit{the deceased} --- a
slightly old-fashioned, almost tender word, softer than \textit{trup}
(corpse), more colloquial than \textit{zmarły} (the departed). Here
in the plural, as a category: the classmates who got there before them.

The image is both physically comic and quietly moving: a small group
weaving behind Little Pea through a Polish cemetery, bottle in hand,
searching for their dead.

% ── Cmentarny opis / Cemetery Survey ────────────────────────
\section{Cmentarny opis / Cemetery Survey}

\poltext{Tandetne walentynkowe grobowce nowych zmarłych z 
asymetrycznymi serduszkami obok zmurszałych, starych niemieckich 
pomników.}
\poltrans{Tawdry Valentine's Day tombs of the newly dead with asymmetric 
little hearts, next to crumbling, old German monuments.}

A sentence with no main verb --- a pure observation, a painter's note.
The eye sweeps the cemetery and records what it sees.

\textbf{Tandetne} --- \textit{tawdry, cheap, tacky}. From
\textit{tandeta} --- cheap goods, kitsch. An unsparing word.

\textbf{walentynkowe} --- \textit{Valentine's Day}, used as an
adjective. Applied to grave monuments: the aesthetic of Valentine's
Day --- pink, heart-shaped, decorative --- transposed onto funerary
architecture.

\textbf{z asymetrycznymi serduszkami} --- \textit{with asymmetric
little hearts}. \textbf{Serduszkami} --- diminutive instrumental plural
of \textit{serduszko}, itself a diminutive of \textit{serce} (heart).
The double diminutive: not \textit{hearts} but \textit{little hearts},
the kind on greetings cards. \textbf{Asymetrycznymi} --- the detail
that makes it real. Not even well-made little hearts.

\textbf{zmurszałych, starych niemieckich pomników} --- \textit{crumbling,
old German monuments}. \textbf{Zmurszałych} --- \textit{crumbling,
decayed, rotted} --- describing stone that has broken down with age
and damp. The cemetery predates the postwar border shifts; this part
of Silesia was German until 1945, and the old German graves remain,
weathering beside their newer Polish neighbours. The aesthetic contrast
is also a historical one.

% ────────────────────────────────────────────────────────────

\poltext{Skremowana klasa ekonomiczna w lastryko obok otoczonej 
masywnym łańcuchem, zapuszczonej mogiły Bojowników o Wolność z Dawno 
Minionych Czasów.}
\poltrans{The cremated economy class in terrazzo, next to the overgrown 
tomb of the Fighters for Freedom from Times Long Past, surrounded by 
a massive chain.}

\textbf{Skremowana klasa ekonomiczna} --- \textit{the cremated economy
class}. \textbf{Klasa ekonomiczna} --- \textit{economy class}, the
airline term for the cheapest seats, applied to those who could not
afford a full grave plot and were cremated into smaller, cheaper
monuments. The narrator has classified the dead by their funerary
budget using the language of air travel.

\textbf{w lastryko} --- \textit{in terrazzo}. \textbf{Lastryko} is
terrazzo --- the composite stone material ubiquitous in communist-era
Polish construction, common for cheaper grave surrounds. Functional,
durable, aesthetically modest.

\textbf{zapuszczonej} --- \textit{overgrown, neglected}. From
\textit{zapuścić} --- to let go, to allow to become overgrown. No one
tends this grave.

\textbf{Bojownicy o Wolność z Dawno Minionych Czasów} --- the official
communist-era designation for resistance fighters and partisans:
\textit{Fighters for Freedom} --- a bureaucratic honorific carved into
stone on countless Polish monuments. \textbf{Z Dawno Minionych Czasów}
--- \textit{from Times Long Past} --- and here the capitals are the
narrator's, not the monument's. The phrase is elevated, almost
fairy-tale in register (\textit{dawno, dawno temu} --- once upon a
time). The official heroism now left to go to seed behind a massive
chain.

% ────────────────────────────────────────────────────────────

\poltext{Sylikonowy anioł z wtryskarki obok sepiowych fotografii z 
niemodnymi krawatami i fryzurami w stylu świętej pamięci Waldemara 
Koconia.}
\poltrans{A silicone angel from an injection moulding machine, next to 
sepia photographs with unfashionable ties and hairstyles in the style 
of the late, lamented Waldemar Kocoń.}

\textbf{Sylikonowy anioł z wtryskarki} --- \textit{a silicone angel
from an injection moulding machine}. \textbf{Wtryskarki} --- genitive
of \textit{wtryskarka}, an injection moulding machine. The angel did
not come from a sculptor's hands but from a factory production line,
extruded in silicone. The word \textit{wtryskarki} contains the entire
critique.

\textbf{sepiowych fotografii} --- \textit{sepia photographs}. The oval
ceramic or glass photographs mounted on Polish graves, showing the
deceased in life --- a widespread and distinctive tradition.

\textbf{z niemodnymi krawatami i fryzurami} --- \textit{with
unfashionable ties and hairstyles}. The photographs capture their
subjects in the fashion of the decade of their death, now dated. The
ties are wide or narrow in the wrong way; the hair is set in a style
that places them precisely in time.

\textbf{w stylu świętej pamięci Waldemara Koconia} --- \textit{in the
style of the late, lamented Waldemar Kocoń}. \textbf{Świętej pamięci}
--- \textit{of blessed/holy memory} --- the traditional Polish
honorific for the deceased. Here deployed with perfect comic precision:
Waldemar Kocoń becomes the unwitting style reference for an entire era
of Polish men's grooming.

% ────────────────────────────────────────────────────────────

\poltext{Nazwiska niemieckie, polskie, nieczytelne. Cmentarny opis 
obyczajów minionego wieku.}
\poltrans{Names German, Polish, illegible. A cemetery account of the 
customs of the past century.}

\textbf{Nazwiska niemieckie, polskie, nieczytelne} --- three adjectives,
no verb. The list moves from legible history --- German, Polish, the
two peoples who have buried their dead here --- to \textit{nieczytelne}:
illegible, worn beyond reading. Time doesn't distinguish between German
and Polish; it erases them equally.

\textbf{Cmentarny opis obyczajów minionego wieku} --- \textit{a
cemetery account of the customs of the past century}. The narrator
steps back and names what this whole passage has been: a
\textit{cmentarny opis}, a survey conducted in a cemetery.
\textbf{Obyczaje} --- customs, manners, the characteristic practices
that define a time and place. The Valentine's hearts, the silicone
angel, the sepia photographs, the crumbling German stone --- all of
it the material of a social historian's survey, conducted on foot,
slightly unsteadily, with a bottle.

% ── Grób Roberta / Robert's Grave ───────────────────────────
\section{Grób Roberta / Robert's Grave}

\poltext{Po krótkich poszukiwaniach stanęliśmy obok grobu, który, 
zamiast pomnika, wieńczyła marmurowa piłka.}
\poltrans{After a brief search we stopped beside a grave which, instead 
of a headstone, was crowned by a marble ball.}

\textbf{stanęliśmy} --- \textit{we stopped, we came to a stand}.
Perfective past of \textit{stanąć}: to stop, to come to a halt, to
take up a position. The movement ceases; they have arrived.

\textbf{który, zamiast pomnika, wieńczyła marmurowa piłka} --- the
relative clause delivers its surprise with structural delay:
\textit{który} (which) --- \textit{zamiast pomnika} (instead of a
headstone) --- \textit{wieńczyła} (was crowned by) ---
\textit{marmurowa piłka} (a marble ball). The subject arrives last.
\textbf{Wieńczyć} --- \textit{to crown, to top, to cap} --- a verb
with ceremonial weight: to crown a building, to crown an achievement.
Here: to sit atop a grave. Not a cross, not an angel, not a stone slab.
A ball. In marble.

% ────────────────────────────────────────────────────────────

\poltext{--- To już druga, mniejsza. Poprzednią zajebali --- wyjaśnił 
nasz przewodnik.}
\poltrans{`This is already the second one, the smaller one. They nicked 
the previous one' --- our guide explained.}

\textbf{To już druga, mniejsza} --- \textit{this is already the second
one, the smaller one}. \textbf{Już} marks the sequence: the marble
ball is a replacement, and a downgraded one.

\textbf{Poprzednią zajebali} --- \textit{they nicked the previous one}.
\textbf{Zajebali} --- vulgar perfective past of \textit{zajebiać}, to
steal, to nick. Thoroughly common in spoken Polish, it does something
no polite synonym could: it combines theft with exasperated resignation.
\textit{Ukradli} (they stole) would be neutral; \textit{zajebali}
implies this is the kind of thing that happens to marble balls left in
cemeteries, and everyone knows it.

\textbf{nasz przewodnik} --- \textit{our guide}. \textbf{Przewodnik}
--- the same word for a tour guide, a guidebook, a scout leader.
Groszek, leading them \textit{niewprawnym zygzakiem}, is now formally
elevated to \textit{nasz przewodnik}.

% ────────────────────────────────────────────────────────────

\poltext{Grób Roberta. Adam leży obok. Adam i Robert Mitwerandu. 
Czarni śląscy bracia. Czarniejsi od węgla!}
\poltrans{Robert's grave. Adam lies beside it. Adam and Robert 
Mitwerandu. Black Silesian brothers. Blacker than coal!}

\textbf{Grób Roberta} --- two words, a new name. The \textit{obchód}
moves on.

\textbf{Adam leży obok} --- \textit{Adam lies beside it}.
\textbf{Leży} --- \textit{lies}, present tense of \textit{leżeć}:
the verb used for lying down but also for the dead in their graves.
\textit{Tu leży} --- \textit{here lies} --- appears on every headstone.
The present tense gives it quiet persistence.

\textbf{Adam i Robert Mitwerandu} --- the surname arrives and
everything shifts. Unmistakably African in origin: these are two
brothers buried side by side in a Silesian cemetery, with a name from
another continent.

\textbf{Czarni śląscy bracia} --- \textit{Black Silesian brothers}.
No verb, just the designation, stated plainly. The combination is not
a paradox to the narrator --- it is simply a fact: they were Black,
they were Silesian, they were brothers, they are here.

\textbf{Czarniejsi od węgla!} --- \textit{Blacker than coal!} The
exclamation mark carries warmth, not shock. Coal is the defining
substance of Silesia --- the region was built on it, its identity is
inseparable from it. To be \textit{czarniejsi od węgla} in Silesia
is to be measured against the local absolute. The voice is that of
someone who knew them, saying this with affection and memory.

% ────────────────────────────────────────────────────────────

\poltext{Owoce internacjonalistyczno-seksualnych uniesień czarnoskórego 
studenta, który po skończeniu w Polsce studiów zabrał się z powrotem 
na Czarny Ląd do proklamowanego w miejsce dawnej Rodezji Zimbabwe, 
zostawiając dwójkę małych Murzynków i ich śląską matkę w infamii.}
\poltrans{The fruits of the internationalist-sexual passions of a 
dark-skinned student, who after finishing his studies in Poland took 
himself back to the Black Continent to the newly proclaimed Zimbabwe 
in place of the former Rhodesia, leaving two small mixed-race children 
and their Silesian mother in disgrace.}

A note on language: \textbf{Murzyn} is an old Polish word for a Black
person, now considered offensive in contemporary usage. It appears here
as the narrator's reported voice, reflecting the language of the
community being described --- not as an endorsement.

\textbf{internacjonalistyczno-seksualnych uniesień} --- a compound of
the narrator's coinage, fusing two registers: \textit{internacjonalistyczny}
(internationalist) --- the official communist vocabulary of solidarity
between peoples --- and \textit{seksualny} (sexual). The hyphen joins
what the state kept carefully separate.

\textbf{czarnoskórego studenta} --- \textit{of a dark-skinned student}.
Under communism, Poland hosted many students from Africa and Asia
through solidarity agreements; this is the specific historical context.

\textbf{zabrał się z powrotem na Czarny Ląd} --- \textit{took himself
back to the Black Continent}. \textbf{Czarny Ląd} --- the traditional
Polish name for Africa.

\textbf{do proklamowanego w miejsce dawnej Rodezji Zimbabwe} --- \textit{to
the newly proclaimed Zimbabwe in place of the former Rhodesia}. The
narrator specifies the historical moment precisely: Zimbabwe was
proclaimed in 1980, dating the story and giving the father's departure
its political context --- he was returning to a newly independent nation.

\textbf{zostawiając\ldots w infamii} --- \textit{leaving\ldots in
disgrace}. \textbf{Infamia} --- public disgrace, loss of good name,
from Latin. In a small Silesian town in 1980, an unmarried woman with
two mixed-race children, abandoned, would have faced exactly this.

The sentence is long because the situation is long: a whole life
compressed into a subordinate clause. Adam and Robert grew up here,
went to school here, are buried here. Their father returned to a newly
free Zimbabwe; their mother stayed in infamy.

% ────────────────────────────────────────────────────────────

\poltext{Nie sprawdziło się czarnowidztwo śląskiego klasyka, Janoscha, 
który pisał o tych stronach: „Bo żaden Murzyn by się tu nie uchował. 
Każdego odmieńca ukradkiem nocą by zatłukli".}
\poltrans{The dark prophecy of the Silesian classic, Janosch, who wrote 
about these parts, did not come true: `Because no Black man would 
survive here. They would beat every outsider to death in secret, 
at night.'}

\textbf{Nie sprawdziło się} --- \textit{it did not come true, it was
not borne out}. \textbf{Sprawdzić się} --- to prove true. The negation
comes first and is the whole point of the sentence.

\textbf{czarnowidztwo} --- \textit{dark prophecy, doom-saying}. From
\textit{czarny} (black) + \textit{widzieć} (to see): literally
\textit{black-seeing}, the habit of seeing only the darkest possible
outcome. A rich compound.

\textbf{Janosch} --- pen name of Horst Eckert (born 1931), a writer
and illustrator born in Zabrze in what was then German Silesia, best
known internationally for his children's books but also the author of
grimly funny, unsentimental prose about Silesian life. A
\textit{klasyk} of this specific region.

\textbf{odmieńca} --- \textit{outsider, deviant, one who is different}.
From \textit{odmienny} (different, other). Stronger than mere
\textit{obcy} (stranger): an \textit{odmieniec} is someone who
deviates from the local norm, who is visibly, fundamentally other.

The quoted words are Janosch's, presented in order to be refuted ---
the graves of Adam and Robert Mitwerandu are the refutation. The
narrator doesn't argue with Janosch; he simply points at the evidence.
Two Black Silesian brothers, buried in their hometown cemetery,
remembered by their classmates with a bottle. The \textit{obchód} is
doing more than commemorating the dead; it is, in its unsteady way,
a testimony.

% ── Bracia Mitwerandu / The Mitwerandu Brothers ──────────────
\section{Bracia Mitwerandu / The Mitwerandu Brothers}

\poltext{Na przekór temu, chłopcy dobrze radzili sobie w dzieciństwie 
pośród śląskich rówieśników ze szkolnego boiska.}
\poltrans{Despite this, the boys managed well in childhood among their 
Silesian peers from the school playground.}

\textbf{Na przekór temu} --- \textit{despite this, in defiance of this}.
\textbf{Na przekór} is a fixed prepositional phrase meaning \textit{against,
contrary to, in spite of} --- with a slight flavour of active defiance
rather than passive endurance. \textbf{Temu} points back at everything
just said: the infamy, the abandonment, Janosch's prediction.

\textbf{dobrze radzili sobie} --- \textit{managed well, coped well}.
\textbf{Radzić sobie} --- \textit{to manage, to cope, to get along}
--- one of the most useful and idiomatic Polish constructions. The
reflexive \textit{sobie} is inseparable from it.

\textbf{pośród śląskich rówieśników} --- \textit{among their Silesian
peers}. \textbf{Pośród} --- \textit{among, amidst} --- slightly more
literary than \textit{wśród}. \textbf{Rówieśnicy} --- \textit{peers,
contemporaries, age-mates} --- a precise and rather beautiful word.

% ────────────────────────────────────────────────────────────

\poltext{Być może pomagała w tym siła pięści i kopniaków Roberta. 
Poza tym dwójce trudniej było spuścić łomot niż jakiejś pojedynczej 
sierotce.}
\poltrans{Perhaps Robert's fists and kicks helped with that. Besides, 
it was harder to give a beating to a pair than to some lone orphan.}

\textbf{siła pięści i kopniaków} --- \textit{the force of fists and
kicks}. \textbf{Pięść} --- \textit{fist}; \textbf{kopniak} ---
\textit{kick}. Robert was the physically formidable one; his reputation
on the playground provided a practical solution to Janosch's predicted
violence.

\textbf{spuścić łomot} --- \textit{to give a beating, to thrash}. A
vivid colloquial expression: \textbf{łomot} is the sound and act of a
heavy blow; \textbf{spuścić} (to deliver) makes the phrase literal and
physical.

\textbf{niż jakiejś pojedynczej sierotce} --- \textit{than to some
lone orphan}. \textbf{Jakiejś} --- \textit{some, any old} --- genitive
of the indefinite \textit{jakiś}, dismissive in register.
\textbf{Sierotce} --- dative of \textit{sierotka}, diminutive of
\textit{sierota} (orphan): the diminutive making it more pitiful, more
defenceless. The sentence is bleakly practical: two children together
were simply a harder target.

% ────────────────────────────────────────────────────────────

\poltext{W każdym razie trzymali się razem i w podstawówce, zamiast 
ofiarami rasizmu, częściej byli obiektami podziwu, zwłaszcza wśród 
koleżanek. Choć nie tylko młodociane adoratorki były pod ich urokiem.}
\poltrans{In any case they stuck together at primary school too, and 
instead of being victims of racism, they were more often objects of 
admiration, especially among the girls. Though it wasn't only their 
youthful admirers who were under their spell.}

\textbf{trzymali się razem} --- \textit{they stuck together, they kept
together}. A warm, solid expression for solidarity.

\textbf{zamiast ofiarami rasizmu} --- \textit{instead of being victims
of racism}. \textbf{Zamiast} + instrumental: \textit{instead of}.
\textbf{Ofiarami} --- instrumental plural of \textit{ofiara} (victim).
The word \textit{rasizm} is stated plainly and without deflection.

\textbf{obiektami podziwu} --- \textit{objects of admiration}.
\textbf{Obiektami} --- instrumental plural of \textit{obiekt}: slightly
clinical, deliberately so --- \textit{objects} rather than
\textit{subjects} of admiration, the gaze of others still constituting
them as something to be looked at, even when that look is favourable.

\textbf{adoratorki} --- \textit{female admirers, devotees} --- a
slightly old-fashioned, romantic word. \textbf{Pod urokiem} ---
\textit{under the spell/charm of} --- a fixed expression.
\textbf{Urok}: charm, spell, enchantment --- the same word used in
fairy tales and in everyday compliments. The final sentence opens a
door that the next passage walks through.

% ────────────────────────────────────────────────────────────

\poltext{--- Zobocz, jak się oni gryfnie nauczyli godać! --- pochwaliła. 
Dopiero gdy z ust chłopaków zaczęły lecieć kurwy i chuje, dodała z 
ledwo wyczuwalną już nutką podziwu: --- A chachory z nich som\ldots}
\poltrans{`See how nicely they've learned to talk!' --- she said 
approvingly. Only when expletives started flying from the boys' mouths 
did she add, with a barely perceptible note of admiration: `And they're 
proper rascals, they are\ldots'}

This passage is dense with Silesian dialect --- arguably the richest
single sentence in the chapter.

\textbf{Zobocz} --- Silesian dialect for \textit{patrz} or
\textit{spójrz} (look, see). The specifically Silesian imperative form,
immediately placing the speaker within the regional vernacular.

\textbf{gryfnie} --- \textit{nicely, beautifully, well}. One of the
most characteristic Silesian dialect words, from the German
\textit{griffig} (handy, neat). In standard Polish: \textit{ładnie} or
\textit{pięknie}. \textit{Gryfnie} is warm and approving, with a
slightly old-fashioned tenderness.

\textbf{godać} --- \textit{to speak, to talk}. The quintessential
Silesian dialect verb, equivalent to standard Polish \textit{mówić}.
To \textit{godać} in Silesia is not merely to speak but to speak
\textit{Silesian} --- it is the word the dialect uses to describe
itself. The boys have been admitted into the linguistic community of
the region.

\textbf{z ust chłopaków zaczęły lecieć kurwy i chuje} --- expletives
\textit{started flying} from the boys' mouths. \textbf{Lecieć} (to
fly) is the standard colloquial verb for language that erupts freely.
The two expletives are the precise linguistic achievement that earned
the boys their final accolade --- fluency in a language includes its
profanity.

\textbf{z ledwo wyczuwalną już nutką podziwu} --- \textit{with a barely
perceptible note of admiration}. \textbf{Ledwo wyczuwalną} ---
\textit{barely detectable}. \textbf{Nutka} --- \textit{a little note,
a hint} --- diminutive of \textit{nuta} (musical note). The admiration
is present but restrained, as if she doesn't quite want to admit it.

\textbf{A chachory z nich som} --- \textit{And they're proper rascals,
they are}. Every word Silesian dialect. \textbf{Chachory} --- rascals,
scamps, rogues: not condemnation but rough, affectionate approval for
someone spirited and irreverent. \textbf{Som} --- Silesian dialect for
\textit{są} (they are). She began praising them in Silesian and ends
her verdict in the same tongue.

The ultimate credential for belonging was not merely speaking the
dialect --- that earned \textit{gryfnie} --- but swearing fluently in
it. The \textit{kurwy i chuje} are the final examination, and the boys
passed. \textit{Chachory z nich som} is an induction: you are one of
ours now.

% ── Robert / Robert ──────────────────────────────────────────
\section{Robert}

\poltext{Groszek zaczął wspominać: --- Robert nawet dobrze się 
zapowiadał jako zawodnik. Miał talent. Grał chyba w okręgówce, ale 
miał przenieść się do pierwszej ligi. Nie pamiętam --- do Gieksy?}
\poltrans{Groszek began to reminisce: `Robert even showed real promise 
as a player. He had talent. He played in the regional league I think, 
but he was due to move up to the first division. I don't remember --- 
to GKS?'}

\textbf{zaczął wspominać} --- \textit{began to reminisce}.
\textbf{Wspominać} --- imperfective: to reminisce, to recall. The
imperfective signals an open-ended flow of memories beginning. They
are standing at Robert's grave; the marble ball now has context.

\textbf{dobrze się zapowiadał jako zawodnik} --- \textit{showed real
promise as a player}. \textbf{Zapowiadać się} --- \textit{to show
promise, to look promising} --- literally \textit{to announce oneself},
to signal what one might become.

\textbf{Grał chyba w okręgówce} --- \textit{he played in the regional
league I think}. \textbf{Chyba} --- \textit{I think, probably} ---
the particle of uncertain recollection, very common in spoken Polish.
\textbf{Okręgówka} --- the colloquial name for the \textit{liga
okręgowa}, the regional/district football league: several tiers below
the top flight.

\textbf{miał przenieść się do pierwszej ligi} --- \textit{he was due
to move up to the first division}. \textbf{Mieć} + infinitive: he was
supposed to, the move was anticipated or arranged.

\textbf{Gieksa} --- the affectionate colloquial name for GKS Katowice
(\textit{Górniczy Klub Sportowy} --- the Miners' Sports Club), one of
the major Silesian football clubs. The name is formed from the letters
G-K-S sounded out, then given the feminine diminutive ending: a club
name made tender by familiarity.

The marble ball on the grave suddenly resolves. It wasn't a quirky
choice --- it was a football. Robert Mitwerandu was a footballer, and
his grave is marked accordingly.

% ────────────────────────────────────────────────────────────

\poltext{--- Nie tylko, grał w ekstraklasie siedem razy, a nawet 
powołano go do reprezentacji Polski juniorów --- Waldi był w kwestiach 
futbolu dobrze zorientowany.}
\poltrans{`Not only that --- he played in the top flight seven times, 
and was even called up to the Polish under-21 squad' --- Waldi was 
well informed on football matters.}

\textbf{grał w ekstraklasie siedem razy} --- \textit{he played in the
Ekstraklasa seven times}. \textbf{Ekstraklasa} --- the top tier of
Polish professional football. Not \textit{pierwsza liga} (which
Groszek hazarded) but \textit{ekstraklasa} --- the highest level.
\textbf{Siedem razy} --- \textit{seven times}: not a long career, but
real appearances, counted and remembered.

\textbf{powołano go do reprezentacji Polski juniorów} --- \textit{he
was called up to the Polish under-21 squad}. \textbf{Powołano} ---
impersonal passive perfective: the official verb for selection to a
national squad. Robert Mitwerandu, Black Silesian, \textit{chachar},
marble-ball-on-the-grave, was called up to represent Poland.

\textbf{Waldi był w kwestiach futbolu dobrze zorientowany} ---
\textit{Waldi was well informed on football matters}. \textbf{W
kwestiach} --- \textit{in matters of} --- a slightly formal
construction giving Waldi's knowledge quasi-expert status. Waldi, on
crutches, \textit{serwis pogwarancyjny}, supplies the precise details
that Groszek's affectionate memory had blurred.

The correction matters enormously. Robert was not merely promising
--- he achieved. The marble ball is not a symbol of what might have
been but of what was.

% ────────────────────────────────────────────────────────────

\poltext{--- I nagle umarł. Podobno na serce. Wada wrodzona? Zawał? 
Doping? Nie wiadomo --- dodała moja siostra, krzywiąc się od 
cmentarnej wyborowej.}
\poltrans{`And then he suddenly died. Apparently of the heart. A 
congenital defect? A heart attack? Doping? Nobody knows' --- my sister 
added, wincing from the cemetery vodka.}

\textbf{I nagle umarł} --- the same \textbf{nagle} from the chapter's
opening sentence (\textit{Fiszer umarł nagle, niespodziewanie i
kłopotliwie}). Here it returns starker, with no companions.

\textbf{Podobno na serce} --- \textit{apparently of the heart}.
\textbf{Podobno} --- \textit{apparently, reportedly, so they say} ---
the particle of unverified information. \textbf{Na serce}: the
preposition \textit{na} with the accusative is used for causes of
death in Polish (\textit{umrzeć na raka} --- to die of cancer).

\textbf{Wada wrodzona? Zawał? Doping?} --- three questions, three
bare nouns. \textbf{Zawał} --- \textit{heart attack}, short for
\textit{zawał serca}. \textbf{Doping?} --- the question that changes
the register: the possibility that the means used to reach the
ekstraklasa may have contributed to what ended him there. No answer
offered.

\textbf{Nie wiadomo} --- \textit{nobody knows, it is not known}. The
impersonal closing of all three questions at once.

\textbf{krzywiąc się od cmentarnej wyborowej} --- \textit{wincing
from the cemetery vodka}. \textbf{Krzywić się} --- to wince, to
grimace. \textbf{Wyborowa} is a well-known Polish vodka brand (the
name means \textit{choice, select, premium}); \textbf{cmentarna} is
the narrator's adjective --- the bottle Groszek produced at the grave,
drunk \textit{z gwinta} with tears in eyes, is still going. The
narrator's sister, who first appeared explaining Sławek's absence
(\textit{wyjaśnia siostra}), delivers the final unknowable fact about
Robert while wincing from it.

% ── Adam / Adam ──────────────────────────────────────────────
\section{Adam}

\poltext{--- A Adam? Dlaczego Adam? --- zapytałem. Groszek miał z nim 
kontakt na studiach: --- On się nigdy nie umiał odnaleźć. Był wybitnie 
zdolny na polibudzie. Z matematyki. Ale to był raczej typ filozofa. 
Kiedy miał dobre dni, dziewczyny za nim szalały. Kiedy miał złe, 
lądował na Korczaka, u czubków.}
\poltrans{`And Adam? Why Adam?' --- I asked. Groszek had been in 
contact with him at university: `He never managed to find himself. He 
was outstandingly gifted at the polytechnic. In mathematics. But he 
was more of a philosopher type. When he had good days, the girls were 
crazy about him. When he had bad ones, he ended up on Korczaka, with 
the loonies.'}

\textbf{On się nigdy nie umiał odnaleźć} --- \textit{he never managed
to find himself}. \textbf{Umiał} --- past of \textit{umieć} (to know
how to, to be capable of) --- not \textit{mógł} (was able to, had the
opportunity): he lacked the capacity itself. \textbf{Odnaleźć się}
--- \textit{to find oneself, to locate oneself in life} --- not merely
\textit{to find one's place} but to successfully inhabit one's own
existence.

\textbf{wybitnie zdolny na polibudzie} --- \textit{outstandingly gifted
at the polytechnic}. \textbf{Wybitnie} is a strong intensifier, beyond
\textit{bardzo} (very). \textbf{Na polibudzie} --- locative of
\textit{polibuda}, the affectionate colloquial name for the
\textit{politechnika} --- the same institution that produced Groszek,
who delivers this judgement with the authority of a fellow alumnus.

\textbf{raczej typ filozofa} --- \textit{more of a philosopher type}.
\textbf{Raczej} --- \textit{rather, more of a}. The mathematical gift
was real, but it sat inside someone whose nature pulled toward
abstraction and inwardness.

\textbf{dziewczyny za nim szalały} --- \textit{the girls were crazy
about him}. \textbf{Szaleć za kimś} --- \textit{to be mad for someone}
--- \textit{szaleć} is \textit{to rage, to go mad}: the girls'
attraction described as productive madness.

\textbf{lądował na Korczaka, u czubków} --- \textit{he ended up on
Korczaka, with the loonies}. \textbf{Lądował} --- \textit{he would
land} --- imperfective, the repeated landing: this happened more than
once. \textbf{Na Korczaka} --- \textit{on Korczaka Street}: the
psychiatric hospital known locally by its address. \textbf{U czubków}
--- \textit{with the loonies}: \textbf{czubek} (colloquially, a mad
person) from \textit{czub} (a tuft, a crest). Groszek's word, spoken
with rough affection.

Adam's portrait is heartbreaking in its symmetry with Robert's. Robert
had the body --- the ekstraklasa, the marble ball, the heart that
failed. Adam had the mind --- the mathematics, the philosophy, the
girls, the psychiatric ward. \textit{On się nigdy nie umiał odnaleźć}:
he had everything except the ability to be at home in his own life.

% ────────────────────────────────────────────────────────────

\poltext{(Dopisek topograficzno-historyczny: To adres lokalnych Tworek, 
w bliskim sąsiedztwie wieży wodnej. Stary budynek szpitala, tak jak 
wieżę, zaprojektowali Zillmannowie. Teraz zrujnowany, w czasach 
świetności stylem przypominał sopocki Grand Hotel, tyle że zamiast na 
molo wychodziło się z niego na końskie doły, dawne wyrobiska po piasku 
do kopalni. O ile się wychodziło, bo z psychiatryka na ogół łatwo nie 
wypuszczali).}
\poltrans{(A topographic-historical note: This is the address of the 
local Tworki, in the close vicinity of the water tower. The old hospital 
building, like the tower, was designed by the Zillmanns. Now ruined, in 
its heyday it was reminiscent in style of the Grand Hotel in Sopot, 
except that instead of onto a pier you stepped out onto horse pits, 
former sand extraction workings for the mine. If you stepped out at all, 
since they didn't generally let you out of a psychiatric hospital 
easily.)}

\textbf{Dopisek topograficzno-historyczny} --- \textit{a
topographic-historical note}. \textbf{Dopisek} --- literally \textit{a
written addition}, a note appended. The compound adjective is the
narrator playing the amateur local historian --- the same impulse that
produced \textit{cmentarny opis obyczajów minionego wieku}.

\textbf{Tworki} --- a colloquial name for a famous psychiatric hospital
outside Warsaw, used here as a common noun: \textit{the local Tworki}
--- the local equivalent. Every Polish town applies this name to its
own facility by analogy with the nationally infamous original.

\textbf{zaprojektowali Zillmannowie} --- \textit{was designed by the
Zillmanns}. A dynasty of architects, German in name, active in Silesia,
responsible for both the hospital and the water tower. Named with the
casual confidence of someone who knows local architectural history.

\textbf{w czasach świetności stylem przypominał sopocki Grand Hotel}
--- \textit{in its heyday reminiscent in style of the Grand Hotel in
Sopot}. The Grand Hotel in Sopot is a landmark of Polish seaside
elegance, a belle époque pile on the Baltic. The comparison is real:
both late nineteenth-century, both grand in ambition.

\textbf{zamiast na molo wychodziło się z niego na końskie doły} ---
\textit{instead of onto a pier you stepped out onto horse pits}.
\textbf{Molo} --- \textit{pier, jetty} --- Sopot's defining feature.
\textbf{Końskie doły} --- \textit{horse pits}: excavation pits named
for the horses that hauled extracted material. The same architectural
grandeur, a completely different view from the terrace.

\textbf{dawne wyrobiska po piasku do kopalni} --- \textit{former sand
extraction workings for the mine}. Sand mined industrially for the
coal mine. Silesian landscape in miniature.

\textbf{O ile się wychodziło} --- \textit{if you stepped out at all}
--- the parenthesis closes on this. The grandeur of the view was, in
any case, largely theoretical for the hospital's residents. The
architecture of aspiration and the landscape of extraction --- the same
Silesian juxtaposition the cemetery survey observed --- and in the
middle of it, Adam, brilliant and lost.

%% ---------------------------------------------------------------
\section{Marzena Orzeł}
%% ---------------------------------------------------------------

\poltext{Idziemy dalej. Grób Marzeny Orzeł. Rak mózgu. Umarła
młodo. Na pogrzebie w odkrytej trumnie leżała z głową w
bandażach. Była inna, nierzeczywista, uroczysta i dostojna.}

\poltrans{We move on. The grave of Marzena Orzeł. Brain cancer.
She died young. At the funeral she lay in an open coffin with her
head in bandages. She was different, unreal, solemn and dignified.}

\textbf{Idziemy dalej} --- \textit{we move on}. Present tense,
two words, the \textit{obchód} continuing. The marble ball, the
gorzała, the hypothesis --- left behind. The group moves to the
next grave.

\textbf{Grób Marzeny Orzeł} --- \textit{the grave of Marzena
Orzeł}. Name and surname, no title. \textbf{Orzeł} ---
\textit{eagle} --- one of the more striking Polish surnames, from
the national symbol.

\textbf{Rak mózgu} --- \textit{brain cancer}. Two nouns, no
verb. The cause of death delivered in the same register as
\textit{endoproteza} and \textit{zawał} --- bare, clinical,
complete.

\textbf{Umarła młodo} --- \textit{she died young}. Three words.
\textbf{Młodo} --- \textit{young} --- the adverb, not the
adjective: she died \textit{in a young way}, while young, at a
young age. The adverb makes it feel more like a manner of living
than a fact about timing.

\textbf{Na pogrzebie w odkrytej trumnie leżała z głową w
bandażach} --- \textit{at the funeral she lay in an open coffin
with her head in bandages}. \textbf{Odkryta trumna} ---
\textit{open coffin}: the Polish funeral tradition of an open
casket, common and accepted. \textbf{Z głową w bandażach} ---
\textit{with her head in bandages} --- the surgery, or the
disease's progression, or both, visible even in death. The detail
is stated without flinching.

\textbf{Była inna, nierzeczywista, uroczysta i dostojna} ---
\textit{she was different, unreal, solemn and dignified}. Four
adjectives, each adding a layer. \textbf{Inna} ---
\textit{different, other} --- not herself, or more than herself.
\textbf{Nierzeczywista} --- \textit{unreal, not of this world}
--- from \textit{rzeczywisty} (real, actual); the \textit{nie-}
prefix makes her belong to another register of existence.
\textbf{Uroczysta} --- \textit{solemn, ceremonial} --- the
adjective of \textit{uroczystość} (ceremony): the quality of a
moment set apart from ordinary time. \textbf{Dostojna} ---
\textit{dignified, stately} --- from \textit{dostojność}
(dignity, majesty): the bearing of someone who occupies a
significant station.

\medskip
The four adjectives move from the observer's disorientation
(\textit{inna, nierzeczywista}) toward a recognition of something
in Marzena herself (\textit{uroczysta, dostojna}). The bandages
on her head, which might have seemed a mark of suffering, become
part of what made her \textit{dostojna}. The narrator is
describing what he saw at her funeral --- a memory that has
stayed, precise and intact.

%% ---------------------------------------------------------------
\section{Zamieniona / Transformation}
%% ---------------------------------------------------------------

\poltext{Zamieniona z niepozornej szkolnej koleżanki w obiekt
adoracji i powód lamentu jej rodziny, a jednocześnie w budzące
nasze niezdrowe zainteresowanie martwe zwłoki. Była delegacja ze
szkoły. To była pierwsza nasza klasowa umarła.}

\poltrans{Transformed from an unremarkable schoolmate into an
object of adoration and the cause of her family's lamentation,
and at the same time into a corpse that aroused our morbid
curiosity. There was a school delegation. She was the first of
our class to die.}

\textbf{Zamieniona z niepozornej szkolnej koleżanki} ---
\textit{transformed from an unremarkable schoolmate}.
\textbf{Zamieniona} --- passive participle of \textit{zamienić}
(to transform, to exchange): she has been changed into something
else. \textbf{Z} + genitive marks the starting point of the
transformation. \textbf{Niepozorna} --- \textit{unremarkable,
unassuming, nondescript} --- from \textit{pozorny} (conspicuous)
with negating \textit{nie-}: someone who didn't particularly
stand out. In life she was \textit{niepozorna}; in death she
became something else entirely.

\textbf{w obiekt adoracji i powód lamentu jej rodziny} ---
\textit{into an object of adoration and the cause of her
family's lamentation}. \textbf{Obiekt adoracji} --- the same
\textit{obiekt} used for Adam and Robert as \textit{obiekty
podziwu}: the clinical noun for a thing that is looked at,
focused upon. In death she became the centre of her family's
grief and worship, elevated to a status she didn't hold in life.
\textbf{Lament} --- \textit{lamentation} --- biblical in
register, the formal cry of the bereaved.

\textbf{a jednocześnie w budzące nasze niezdrowe zainteresowanie
martwe zwłoki} --- \textit{and at the same time into a corpse
that aroused our morbid curiosity}. \textbf{Jednocześnie} ---
\textit{simultaneously} --- the transformation is double: she is
both the \textit{obiekt adoracji} for her family and the
\textbf{martwe zwłoki} (corpse --- the same \textit{zwłoki} from
the chapter's opening, Fiszer's widow transporting them from
France) for the school delegation. \textbf{Niezdrowe
zainteresowanie} --- \textit{morbid, unhealthy curiosity} ---
literally \textit{unhealthy}: the narrator does not flatter
himself or his classmates. They came to the open coffin drawn by
something they recognise as not entirely wholesome. Children at
their first encounter with death.

\textbf{Była delegacja ze szkoły} --- \textit{there was a school
delegation}. Impersonal past. A formal group sent to represent
the school at the funeral --- a common Polish practice.

\textbf{To była pierwsza nasza klasowa umarła} --- \textit{she
was the first of our class to die}. \textbf{Klasowa umarła} ---
\textit{the class's dead girl} --- the feminine form of
\textit{umarły}, an adjective used as a noun. Earlier we had
\textit{klasowi umarli} (the class dead, plural, masculine); here
the singular feminine form, the first one, Marzena. The chapter
has been about a group of people who have accumulated dead; here
is where it began, the first death, the one that opened the
account. All the deaths that followed --- Fiszer, Robert, Adam,
and however many others the \textit{obchód} will yet reveal ---
were preceded by this one.

%% ---------------------------------------------------------------
\section{Gutek}
%% ---------------------------------------------------------------

\poltext{--- A co stało się z Guciem? --- pytam, przypominając
sobie szkolnych kolegów.}

\poltrans{``And what happened to Gutek?'' --- I ask, calling to
mind my school friends.}

\textbf{A co stało się z Guciem?} --- \textit{and what happened
to Gutek?} \textbf{Stało się} --- \textit{happened, occurred}
--- the same impersonal \textit{stać się} from \textit{co ci się
stało w nogę} earlier, but here without the dative of the
affected person: just \textit{what happened}, free-floating.
\textbf{Guciem} --- instrumental of \textit{Gutek}, itself a
diminutive of \textit{August} or \textit{Gustaw}: a boyhood
nickname, the kind that persists.

\textbf{pytam} --- \textit{I ask} --- present tense, not past.
A shift: the narrator has been using past tense throughout, but
here \textit{pytam} --- present, as though we are watching him
ask it in real time, the question forming at the grave, in the
moment of telling.

\textbf{przypominając sobie szkolnych kolegów} --- \textit{calling
to mind my school friends}. \textbf{Przypominać sobie} ---
\textit{to recall, to call to mind} --- the reflexive
\textit{sobie} marking that the recollection is internal, his
own. \textbf{Szkolnych kolegów} --- \textit{school friends},
genitive plural: the whole class, mentally assembled. He is
running through them, and Gutek is the one whose name surfaces
with a question mark attached.

\medskip
The question is the most open one in the chapter so far --- not
\textit{dlaczego} (why) but \textit{co stało się} (what
happened): Gutek's fate is entirely unknown to the narrator.
After Groszek's precise knowledge of Robert's statistics and
Adam's polytechnic record, he finds himself with a name and no
information. The \textit{obchód} is also a survey of what is
known and what has been lost to silence.

%% ---------------------------------------------------------------
\section{Groszek odpowiada / Groszek Answers}
%% ---------------------------------------------------------------

\poltext{--- Jego ojciec był wojskowym, mieszkali w wojskowych
blokach. Wyjechali potem gdzieś pod granicę. Nie wiem którą.
Zapił się podobno. Podobno. Ale może chodziło o kogoś innego
--- Groszek nie jest pewny.}

\poltrans{``His father was in the military, they lived in the
military blocks. They moved away somewhere near the border
afterwards. I don't know which one. He drank himself to ruin
apparently. Apparently. But maybe it was someone else'' ---
Groszek isn't sure.}

\textbf{Jego ojciec był wojskowym} --- \textit{his father was a
military man}. \textbf{Wojskowy} --- \textit{military man,
serviceman} --- used as a noun in the instrumental after
\textit{był}. A single defining fact about Gutek's family,
supplied by way of context.

\textbf{mieszkali w wojskowych blokach} --- \textit{they lived
in the military blocks}. \textbf{Bloki} --- the standard Polish
word for apartment blocks, the large concrete residential
buildings of the communist era. \textbf{Wojskowych} ---
\textit{military}: purpose-built housing for military families,
a self-contained world within the town.

\textbf{Wyjechali potem gdzieś pod granicę} --- \textit{they
moved away somewhere near the border afterwards}.
\textbf{Wyjechali} --- \textit{they left, they moved away} ---
perfective, a departure completed. \textbf{Gdzieś} ---
\textit{somewhere} --- the vagueness of half-remembered
information. \textbf{Pod granicę} --- \textit{near the border,
toward the border} --- accusative after \textit{pod} expressing
direction or proximity.

\textbf{Nie wiem którą} --- \textit{I don't know which one}.
Groszek's honesty about the limits of his knowledge, offered
without embarrassment. Four words and the information stops
there.

\textbf{Zapił się podobno} --- \textit{he drank himself to ruin
apparently}. \textbf{Zapić się} --- \textit{to drink oneself
into destruction}: the prefix \textit{za-} marks a process
carried to its ruinous completion. Not merely \textit{pił} (he
drank) but \textit{zapił się} --- he drank until there was
nothing left. \textbf{Podobno} --- \textit{apparently,
reportedly} --- the same particle of unverified rumour from
\textit{podobno na serce}. What Groszek has heard, not what he
knows.

\textbf{Podobno.} --- \textit{Apparently.} Alone, the word
repeated as its own sentence. A pause. The repetition signals
that Groszek himself is uncertain whether to credit it --- the
second \textit{podobno} questions the first.

\textbf{Ale może chodziło o kogoś innego} --- \textit{but maybe
it was someone else}. \textbf{Chodziło o} --- \textit{it was
about, it concerned} --- the impersonal construction. The rumour
of drinking may belong to a different person entirely, confused
with Gutek in the passage of years. Memory doing what memory
does.

\textbf{Groszek nie jest pewny} --- \textit{Groszek isn't sure}.
Present tense, the narrator stepping back to observe.
\textbf{Pewny} --- \textit{certain, sure}. After Robert's seven
appearances in the \textit{ekstraklasa} and Adam's polytechnic
mathematics, Groszek's certainty runs out at Gutek. He has a
fragment --- military father, border town, a rumour of drink ---
and the honesty to say it might not even be right. Memory
reliable here, vague there; precise on football statistics,
unreliable on a childhood friend who moved away. The repeated
\textit{podobno} is a small masterpiece: the word meaning
\textit{apparently} appearing twice, the second time apparently
doubting the first.

%% ---------------------------------------------------------------
\section{Gorol}
%% ---------------------------------------------------------------

\poltext{--- A Gorol? --- przypominam sobie ksywę dawnego osiłka
o ospowatej twarzy.}

\poltrans{``And Gorol?'' --- I recall the nickname of a former
muscleman with a pockmarked face.}

The narrator now produces a name himself, without prompting. The
\textit{obchód} continues its roll call, but the direction has
reversed: no longer asking the group, he is surfacing memories
from inside himself.

\textbf{A Gorol?} --- the pivot. \textbf{Gorol} (also
\textit{górol}) is the Silesian term for an outsider: someone
from outside Silesia, from the mountains or from central Poland,
who has come to live among the locals. All such regional
designations carry the same dual charge --- descriptive on the
surface, exclusionary underneath. That this was a school
\textit{ksywa} tells us the boy was marked as not-from-here, and
that the marking was adhesive enough to replace whatever his
actual name was. He moved through their world as \textit{Gorol}
--- \textit{The Outsider}.

\textbf{przypominam sobie} --- \textit{I recall, I call to mind}
--- present tense, as throughout this sequence. The narrator's
memory is an active process, unfolding visibly in the present
moment of the grave visit.

\textbf{ksywę} --- accusative of \textit{ksywa}, the colloquial
and slightly underworld-flavoured word for a nickname or alias.
More charged than the neutral \textit{przezwisko}; a
\textit{ksywa} is assigned from outside and sticks. Here it
performs an irony: the name that names the nameless.

\textbf{dawnego osiłka} --- \textit{of a former muscleman}.
\textbf{Osiłek}: a big, strong, physically imposing lad ---
bulk rather than grace. The diminutive suffix \textit{-ek} gives
it a rough affection rather than contempt. \textbf{Dawnego}
--- \textit{former, of former times} (from \textit{dawno},
long ago) --- this is who he was then, in the school
years; what he became since is unspoken.

\textbf{o ospowatej twarzy} --- \textit{with a pockmarked face}.
\textbf{Ospowaty}: marked by \textit{ospa} (smallpox,
chickenpox), the scarring left by the disease. A single precise
physical detail, recovered across decades. The face returns
before the person --- or rather, instead of the person, since
the person was always, only, \textit{Gorol}.

\medskip
\noindent\textit{Grammar note:} \textbf{o ospowatej twarzy} is
an instrumental phrase of description: \textit{o} + instrumental
= ``with (a feature)'', the standard construction for
characterising someone by a physical attribute. The same
structure governs \textit{chłopak o niebieskich oczach} (a boy
with blue eyes) throughout Polish narrative prose.

%% ---------------------------------------------------------------
\section{Na dobrej drodze / On the Right Track}
%% ---------------------------------------------------------------

\poltext{--- Ten chyba jeszcze się nie zapił. Ale jest na dobrej
drodze. Możesz go spotkać na osiedlu pod sklepem. Razem z
innymi: Dyjakiem, Maligłówką\ldots{} Pamiętasz ich jeszcze?}

\poltrans{``This one probably hasn't drunk himself to ruin yet.
But he's on the right track. You can find him on the estate by
the shop. Together with the others: Dyjak, Maligłówka\ldots{}
Do you still remember them?''}

Groszek corrects the conversation's direction. The \textit{zapił
się} story belonged to someone else; Gorol is alive and
locatable --- though the news is not straightforwardly good.

\textbf{Ten chyba jeszcze się nie zapił} --- \textit{this one
probably hasn't drunk himself to ruin yet}. \textbf{Ten} ---
\textit{this one} --- the demonstrative used familiarly, almost
dismissively, referring back to Gorol. \textbf{Chyba} ---
\textit{probably, I think} --- Groszek's characteristic hedging
particle. \textbf{Jeszcze} --- \textit{yet, still} --- doing
quiet work: not \textit{didn't} but \textit{hasn't yet}, leaving
the future pointedly open. The same \textit{zapić się} from the
Gutek passage, now negated --- the word connects the two men:
one apparently did, this one apparently hasn't. Yet.

\textbf{Ale jest na dobrej drodze} --- \textit{but he's on the
right track}. A piece of bitter irony. \textbf{Dobra droga} ---
\textit{good road, right path} --- an idiom that normally signals
someone is on track for success. Here Groszek deploys it to mean
the opposite: Gorol is on track toward the same ruin. The
compliment form delivering the insult content.

\textbf{Możesz go spotkać na osiedlu pod sklepem} --- \textit{you
can find him on the estate by the shop}. \textbf{Możesz} ---
\textit{you can} --- second person singular, addressing the
narrator directly. \textbf{Na osiedlu} --- \textit{on the
estate}, locative of \textit{osiedle}: the housing estate, the
\textit{bloki}. \textbf{Pod sklepem} --- \textit{by the shop}
--- the classic post-communist gathering point, the patch of
ground outside the local \textit{sklep} where men congregate.
The address is precise without being an address: everyone knows
where \textit{pod sklepem} is.

\textbf{Razem z innymi} --- \textit{together with the others}.
Not alone. There is a group there, a cohort, a recognisable
social formation.

\textbf{Dyjakiem, Maligłówką} --- two names in the instrumental
after \textit{z} (with). Both clearly nicknames.
\textbf{Dyjak} --- obscure origin, possibly from \textit{diak}
(deacon) or a regional coinage. \textbf{Maligłówka} ---
transparent: \textit{mała główka} compressed --- \textit{little
head}, a physical nickname. The instrumental signals membership
of the \textit{razem z} group: Gorol can be found among them.

\textbf{Pamiętasz ich jeszcze?} --- \textit{do you still
remember them?} A sudden turn toward the narrator --- and toward
the reader. \textbf{Jeszcze} again: \textit{still}, implying
time has passed and forgetting is natural. Groszek tests whether
the Warsaw man's memory extends to these peripheral figures: not
the dead, but the living-and-lost, the ones drinking
\textit{pod sklepem}.

\medskip
The passage quietly redraws the social map. There is a category
the \textit{obchód} hasn't yet reckoned with: those who aren't
dead but who have come to rest in a different kind of permanence
--- the estate bench, the shop doorway, the daily bottle. Not a
grave to visit, but a spot where you can still find them. Gorol
is alive; he just hasn't gone anywhere.

%% ---------------------------------------------------------------
\section{To nie Gorol / That's Not Gorol}
%% ---------------------------------------------------------------

\poltext{--- To nie Gorol --- zaprzecza Hanka. --- Gorol już tu
dawno nie mieszka.\\
--- A kto? --- pyta Groszek.\\
--- A bo ja wiem? Wszyscy są do siebie podobni.}

\poltrans{``That's not Gorol,'' Hanka contradicts. ``Gorol
hasn't lived here for a long time.''\\
``Then who?'' asks Groszek.\\
``How would I know? They all look alike.''}

A miniature comedy of errors, three lines, and then a sentence
that lands with uncomfortable weight.

\textbf{To nie Gorol} --- \textit{that's not Gorol}. Flat,
immediate contradiction. Hanka has been listening, and she knows
something Groszek doesn't. \textbf{To} --- \textit{that, it}
--- the blunt demonstrative.

\textbf{zaprzecza Hanka} --- \textit{Hanka contradicts}.
\textbf{Zaprzeczać} --- \textit{to contradict, to deny} --- more
formal than simply \textit{mówi} (says); it registers that she
is specifically negating what was just said. Hanka has had
quieter moments so far --- lighting candles, producing
cigarettes --- and here she cuts in with authority.

\textbf{Gorol już tu dawno nie mieszka} --- \textit{Gorol hasn't
lived here for a long time}. \textbf{Już} --- \textit{already,
no longer} --- paired with negation it means \textit{no longer}:
he has already departed, and the departure is settled.
\textbf{Tu} --- \textit{here}. \textbf{Dawno} --- \textit{long
ago, for a long time} --- his absence isn't recent. Groszek's
confident \textit{możesz go spotkać na osiedlu pod sklepem} was
wrong: the intelligence is out of date, or it was never Gorol
to begin with.

\textbf{A kto?} --- \textit{then who? so who is it?} Groszek's
response, two words. \textbf{A} --- the pivoting \textit{a}
again, now expressing genuine bafflement. If it's not Gorol
\textit{pod sklepem}, who is Groszek thinking of?

\textbf{pyta Groszek} --- \textit{Groszek asks}. Present tense
throughout this exchange --- the dialogue is immediate, alive.

\textbf{A bo ja wiem?} --- \textit{how would I know?} One of
the most characteristically Polish rhetorical constructions.
\textbf{A bo} --- a colloquial particle combination expressing
mild exasperation or rhetorical deflection. \textbf{Ja wiem}
--- \textit{I know} --- but as a question it means the opposite:
\textit{I don't know, and why would you expect me to?} A shrug
rendered as syntax.

\textbf{Wszyscy są do siebie podobni} --- \textit{they all look
alike, they all resemble each other}. \textbf{Do siebie podobni}
--- \textit{similar to each other, resembling one another} ---
the reflexive \textit{siebie} making it mutual: each resembles
all the others. \textbf{Wszyscy} --- \textit{everyone, all of
them}.

\medskip
The sentence is the passage's sting. Hanka is speaking about
the men \textit{pod sklepem} --- the Gorols, the Dyjaks, the
Maligłówkas of the estate --- and her words carry the flat
dismissal of someone who has watched this happen too many times
to distinguish between individuals. They've blurred into a
single type: the man by the shop, the bottle, the bench, the
slow ruin. Not cruel so much as exhausted.

\medskip
\noindent\textit{Vocabulary note:} \textbf{Podobni} connects
back to the chapter's recurring \textit{podobno} (apparently,
reportedly) and to the earlier motif of recognition and
resemblance --- \textit{podobny} (similar, resembling) threads
through the text's concern with who can be told apart from whom,
who is remembered as a person and who becomes a category.

%% ---------------------------------------------------------------
\section{Szalony Grzesio / Crazy Grzesio}
%% ---------------------------------------------------------------

\poltext{--- Grzesio też z nimi stoi? --- jestem ciekaw. ---
Wiecie, Szalony Grzesio. Piroman. Na parterze mieszkał. Kiedyś
podpalił drzwi Aście.}

\poltrans{``Is Grzesio standing with them too?'' I'm curious.
``You know --- Crazy Grzesio. Pyromaniac. He lived on the ground
floor. He once set fire to Aśka's door.''}

The narrator steps into the conversation for the first time ---
not to report it, but to participate. And characteristically,
he asks the question and then immediately answers it himself,
supplying Grzesio's dossier in four fragments before anyone
else can respond.

\textbf{Grzesio też z nimi stoi?} --- \textit{Is Grzesio
standing with them too?} \textbf{Grzesio} --- diminutive of
\textit{Grzegorz} (Gregory), affectionate and familiar.
\textbf{Też} --- \textit{too, also} --- extending the roll
call: Dyjak, Maligłówka\ldots{} and now, possibly, Grzesio.
\textbf{Z nimi stoi} --- \textit{standing with them}: \textit{z}
+ instrumental (\textit{nimi}), the same construction as
\textit{razem z innymi} in the previous passage. The gathering
\textit{pod sklepem} has its own grammar: you stand \textit{z}
someone, not merely near them.

\textbf{jestem ciekaw} --- \textit{I am curious, I wonder}.
Present tense breaking into the past-tense narration --- the
narrator's living curiosity surfacing mid-sentence, as if the
memory is being worked through in real time. Not \textit{byłem
ciekaw} (I was curious) but \textit{jestem}: he wonders now,
telling us.

\textbf{Wiecie} --- \textit{you know, you see}. Second person
plural, addressing Hanka and Groszek simultaneously --- but
the plural also catches the reader. A narrator's aside, the
\textit{you know} of a storyteller who remembers he has an
audience.

\textbf{Szalony Grzesio} --- \textit{Crazy Grzesio}. From
\textit{szaleć} --- to be mad, to rage, to run wild.
\textbf{Szalony} as a nickname is an accolade of a particular
kind: not dangerous-mad but memorable-mad, the sort of
\textit{szaleniec} who does things others only think about.
The name precedes any explanation and makes explanation both
necessary and superfluous.

\textbf{Piroman} --- \textit{pyromaniac}. One word, unglossed,
delivered like a card from a file. The international form
(from Greek \textit{pyr}, fire) lends it a clinical flavour
that the nickname entirely undercuts. Together they form a
complete portrait: \textit{Szalony} from the inside,
\textit{Piroman} from the outside.

\textbf{Na parterze mieszkał} --- \textit{he lived on the
ground floor}. \textbf{Parter} --- borrowed from French
\textit{par terre}, on the ground --- the ground floor of a
\textit{blok}. Locative: \textit{na parterze}. The detail is
architectural and tactical: ground floor, close to everyone
else's door. The placement sets up what follows.

\textbf{Kiedyś podpalił drzwi Aście} --- \textit{he once set
fire to Aśka's door}. \textbf{Kiedyś} --- \textit{once, at
some point in the past} --- indefinite, which gives the act
a slightly legendary quality: it happened, it is established,
the exact when has dissolved. \textbf{Podpalił} --- perfective
past of \textit{podpalić}: to set fire to, to ignite (with
intent). The \textit{pod-} prefix implies application from
below or close contact --- not an accidental blaze but a
deliberate act at ground level. \textbf{Drzwi} --- \textit{door}
--- a plural-only noun in Polish, neuter. \textbf{Aście} ---
dative of \textit{Aśka}, a diminutive of \textit{Asia} (itself
a diminutive of \textit{Joanna}).

\medskip
\noindent\textit{Grammar note:} \textbf{Podpalił drzwi Aście}
uses the dative of personal interest rather than the genitive
of possession. One could say \textit{podpalił drzwi Aśki}
(genitive: Aśka's door, mere ownership), but \textit{drzwi
Aście} places the harm on Aśka directly: the dative signals
that the act was done \textit{to her}, not just to her
property. It is a small grammatical intensification that
makes the victim present in the sentence.

\medskip
The micro-portrait is complete in four beats: nickname,
category, location, defining act. This is how the \textit{obchód}
works as a narrative method --- not biography but evidence,
the minimum required to fix a person in memory. Grzesio
is given no face, no family, no further history. He has a
name everyone already knows, a word that explains it, a floor,
and a fire. That is enough to bring him back.

%% ---------------------------------------------------------------
\section{Znikali po kolei / They Disappeared One by One}
%% ---------------------------------------------------------------

\poltext{--- Coś ty? To ty nie wiesz? --- Groszek kręci
z dezaprobatą głową. --- Został dyrektorem ekonomicznym huty.
Ale teraz jest na rencie.\\
\hspace*{1em} A inni? Znikali po kolei. Jeden po drugim.}

\poltrans{``What? You don't know?'' Groszek shakes his head
with disapproval. ``He became the economic director of the
smelter. But now he's on disability pension.''\\
\hspace*{1em} And the others? They disappeared one by one.
One after another.}

Groszek's disbelief at the narrator's ignorance, a quick
biographical reversal for Grzesio, and then --- unprompted,
unmarked by any speaker attribution --- the chapter's pivot.

\textbf{Coś ty?} --- a compressed expression of startled
disbelief, almost untranslatable literally. The \textbf{coś}
is an archaic or colloquial contraction of \textit{co} (what)
with the second-person auxiliary \textit{ś} (from
\textit{jesteś}): roughly \textit{what are you}. As an
exclamation it means something between \textit{are you
serious?}, \textit{what's wrong with you?}, and \textit{come
off it} --- a reflexive noise of disbelief before the actual
sentence arrives.

\textbf{To ty nie wiesz?} --- \textit{so you don't know?}
\textbf{To} pivots, as it so often does in Polish dialogue:
\textit{in that case, then}. The stress falls on \textbf{ty}
--- \textit{you}, specifically you, the man from Warsaw who
has lost touch. The question is a mild rebuke dressed as
an inquiry.

\textbf{Groszek kręci z dezaprobatą głową} --- \textit{Groszek
shakes his head with disapproval}. \textbf{Kręcić głową} ---
\textit{to shake one's head} --- is the standard idiom:
\textit{kręcić} (to turn, to rotate) + \textbf{głową}
(instrumental of \textit{głowa}). The instrumental is
required: one shakes \textit{by means of} one's head, as
with all body-part gestures in Polish. \textbf{Z dezaprobatą}
--- \textit{with disapproval} --- instrumental of
\textit{dezaprobata}, a formal Latinate word inserted into
this entirely colloquial exchange. The elevated noun lands
slightly dryly: Groszek's gesture is not merely a shake
but a registered verdict.

\textbf{Został dyrektorem ekonomicznym huty} --- \textit{he
became the economic director of the smelter}.
\textbf{Zostać} + instrumental is the standard construction
for becoming something: \textit{dyrektorem} and
\textit{ekonomicznym} both in the instrumental, agreeing
with each other. \textbf{Huty} --- genitive of
\textbf{huta} --- the smelter, the steelworks.

\medskip
\noindent\textit{Context note:} This is the first appearance
of \textbf{huta} as a named institution rather than a backdrop.
The book's title is \textit{Ołowiane dzieci} --- \textit{Lead
Children} --- and the \textit{huta} is the lead-smelting
complex whose contamination the book investigates. Grzesio,
the pyromaniac of the ground floor, became its economic
director. The \textit{obchód} has just produced its first
piece of evidence.

\textbf{Ale teraz jest na rencie} --- \textit{but now he's
on disability pension}. \textbf{Renta} is a disability or
invalidity pension, distinct from \textit{emerytura}
(retirement pension earned by age). \textbf{Na rencie} ---
locative, the standard phrase: one is \textit{on} a pension
as if resting upon it. The \textbf{ale} (but) performs the
chapter's characteristic move: a career peak, then the
counterweight. Director of the smelter --- now on
disability. The two facts sit next to each other without
comment, and the comment is unnecessary.

\medskip

\textbf{A inni?} --- \textit{and the others?} No speaker
is named. This is either Groszek continuing, or the narrator
stepping back to address us directly, or both at once ---
the transition is deliberately unmarked. \textbf{A} performs
its widening function: having accounted for Grzesio, the
gaze pulls back to the whole cohort.

\textbf{Znikali po kolei} --- \textit{they disappeared one
by one, in turn}. \textbf{Znikać} --- imperfective,
\textit{to disappear, to vanish} --- and the imperfective
past \textbf{znikali} is important: not a single collective
event but a repeated, ongoing process, each disappearance
separate, the same thing happening again and again.
\textbf{Po kolei} --- \textit{in turn, one after another,
in sequence} --- an adverbial phrase of orderly succession.
The order was not random; there was a sequence, a pattern.

\textbf{Jeden po drugim} --- \textit{one after another}.
Where \textit{po kolei} describes the pattern, \textit{jeden
po drugim} insists on the arithmetic: one, then the next
one, then the next. The sentence is a repetition and an
intensification simultaneously.

\medskip
The final two sentences are the chapter's turn. Everything
before this point has been the \textit{obchód}: individual
portraits, nicknames, brief flares of recovered memory,
the dead distinguished from the merely-lost. Now the
individuals dissolve into a movement. \textit{Znikali} ---
they vanished, repeatedly, one after another --- and the
imperfective aspect makes it a process still visible in
the narration, not a completed fact but something the
speaker watched happening over years.

This is the book's first statement of its subject. The
lead smelter has been named. The economic director is on
disability pension. The others disappeared in sequence.
The word \textit{epidemia} in the subtitle now has its
grammar: not sudden, not simultaneous, but \textit{po
kolei} --- methodical, patient, one by one.

%% ---------------------------------------------------------------
\section{Z naszej półki / From Our Shelf}
%% ---------------------------------------------------------------

\poltext{--- Coś wyjątkowo często sięgają z naszej półki ---
zauważam.}

\poltrans{``They seem to reach from our shelf rather
exceptionally often,'' I notice.}

One sentence, one metaphor, one verb of noticing. It is the
narrator's first full contribution to the conversation ---
and it is the most compressed statement of the book's premise
that has appeared so far.

\textbf{Coś} --- here not the exclamatory \textit{coś ty?}
of the previous passage but the adverbial \textit{coś}: a
hedging intensifier meaning \textit{somehow, rather, a bit,
strangely}. \textit{Coś} in this function softens or
qualifies what follows --- \textit{coś za drogo} (a bit too
expensive), \textit{coś mi się wydaje} (it seems to me
somehow). Here it gives the observation a deliberately
understated, almost tentative tone: not \textit{they are
dying far too often} but \textit{somehow they seem to be
reaching from our shelf rather a lot}. The hedging is
ironic; the observation is devastating.

\textbf{wyjątkowo często} --- \textit{exceptionally often,
with unusual frequency}. \textbf{Wyjątkowo} --- from
\textit{wyjątek} (exception) --- means \textit{exceptionally,
unusually, out of the ordinary}. Paired with \textbf{często}
(often), it builds to: \textit{with a frequency that itself
is exceptional}. The two adverbs together form the
statistical claim the narrator will not make explicitly.

\textbf{sięgają z naszej półki} --- \textit{they reach from
our shelf}. \textbf{Sięgać} --- \textit{to reach, to
stretch for something} --- combined with \textbf{z} (from)
and \textbf{półki} (genitive of \textit{półka}, shelf):
\textit{to take from a shelf}, as one pulls a book.
The agent of \textbf{sięgają} is unstated --- the third
person plural without a subject, the Polish impersonal
construction: \textit{they reach, one reaches, someone
reaches}. Death, fate, whatever force removes people, is
present only as a grammatical third person who never
appears in the nominative.

\medskip
\noindent\textit{The metaphor:} \textbf{Półka} --- a shelf,
specifically a library shelf or a storage shelf --- turns
a generation into an inventory. The dead are not lost or
taken or fallen; they are \textit{reached for}, selected,
lifted from a particular section. \textit{Nasza półka}
--- \textit{our shelf} --- is the cohort: their year,
their estate, their school. Someone is browsing that
section with notable frequency. The image is almost
bureaucratic in its tidiness, which is precisely what
makes it chilling. It also carries a faint absurdist
gallows-humour that is entirely consistent with the book's
voice: the narrator noticing, as if commenting on a
library's circulation statistics, that books from one
particular shelf keep going missing.

\textbf{zauważam} --- \textit{I notice, I observe}.
Present tense, first person singular, set apart after the
dash as a speech-tag turned inward. Not \textit{mówię}
(I say) or \textit{dodaję} (I add) but \textit{zauważam}
--- \textit{I notice}: the verb of careful, slightly
clinical attention. The same present tense as
\textit{jestem ciekaw} two passages ago --- the narrator's
consciousness keeps surfacing in the present, thinking
through memory in real time. He is not merely recounting
the cemetery walk; he is arriving at something.

\medskip
With this sentence the \textit{obchód} completes its turn.
The chapter opened with individual names --- Fiszer,
Gutek, Gorol, Grzesio --- and has moved through the
living-and-lost to the smelter director on disability and
now to this: a pattern, a frequency, a shelf being
visited too often. The narrator has not yet said
\textit{ołów} (lead) or \textit{trucizna} (poison). He
has said \textit{nasza półka}, and let the grammar do
the rest.

%% ---------------------------------------------------------------
\section{Okazy zdrowia / Specimens of Health}
%% ---------------------------------------------------------------

\poltext{Biorąc pod uwagę, że nasza garstka żywych też nie
prezentowała się jak okazy zdrowia, uwaga była jak najbardziej
uzasadniona. Walencki --- endoproteza, Daisy --- choroba
układu odpornościowego zwana też autoagresją, ja ---
przewlekle, nieuleczalnie i postępująco. WTF?}

\poltrans{Taking into account that our handful of survivors
didn't exactly present themselves as specimens of health
either, the observation was more than justified. Walencki
--- joint replacement, Daisy --- an immune system disorder
also known as autoaggression, me --- chronically,
incurably, and progressively. WTF?}

The narrator steps out of the dialogue entirely and
addresses the reader directly, turning the observation
of \textit{nasza półka} onto the living remainder.
The dossier format returns --- three entries, each a
name and a condition --- and then the paragraph ends
with two letters and a punctuation mark that arrives
like a slap.

\textbf{Biorąc pod uwagę, że} --- \textit{taking into
account that, given that}. A formal participial opener ---
the same register as \textit{zważywszy, że} in the book's
first paragraph, used there to ironise children smoking in
bushes. Here it frames a medical audit. \textbf{Biorąc}
is the imperfective active participle of \textit{brać}
(to take); \textbf{pod uwagę} --- \textit{into
consideration, under attention} --- a fixed phrase:
literally \textit{under attention}. The bureaucratic
construction applied to survival statistics.

\textbf{nasza garstka żywych} --- \textit{our handful of
the living}. \textbf{Garstka} --- diminutive of
\textit{garść} (a handful, a fistful) --- conveys both
smallness and the cupped-hand image: what remains is
what fits in a palm. \textbf{Żywych} --- genitive plural
of \textit{żywy} (alive, living) --- the living as a
countable, diminishing category. Placed against
\textit{znikali po kolei} of the previous passage, the
arithmetic is visible: many have gone; a handful remain;
and the handful is not well.

\textbf{nie prezentowała się jak okazy zdrowia} ---
\textit{didn't present itself as specimens of health}.
\textbf{Prezentować się} --- \textit{to present oneself,
to look, to appear} --- slightly formal, almost clinical.
\textbf{Okazy zdrowia} --- \textit{specimens of health},
genitive plural: the phrase \textit{okaz zdrowia} (a
specimen or exemplar of health) is common in Polish,
typically used ironically or in negation, as here.
\textbf{Okaz} (specimen, exemplar) from biology and
natural history --- the word used for museum exhibits,
pinned butterflies, things that are preserved and
labelled. The narrator is not using it neutrally.

\textbf{uwaga była jak najbardziej uzasadniona} ---
\textit{the observation was more than justified, entirely
warranted}. \textbf{Jak najbardziej} --- literally
\textit{as most as possible}, idiomatically \textit{entirely,
absolutely, most definitely} --- a Polish intensifier of
agreement or confirmation. \textbf{Uzasadniona} ---
\textit{justified, warranted, well-founded} --- from
\textit{zasada} (principle, basis). The \textit{uwaga}
(observation, remark) refers back to his own
\textit{zauważam} of the previous sentence: the noticing
was not idle; the data bear it out.

\medskip

The three entries arrive in the same compressed dossier
format as the earlier nicknames --- name, dash, condition:

\textbf{Walencki --- endoproteza} --- \textit{Walencki
--- joint replacement}. \textbf{Endoproteza} ---
an endoprosthesis, an internal prosthetic implant,
most commonly a hip or knee replacement. The medical
term, unglossed and unadorned, in the place where a
nickname epithet would have stood. Walencki is defined
by his replacement joint the way Grzesio was defined
by his fire.

\textbf{Daisy --- choroba układu odpornościowego zwana
też autoagresją} --- \textit{Daisy --- an immune system
disease also called autoaggression}. A longer entry:
\textbf{układ odpornościowy} --- \textit{immune system}
(literally \textit{defensive system}); \textbf{zwana też}
--- \textit{also known as} --- the passive participle
\textit{zwana} (called, named), feminine to agree with
\textit{choroba}. \textbf{Autoagresja} --- the Polish
term for autoimmune disease: \textit{autoaggression},
the body attacking itself. The word is vivid in a way
the English clinical term ``autoimmune'' is not ---
the self turning against the self, aggression directed
inward. And the name \textbf{Daisy} stands out sharply
among the Polish names: English, feminine, incongruous
--- which is presumably why she was called it.

\textbf{ja --- przewlekle, nieuleczalnie i postępująco}
--- \textit{me --- chronically, incurably, and
progressively}. The narrator's own entry. Where the
others receive a named condition, he gives only three
adverbs --- the manner of the illness, not its name.
\textbf{Przewlekle} --- \textit{chronically}, from
\textit{przewlekły} (protracted, lingering).
\textbf{Nieuleczalnie} --- \textit{incurably}, from
\textit{leczyć} (to treat, to cure) with the negating
\textit{nie-} prefix. \textbf{Postępująco} ---
\textit{progressively}, from \textit{postępować}
(to advance, to proceed) --- a medical term for a
condition that worsens over time. Three adverbs in
ascending dread. He does not name the disease; the
adverbs are enough, and more intimate for the
withholding.

\medskip
\noindent\textit{Grammar note:} The three entries use
the dash construction (\textbf{Walencki --- endoproteza})
in place of any verb --- no \textit{ma} (has), no
\textit{jest} (is). The nominal predicate without
copula is entirely natural in Polish: \textit{on
doktor} (he is a doctor) requires no \textit{jest}
in informal or clipped contexts. Here the omission
performs the clinical shorthand of a medical file:
patient name, dash, diagnosis.

\medskip

\textbf{WTF?} The paragraph's final word --- or rather
its final three letters. The English internet acronym,
unpacked in any language as an expression of stupefied
disbelief, lands in the middle of formal Polish prose
with complete deliberateness. After \textit{biorąc pod
uwagę}, \textit{uzasadniona}, \textit{endoproteza},
\textit{autoagresją} --- the Latinate, clinical,
bureaucratic register collapses into two decades of
screen-mediated exasperation. The effect is not cheap:
the acronym does what none of the formal vocabulary
could, which is to express the narrator's actual
response to the data he has just presented. The medical
terms described; \textit{WTF?} reacts. It is also, of
course, funny --- and the humour is load-bearing.
A man who can write this sentence about his own
incurable condition is telling you something about
how he intends to survive the telling of this story.

