% ============================================================
%  ch02_ostatni_dzien_wakacji.tex  —  Chapter 2
%  Ostatni dzień wakacji. Osiedle na Burowcu (pp. TBC)
% ============================================================
%
% ── STYLE RULES FOR THIS FILE ───────────────────────────────
%
%  These rules must be followed consistently throughout.
%  Drift from these rules has occurred before — check carefully.
%
%  HEADINGS:
%  - Use \section{} for ALL passage headings without exception.
%  - \subsection{} and \subsection*{} are NOT used in this file.
%  - Section titles follow the pattern:
%      Polish title / English title   (e.g. \section{Obchód / The Rounds})
%    or just a name if self-evident   (e.g. \section{Groszek})
%  - \section{} renders in blue bold with a rule beneath
%    (defined in olowiane_dzieci_analiza.tex).
%
%  POLISH SOURCE TEXT:
%  - Always use \poltext{...} — never \begin{quote}...\end{quote}
%    and never plain \textit{...} or any other wrapper.
%  - \poltext{} renders the text in red italic (keywordred).
%
%  ENGLISH TRANSLATION:
%  - Always use \poltrans{...} immediately after \poltext{}.
%  - \poltrans{} renders in blue (polishblue) with a rule beneath.
%  - No label precedes the translation — colour contrast suffices.
%
%  ANALYSIS NOTES:
%  - Follow \poltrans{} with free prose commentary.
%  - Bold Polish terms with \textbf{}, gloss in \textit{}.
%  - Standard pattern: \textbf{term} --- \textit{gloss} --- analysis.
%  - Use \medskip before thematic summary paragraphs.
%  - Grammar or vocabulary notes may be prefixed with:
%    \noindent\textit{Grammar note:} or \noindent\textit{Vocabulary note:}
%
%  PASSAGE SEPARATORS:
%  - Use % ──────────────────────────────────────────────────
%    (single % with em-dashes) between passages within a section.
%    Do NOT use the %% --- style.
%
%  UPDATE PROTOCOL:
%  - The LaTeX file is updated after each passage is confirmed
%    by the user — not in batches.
%  - Always append new content by str_replace on the final line.
%  - Verify line count after each update with wc -l.
%
% ── SCREENSHOT PROTOCOL ─────────────────────────────────────
%
%  The user supplies passages as screenshots of running book
%  text. Because screenshots capture arbitrary page regions,
%  they often begin or end mid-sentence. The rule is:
%  - Analyse ONLY complete sentences visible in the screenshot
%    (i.e. sentences where both the opening capital and the
%    closing period are present).
%  - Ignore any sentence fragment at the START of a screenshot
%    (already covered by the previous passage).
%  - Ignore any sentence fragment at the END of a screenshot
%    (will be covered in the next passage).
%  - When in doubt, ask the user for the next screenshot
%    rather than guessing at truncated text.
%
% ── WHERE THE READING CURRENTLY STANDS ─────────────────────
%
%  Last passage analysed (session ending 2026-03-07):
%  "Aż zamykano okna." — the poisonous smoke, the four
%  neighbourhoods, and the windows closed against the air.
%  See last \poltext{} block at end of file.
%
%  Chapter 2 is in progress.
%
% ============================================================


\chapter{Ostatni dzień wakacji. Osiedle na Burowcu}
\markboth{Ostatni dzień wakacji. Osiedle na Burowcu}{}

% ── Chapter title ────────────────────────────────────────────
\section{Tytuł rozdziału / Chapter Title}

\poltext{Ostatni dzień wakacji. Osiedle na Burowcu.}
\poltrans{The Last Day of the Holidays. The Estate at Burowiec.}

\textbf{Ostatni dzień wakacji} --- \textit{the last day of
the holidays}. \textbf{Ostatni} --- \textit{last, final} ---
from \textit{ostać się} (to remain); the superlative of
lateness, the day at the end of summer that everyone
who has been a schoolchild knows by feel.
\textbf{Wakacje} --- \textit{school holidays, summer
vacation} --- plural only, like \textit{drzwi} (door)
or \textit{nożyczki} (scissors): summer does not come
in the singular. The word is borrowed from Latin
\textit{vacatio} (freedom, exemption) via the academic
calendar.

\textbf{Osiedle na Burowcu} --- \textit{the estate at
Burowiec}. \textbf{Osiedle} --- the housing estate, the
\textit{bloki}: the word that has governed the spatial
world of Chapter~1. \textbf{Na Burowcu} --- locative
of \textbf{Burowiec}: a specific named estate or district,
\textit{na} indicating location on or at a place
(used with open areas, estates, squares, as distinct
from \textit{w} for enclosed spaces). Burowiec is a
district of Olkusz, a town in the Małopolska region
of southern Poland, southeast of Kraków --- the area
whose proximity to lead and zinc smelting operations
is the book's subject.

\medskip
Where Chapter~1 opened at a funeral and moved through
the present-tense reckoning of a generation's losses,
Chapter~2 announces its coordinates in time and space:
the last day of summer, a specific estate in a specific
town. The chapter will move into the past, into the
children themselves --- before the vanishing, at the
moment just before school resumes and the world of the
estate is still fully inhabited.

%% ---------------------------------------------------------------
\section{Wszystko zaczęło się od huty / It All Started with the Smelter}
%% ---------------------------------------------------------------

\poltext{Co do znikania moich rówieśników, wszystko zaczęło
się od huty. Przez sto pięćdziesiąt lat jej istnienia różnie
ją nazywano: Uthemann, Zakłady Cynkowe Szopienice, blajówa,
Huta Metali Nieżelaznych. Jak zwał, tak zwał. Tu wszystko
zaczynało i kończyło się na hucie.}

\poltrans{As for the disappearing of my contemporaries,
it all started with the smelter. Over a hundred and fifty
years of its existence it was called by various names:
Uthemann, the Szopienice Zinc Works, \textit{blajówa},
the Non-Ferrous Metals Smelter. Whatever it was called,
it was called. Here everything began and ended with the
smelter.}

The chapter opens with a declaration, a list, a proverb,
and an axiom. In four sentences Jędryk names the smelter
as the origin of everything, passes through 150 years of
its history in a list of aliases, dismisses the question
of names with a shrug, and then restates the first claim
with the finality of a closing argument. The book has
found its subject; Chapter~2 announces it without ceremony.

\textbf{Co do znikania moich rówieśników} ---
\textit{as for the disappearing of my contemporaries}.
\textbf{Co do} + genitive: a formal topic-introducer,
\textit{as for, regarding, on the matter of} ---
the register of a deposition or a report. \textbf{Znikania}
--- genitive of \textit{znikanie}, the verbal noun from
\textit{znikać}: \textit{the disappearing}, the process
nominalised. Chapter~1 used \textit{znikały} and
\textit{znikali} as verbs; now the disappearing is a
noun, a phenomenon to be accounted for. \textbf{Moich
rówieśników} --- \textit{of my contemporaries, of my
peers}: genitive plural of \textit{rówieśnik} (from
\textit{równy} --- equal, and \textit{wiek} --- age):
one who is of equal age, a peer. \textbf{Moich} ---
\textit{my} --- the possessive confirms the narrator's
place inside the cohort: not an external observer but
a member of the disappearing generation.

\textbf{wszystko zaczęło się od huty} ---
\textit{it all started with the smelter}. The thesis.
\textbf{Wszystko} --- \textit{everything, it all}.
\textbf{Zaczęło się} --- perfective reflexive past of
\textit{zaczynać się}: \textit{it began, it started}
--- the reflexive \textit{się} makes the beginning
intransitive, self-initiated, as if the thing started
of its own accord. \textbf{Od huty} --- \textit{from
the smelter}: \textit{od} + genitive of origin and
cause. The smelter is the source, the point of
departure, the \textit{od} from which all consequences
flow. The sentence could close the book as easily as
open a chapter.

\textbf{Przez sto pięćdziesiąt lat jej istnienia} ---
\textit{over a hundred and fifty years of its existence}.
\textbf{Przez} + accusative: duration, passage through
time. \textbf{Sto pięćdziesiąt lat} --- one hundred
and fifty years. \textbf{Jej istnienia} --- genitive
of \textit{jej istnienie}: \textit{its existence},
the smelter's span of operation. A century and a half:
the smelter predates the narrator's grandparents,
predates modern Poland, predates two world wars.

\textbf{różnie ją nazywano} --- \textit{it was called
by various names, people called it variously}.
\textbf{Różnie} --- \textit{variously, in different
ways}. \textbf{Ją} --- accusative of \textit{ona}:
\textit{her/it}, referring to \textit{huta} (feminine).
\textbf{Nazywano} --- impersonal passive past of
\textit{nazywać}: \textit{one called, it was called}
--- the unnamed collective of namers, the succession
of owners, regimes, and bureaucracies that renamed
it over 150 years.

\medskip
The list of names that follows is a compressed history
of the region:

\textbf{Uthemann} --- the original founder's name:
Friedrich Uthemann, a German industrialist who
established a zinc-smelting operation in Szopienice
(now a district of Katowice, in Upper Silesia) in
the nineteenth century. The German surname places the
smelter's origins in the industrial colonisation of
Silesia under Prussian rule.

\textbf{Zakłady Cynkowe Szopienice} --- \textit{the
Szopienice Zinc Works}: the interwar or early
communist-era designation, a sober administrative
title. \textbf{Zakłady} --- \textit{works, plant,
establishment} (plural of \textit{zakład}).
\textbf{Cynkowe} --- \textit{zinc} (adjectival form
of \textit{cynk}).

\textbf{blajówa} --- the workers' name, the local
colloquial name, set deliberately among the official
designations. From German \textit{Blei} (lead) +
the Polish suffix \textit{-ówa} (used to form
informal place-names and nicknames, especially
feminine): \textit{the lead-place, the leady one}.
The suffix makes it familiar, almost affectionate ---
the name the neighbourhood used, not the name on
the letterhead. Its presence in this list is the
list's most important entry: it is the only name
that tells you what was actually produced there.

\textbf{Huta Metali Nieżelaznych} --- \textit{the
Non-Ferrous Metals Smelter}: the late communist-era
official designation, bureaucratically precise and
informationally evasive. \textbf{Nieżelaznych} ---
\textit{non-ferrous}: genitive plural of
\textit{nieżelazny} (\textit{nie-} + \textit{żelazny},
iron): not iron, i.e.\ lead, zinc, cadmium. The
name describes the product category without naming
any specific metal. The more official the name, the
less it says.

\textbf{Jak zwał, tak zwał.} --- \textit{Whatever
it was called, it was called. A name's a name.}
A Polish proverb-formula: \textbf{jak\ldots{}
tak\ldots{}} sets up a parallelism of indifference.
\textbf{Zwał} is the past tense third person
singular of \textit{zwać} --- an archaic/literary
verb meaning \textit{to call, to name} --- used
here as the impersonal \textit{one called it}.
The formula means: \textit{it doesn't matter what
you called it; the thing was what it was}. The
proverb dismisses the whole question of naming
that the preceding list raised, and in doing so
makes a point: the names changed (German founder,
zinc works, workers' slang, communist designation),
but the smelter and its emissions did not.

\textbf{Tu wszystko zaczynało i kończyło się na
hucie.} --- \textit{Here everything began and ended
with the smelter.} The chapter's axiom, restating
the opening thesis with added weight. \textbf{Tu}
--- \textit{here}: the estate, the neighbourhood,
the world of these children. \textbf{Zaczynało i
kończyło się} --- imperfective reflexive past:
\textit{began and ended}, the two verbs paired and
balanced, \textit{zaczynać się} (to begin) and
\textit{kończyć się} (to end). \textbf{Na hucie}
--- \textit{at/with the smelter}: locative of
\textit{huta}, the \textit{na} of attachment and
dependence. Everything in this place had its origin
and its terminus in the smelter: employment, economy,
community --- and, the book argues, illness and death.

\medskip
\noindent\textit{Context note:} The smelter in
question --- operating under its various names in
Szopienice and the surrounding Silesian--Małopolska
region --- processed lead and zinc ore for over a
century, emitting lead particulates that settled on
the surrounding estates. The children who grew up
there in the 1970s and 1980s were breathing and
ingesting lead dust at levels now understood to
cause neurological damage, immune dysfunction, and
early death. The \textit{blajówa} --- the lead-place
--- named itself honestly. The official designations
did not.

\medskip
The opening of Chapter~2 performs the same movement
as the close of Chapter~1, but inverted: where
\textit{znikającą klasę} arrived at the end as a
discovery, \textit{wszystko zaczęło się od huty}
is stated at the start as a premise. The chapter
will now go back to show how the premise was lived.

%% ---------------------------------------------------------------
\section{Sierpień 1975 / August 1975}
%% ---------------------------------------------------------------

\poltext{Sierpień 1975 roku był na Śląsku upalny i duszny.}

\poltrans{August of 1975 was hot and stuffy in Silesia.}

One sentence. The chapter steps back in time and sets
the scene in seven words. After the thesis of the
opening paragraph --- the smelter, the names, the
axiom --- the narrative descends into a specific
summer, a specific region, a specific quality of air.

\textbf{Sierpień 1975 roku} --- \textit{August of
1975}. \textbf{Sierpień} --- \textit{August}: the
month whose name derives from \textit{sierp}, the
sickle --- harvest month, the blade that cuts the
summer. \textbf{Roku} --- genitive of \textit{rok}
(year): \textit{of the year 1975}, the standard
Polish construction for specifying a year. The date
is precise: not \textit{kiedyś} (once, at some
unspecified time) but a named year, the summer before
the children's disappearance becomes the book's
subject.

\medskip
\noindent\textit{Vocabulary note:} Polish month names
are almost entirely Slavic in origin, unlike their
Latin counterparts in most Western European languages,
and they describe what is actually happening in the
natural world. \textbf{Sierpień} --- the sickle month,
when grain is cut. \textbf{Styczeń} (January) ---
from \textit{stykać się} (to meet, to touch): the
junction of the year. \textbf{Kwiecień} (April) ---
from \textit{kwiat} (flower): the flowering month.
\textbf{Październik} (October) --- from
\textit{październ}, the stalk left after flax is
stripped. The calendar in Polish is a record of
agricultural labour, each month named for the task
that defined it. \textit{Sierpień} opens this chapter
with a harvesting blade --- quietly apt for a summer
that will turn out to be the last of its kind.

\textbf{był na Śląsku} --- \textit{was in Silesia}.
\textbf{Na Śląsku} --- locative of \textit{Śląsk}
(Silesia): the \textit{na} is obligatory with
\textit{Śląsk}, as with most regional names in Polish
(\textit{na Mazowszu}, \textit{na Podlasiu},
\textit{na Kujawach}) --- the region treated as a
surface or territory one is \textit{upon} rather
than enclosed within. Silesia: the industrial
heartland of southern Poland, the coal and zinc
and lead country, the \textit{blajówa}'s landscape.

\textbf{upalny i duszny} --- \textit{hot and stuffy,
sultry and airless}. Two adjectives, paired without
elaboration. \textbf{Upalny} --- \textit{scorching,
intensely hot} --- from \textit{upał} (heat, intense
heat), stronger than \textit{gorący} (warm, hot):
the heat of a heat wave, the kind that settles on
industrial cities and does not lift. \textbf{Duszny}
--- \textit{stuffy, airless, oppressive, suffocating}
--- from \textit{dusić} (to choke, to stifle, to
suffocate). The word is used for rooms without
ventilation, for crowds, for air that is too thick
to breathe comfortably. In an industrial city whose
smelter has been processing lead and zinc for 150
years, \textit{duszny} is not merely meteorological.
The air is described as oppressive before we are told
what is in it.

\medskip
The sentence performs a classical narrative move ---
the establishing shot, the weather as atmosphere ---
but the word choice is not neutral. \textit{Upalny}
places us in summer heat; \textit{duszny} places us
in air that does not circulate, air that stays close
to the ground, air that is breathed and breathed again.
In the context of a book about lead contamination from
a smelter's emissions, the choice of \textit{duszny}
over, say, \textit{parny} (humid) or \textit{ciepły}
(warm) carries a quiet secondary meaning that will
only become fully legible as the chapter continues.
The children are about to be introduced into this air.
It was August. It was hot. It was hard to breathe.

%% ──────────────────────────────────────────────────────────
\poltext{Do saturatorów w nagrzanym mieście ustawiały się
kolejki spragnionych gazowanej wody z sokiem za złoty piątka
lub samej gruźliczanki za złotówkę.}

\poltrans{At the soda-water machines in the overheated city,
queues of people thirsty for fizzy water with syrup formed
--- five groszy for the flavoured version, or plain
carbonated water on its own for a złoty.}

A sentence that conjures an entire lost world in one image:
the Soviet-era soda machine, the summer queue, the two-tier
pricing of fizz. The heat established in the previous
sentence now has its human consequence.

\textbf{Do saturatorów} --- \textit{to/at the soda machines}.
\textbf{Saturator} --- the pressurised carbonated-water
dispenser that was a fixture of Polish and Soviet public
life from the 1950s through the 1980s: a metal cylinder
with a tap, dispensing chilled sparkling water, usually
with a choice of fruit syrup. The word comes from
\textit{saturować} (to saturate, to carbonate). These
machines stood on street corners, in parks, in train
stations; a single glass was rinsed and reused by every
customer in sequence, a detail of communal hygiene that
was unremarkable at the time. \textbf{Do} + genitive:
movement toward.

\textbf{w nagrzanym mieście} --- \textit{in the
overheated city}. \textbf{Nagrzany} --- passive
participle of \textit{nagrzać} (to heat up, to warm
through): the city has been heated, thoroughly and
completely, by the August sun. Not merely \textit{ciepły}
(warm) or \textit{gorący} (hot) but \textit{nagrzany}
--- heated through like a stone that has been in the
sun all day and radiates warmth into the evening.

\textbf{ustawiały się kolejki spragnionych} ---
\textit{queues of thirsty people formed}.
\textbf{Ustawiać się} --- reflexive imperfective:
\textit{to form up, to arrange oneself in a line} ---
the queue assembles itself, as if by natural process.
\textbf{Kolejki} --- diminutive plural of
\textit{kolejka} (queue, line): the diminutive is
standard usage, not affectionate --- \textit{kolejka}
is simply the word for a queue, smaller and more
ordinary than \textit{kolejność}. \textbf{Spragnionych}
--- genitive plural of \textit{spragniony} (thirsty,
craving): the adjective used as a noun, \textit{the
thirsty ones}. From \textit{pragnąć} (to thirst, to
crave, to desire) --- the same root as in the formal
\textit{pragnę} (I desire, I wish), here reduced to
physical need.

\textbf{gazowanej wody z sokiem za złoty piątka} ---
\textit{fizzy water with syrup for five groszy}.
\textbf{Gazowana woda} --- \textit{carbonated water}
(from \textit{gaz}, gas). \textbf{Z sokiem} ---
\textit{with syrup/juice}: \textit{sok} is both
juice and flavoured syrup; at the saturator, it
meant a splash of coloured fruit syrup. \textbf{Za}
+ accusative: \textit{for} a price. \textbf{Złoty
piątka} --- \textit{five groszy}: \textbf{piątka}
is \textit{a five} (from \textit{pięć}), and
\textbf{złoty} here is the adjective \textit{złoty}
used as a colloquial term for a grosz coin ---
specifically \textit{złoty grosz}, a copper coin.
\textit{Za złoty piątka} means for five of those
copper coins, i.e.\ five groszy. The communist-era
pricing was deliberately minimal; a glass of
flavoured soda water cost almost nothing.

\textbf{lub samej gruźliczanki za złotówkę} ---
\textit{or plain carbonated water on its own for
a złoty}. \textbf{Samej} --- genitive of \textit{sama}
(alone, by itself, plain): the water without syrup,
unaccompanied. \textbf{Gruźliczanka} --- this is
the sentence's most striking word: the popular
name for plain unflavoured carbonated water from
the saturator, from \textit{gruźlica} (tuberculosis).
The name derives from the folk belief, or perhaps
the medical recommendation, that mineral or
carbonated water was beneficial for tuberculosis
sufferers --- the \textit{-anka} suffix forming
a familiar noun. Plain fizzy water was thus
\textit{gruźliczanka}: TB-water, the invalid's
drink, what you got if you couldn't afford the
syrup. \textbf{Za złotówkę} --- \textit{for one
złoty}: one full coin, twice the price of the
flavoured version. The pricing is counterintuitive
until you remember that syrup was cheap and
abundant while plain water required more
carbonation pressure --- or simply that the
economics of the saturator had their own logic.

\medskip
The sentence places the narrative in communist-era
Poland with the precision of a single prop. The
\textit{saturator} on a street corner in a Silesian
town in August 1975, the queue of thirsty people,
the two prices --- five groszy with syrup,
a złoty without --- these are coordinates that
any Polish reader of a certain age will locate
themselves within immediately. The \textit{gruźliczanka}
carries its own small history: a folk name that
preserves a memory of tuberculosis as the great
urban disease of an earlier era, now repurposed
as slang for the cheaper option at a soda machine.
The city is hot. The people are thirsty. They
queue for water that costs almost nothing.

%% ──────────────────────────────────────────────────────────
\poltext{Któregoś dnia wujek Piotr zabrał mnie na Falę.}

\poltrans{One day, Uncle Piotr took me to the Fala.}

The narrator enters the scene in the first person for
the first time in this chapter. A child, an uncle, a
destination. Seven words that shift the narrative from
the general (the summer, the city, the queues) to the
personal and specific.

\textbf{Któregoś dnia} --- \textit{one day, on some
day}. \textbf{Któregoś} --- genitive of the indefinite
pronoun \textit{któryś} (some, a certain one):
\textit{dnia} is the genitive of \textit{dzień} (day),
giving \textit{of some day, on a certain day}. The
construction is the Polish equivalent of the narrative
\textit{one day} --- indefinite, legendary in register,
the way stories begin. Not \textit{w sierpniu} (in
August) or \textit{pewnego dnia} (on a particular day)
but \textit{któregoś dnia}: one day, as children
remember things, without a date attached.

\textbf{wujek Piotr} --- \textit{Uncle Piotr}.
\textbf{Wujek} --- the diminutive-familiar form of
\textit{wuj} (maternal uncle): \textit{wujek} is
warmer, more childlike in register, the way a child
would actually say it. In Polish family address, the
diminutive is standard when speaking to or of uncles,
aunts, and close family friends; \textit{wuj} alone
would sound stiff. The name \textbf{Piotr} is the
Polish form of Peter. The uncle arrives without
introduction or explanation, as figures in childhood
memory do: he is simply there, already known,
requiring no backstory.

\textbf{zabrał mnie} --- \textit{took me, brought me
along}. \textbf{Zabrać} --- perfective of
\textit{zabierać}: to take along, to take with one.
Not \textit{wziął mnie} (took me, in the sense of
picking up) but \textit{zabrał}: the sense of
including someone in a journey, bringing them
into the activity. The child is passenger, beneficiary.

\textbf{na Falę} --- \textit{to the Fala}.
\textbf{Na} + accusative: movement toward a place
treated as a surface or open venue --- the same
\textit{na} as \textit{na Śląsku}, \textit{na
osiedlu}. \textbf{Fala} --- \textit{wave}: a proper
noun here, the name of a place. \textit{Fala} was
a common name for open-air lidos, swimming pools,
and recreation grounds in communist-era Poland ---
the wave, the place of water and summer leisure.
Several towns had a \textit{Fala}; it was a generic
brand name for public aquatic recreation, the kind
of facility every working-class estate aspired to
have nearby.

\medskip
The sentence is a hinge. The chapter has moved from
thesis (the smelter as origin of everything) to
landscape (August heat, the queues) and now into
the narrator's own childhood memory: a specific
uncle, a specific outing, \textit{któregoś dnia}.
The \textit{Fala} --- the wave, the lido, the place
children went in August heat --- is about to become
the scene for something. The word \textit{fala}
also carries its natural meaning alongside the
proper noun: a wave, a surge, a movement that
spreads outward from a point. In a book about
a contamination that spread across a generation,
the name will not stay innocent for long.

%% ──────────────────────────────────────────────────────────
\poltext{Podarował mi czarną maskę do nurkowania.}

\poltrans{He gave me a black diving mask.}

Five words. The uncle gives the child a gift; the
child receives it. The sentence is plain, almost
bare --- and the object it names is precise enough
to do its own work.

\textbf{Podarował mi} --- \textit{he gave me, he
presented me with}. \textbf{Podarować} --- perfective
of \textit{darować}: to give as a gift, to present.
The \textit{po-} prefix marks completion and
generosity. The distinction from the simpler
\textbf{dał} (from \textit{dać}, to give, to hand
over) is worth holding: \textit{dać} is transactional
--- to pass something across --- while
\textit{podarować} is a gift, freely made, with
the full weight of the act behind it. The uncle
didn't hand the child a mask; he gave it to him.
\textbf{Mi} --- dative of \textit{ja}: to me, the
indirect object, the recipient.

\textbf{czarną maskę do nurkowania} --- \textit{a
black diving mask}. \textbf{Czarną maskę} ---
accusative of \textit{czarna maska}: the adjective
\textit{czarny} (black) agreeing with the feminine
\textit{maska} in the accusative. The colour is
specified --- not just any mask but a black one.
Children remember the precise detail: the colour
of the rubber, the smell of it, the specific weight
of a thing that was new and slightly grown-up, a
real piece of equipment rather than a toy. That
\textit{czarną} is the mark of genuine memory
rather than reconstruction. \textbf{Do nurkowania}
--- \textit{for diving}: \textit{do} + genitive of
the verbal noun \textit{nurkowanie} (diving, from
\textit{nurkować} --- to dive, to submerge). The
\textit{do} of purpose: a mask made for going under.

\medskip
The diving mask is the chapter's first object given
directly to the narrator as a child. In the
overheated city of August 1975, taken by his uncle
to the \textit{Fala}, he is given something that
lets him see underwater --- that transforms the
pool from a surface to a depth, that makes the
invisible visible. In a book about a contamination
that was invisible to those living inside it,
the gift of an instrument of seeing-beneath-surfaces
carries a weight that exceeds its five words.
Whether Jędryk intends the symbolic register or
not, the mask will be worth remembering.

%% ──────────────────────────────────────────────────────────
\poltext{Pachniała rozgrzaną gumą.}

\poltrans{It smelled of warm rubber.}

Three words. The child has the mask; now he smells
it. The sentence is pure sensory memory — the kind
that bypasses conscious recollection entirely and
arrives as physical fact.

\textbf{Pachniała} --- \textit{it smelled, it had
the scent of}. \textbf{Pachnieć} --- imperfective:
to smell (of something pleasant or at least
neutral), to have a scent. The imperfective aspect
here signals a continuous, ambient quality: the
mask was not merely smelling of rubber at one
moment but carrying that smell as a property.
Compare \textit{śmierdziała} (it stank) --- the
choice of \textit{pachniała} rather than
\textit{śmierdziała} tells you this is a child's
memory, and the smell is a good one, or at least
a vivid one that the child associated with
excitement and novelty rather than unpleasantness.
The rubber smell of new equipment: an olfactory
marker of something desirable.

\textbf{rozgrzaną gumą} --- \textit{of warm rubber,
of heated rubber}. \textbf{Rozgrzaną} --- the
instrumental of the passive participle
\textit{rozgrzany} (warmed up, heated through),
from \textit{rozgrzać} (to heat up, to warm
through) --- the same root as \textit{nagrzanym}
used of the city two sentences earlier. The August
heat has reached the mask too; it has been sitting
in the sun, or in a bag, and the rubber is warm
to the touch and to the nose. \textbf{Gumą} ---
instrumental of \textit{guma} (rubber, gum): the
smell of the material itself, specific and
unmistakable.

\medskip
\noindent\textit{Grammar note:} \textbf{Pachniała
rozgrzaną gumą} uses the instrumental without a
preposition to express the content of the smell ---
the thing smelled \textit{of} rubber is rendered as
smelling rubber-in-the-instrumental. Polish uses
this construction consistently with verbs of
perception that take an object-of-sensation:
\textit{pachnieć czymś} (to smell of something),
\textit{smakować czymś} (to taste of something).
The instrumental is the case of the instrument
and the medium, and scent here is the medium
through which the rubber makes itself known.

\medskip
The sentence continues the chapter's movement
from the general to the intimate. The heat of
Silesia in August, the queues at the saturator,
the uncle, the outing, the gift --- and now the
smell of the mask in the child's hands. Jędryk
is zooming in, one register at a time, until we
arrive at the body of a specific child holding
a specific object and breathing it in. The word
\textit{rozgrzaną} connecting the mask to the
overheated city (\textit{nagrzanym mieście}) is
not coincidental: everything in this August is
heated through. The air, the city, the rubber.
The child breathes it all.

%% ──────────────────────────────────────────────────────────
\poltext{Rozpychałem się w niej wśród tłumów w basenie
z prawdziwymi sztucznymi falami.}

\poltrans{I elbowed my way through the crowds in it in
the pool with real artificial waves.}

The child is in the pool, masked, pushing through
the crowd, in water that is both genuinely wavy and
mechanically produced. The sentence earns its oxymoron
at the end.

\textbf{Rozpychałem się} --- \textit{I elbowed my way,
I pushed through, I shoved my way}. \textbf{Rozpychać
się} --- imperfective reflexive: to push oneself
through, to make room for oneself by shoving outward.
The \textit{roz-} prefix suggests outward dispersal
--- pushing in all directions to create space.
\textit{Pychać} is a colloquial verb for pushing or
shoving; \textit{rozpychać się} is the action of a
child in a crowded pool who is not going to wait
politely. The imperfective aspect holds him in the
act: he kept pushing, throughout the visit. The
reflexive \textit{się} marks the action as
self-directed and self-propelled.

\textbf{w niej} --- \textit{in it}: \textit{w} +
locative of \textit{ona} (she/it), referring back
to the mask (\textit{maska}, feminine). He is
wearing the mask while doing all this shoving ---
the mask is still on, the child is submerging and
surfacing and pushing through crowds with his new
equipment on his face. The pronoun keeps the mask
present in the sentence without naming it again.

\textbf{wśród tłumów} --- \textit{among crowds,
amid the throngs}. \textbf{Wśród} + genitive:
\textit{among, in the midst of}. \textbf{Tłumów}
--- genitive plural of \textit{tłum} (crowd, mass
of people): not one crowd but \textit{crowds},
plural, the Fala packed on an August day. The
public lido in a Silesian town in the last summer
of term would have been exactly this: heaving,
loud, full of children and their families and the
general press of a hot city seeking water.

\textbf{w basenie z prawdziwymi sztucznymi falami}
--- \textit{in the pool with real artificial waves}.
\textbf{Basenie} --- locative of \textit{basen}:
swimming pool. \textbf{Z prawdziwymi sztucznymi
falami} --- \textit{with real artificial waves}:
the oxymoron the sentence has been building toward.
\textbf{Prawdziwymi} --- instrumental plural of
\textit{prawdziwy} (real, genuine, true).
\textbf{Sztucznymi} --- instrumental plural of
\textit{sztuczny} (artificial, man-made, fake),
from \textit{sztuka} (art, craft): made by human
craft rather than by nature. \textbf{Falami} ---
instrumental plural of \textit{fala} (wave) ---
the pool's name made literal: the \textit{Fala}
had actual waves, mechanically generated.

\medskip
\noindent\textit{On the oxymoron:} \textbf{prawdziwe
sztuczne fale} --- \textit{real artificial waves}
--- is not a contradiction the child finds amusing;
it is simply accurate. The waves are genuinely
present (you can feel them, they knock you off
your feet, they are \textit{prawdziwe} in every
physical sense) and mechanically produced
(\textit{sztuczne}: made by a machine, not by
wind or tide). The child's formulation captures
both truths without irony. It takes an adult
reader to notice the paradox --- and to notice
that \textit{fala} has now appeared three times:
as the pool's name, as the physical wave in the
water, and now explicitly as the thing the pool
was built to produce. The word has earned its
place. A wave that is real and artificial
simultaneously: in a book about lead contamination
that was invisible and undeniable at the same time,
the oxymoron will resonate beyond the pool.

%% ──────────────────────────────────────────────────────────
\poltext{Wracaliśmy z kąpieliska w chorzowskim parku
z przesiadką na rondzie i dalej nagrzaną, czerwoną,
przegubową czterdziestką na Burowiec.}

\poltrans{We were making our way back from the lido
in the Chorzów park, with a change at the roundabout
and then onward by the hot, red, articulated number
forty to Burowiec.}

The journey home, rendered in a single sentence of
accumulating precision. We move from the lido to
the park to the roundabout to the bus number to the
estate, each detail layered onto the last, the whole
route reconstructed from a child's memory with the
specificity that only genuine experience produces.

\textbf{Wracaliśmy} --- \textit{we were making our
way back, we were returning}. \textbf{Wracać} ---
imperfective: the return is in progress, ongoing,
not yet complete. First person plural: the child
and the uncle together, going home. The imperfective
holds them on the journey for the duration of the
sentence.

\textbf{z kąpieliska w chorzowskim parku} ---
\textit{from the lido in the Chorzów park}.
\textbf{Kąpielisko} --- \textit{lido, bathing
place, outdoor swimming facility}: a more formal
word than the colloquial \textit{Fala}, the
official designation for the same place.
\textbf{W chorzowskim parku} --- \textit{in the
Chorzów park}: locative of \textit{chorzowski park}
--- the park in Chorzów, now identified by name.
Chorzów is a city in Upper Silesia, part of the
conurbation that includes Katowice and Szopienice
--- industrial heartland, the territory of the
\textit{blajówa}. The park in question is Wojewódki
Park Kultury i Wypoczynku (the Provincial Park of
Culture and Recreation), one of the largest urban
parks in Europe, built on the site of former
collieries and containing, among much else, a
large outdoor lido. The child travelled from his
estate in Olkusz or its environs to the big park
in Chorzów for this outing --- a significant trip,
an event, the kind of excursion that explains why
the uncle and the diving mask mattered.

\textbf{z przesiadką na rondzie} --- \textit{with a
change at the roundabout}. \textbf{Przesiadka} ---
\textit{transfer, change} (of tram or bus): the
noun of the act of changing vehicles, from
\textit{przesiąść się} (to transfer, to move from
one vehicle to another). \textbf{Na rondzie} ---
\textit{at the roundabout}: locative of
\textit{rondo}. Every city child knows which
rondo is the one --- no further identification
needed.

\textbf{dalej nagrzaną, czerwoną, przegubową
czterdziestką} --- \textit{and then onward by the
hot, red, articulated number forty}. Three
adjectives in sequence before the noun, each one
adding a layer: \textbf{nagrzaną} --- heated
through, the bus warm from the August sun, the
same \textit{nagrzany} as the city and the mask's
rubber; \textbf{czerwoną} --- red, the colour of
the bus, which the child specifies as if to
distinguish it from buses of other colours, or
simply because red is what it was and the memory
is exact; \textbf{przegubową} --- articulated,
from \textit{przegub} (joint, hinge, articulation):
the bendy bus, the kind with a concertina middle
section that swings around corners. The articulated
bus was a feature of communist-era Polish city
transport, longer than a standard bus, able to
carry more passengers, beloved by children for
the moving floor-section in the middle joint.
\textbf{Czterdziestką} --- instrumental of
\textit{czterdziestka} (a forty, the number-forty):
the bus route number used as a noun, the way
Poles still refer to bus and tram lines.
\textbf{Na Burowiec} --- \textit{to Burowiec}:
accusative of direction, the estate that is the
chapter's second coordinate and the narrator's
home.

\medskip
The sentence is a feat of affectionate topography.
The child does not say \textit{we went home by
bus}; he reconstructs the exact route ---
kąpielisko, park, rondo, czterdziestka,
Burowiec --- with the unselfconscious precision
of someone for whom these coordinates were once
the entire known world. The three adjectives
applied to the bus (\textit{nagrzana, czerwona,
przegubowa}) are a child's catalogue of its
notable properties: warm, red, bendy. None of
them is the route number or the destination;
all of them are what made this particular bus
this bus, distinguishable from all others.
\textit{Nagrzaną} connects the bus to the
heated city and the warm rubber mask, completing
the chapter's thermal motif: in August 1975,
everything is hot, everything is suffused with
the same Silesian summer heat.

%% ──────────────────────────────────────────────────────────
\poltext{Przepełnione autobusy żłobiły koleiny
w rozmiękłym od upałów asfalcie.}

\poltrans{Overloaded buses were cutting ruts into the
asphalt softened by the heat.}

The sentence pulls back from the child on the bus
to the street outside, and the image it delivers
is quietly extraordinary: the road itself yielding
under the weight of the buses, the summer heat
having softened the asphalt to the point where
vehicles leave permanent marks. The city is not
merely hot; it is being physically deformed by
the heat.

\textbf{Przepełnione autobusy} --- \textit{overloaded
buses, overfull buses}. \textbf{Przepełniony} ---
passive participle of \textit{przepełnić} (to
overfill, to fill beyond capacity): the \textit{prze-}
prefix signals excess, going beyond a limit.
The buses the child rode home were packed --- a
detail confirmed by \textit{rozpychałem się wśród
tłumów} in the pool, now extended to the transport:
the crowds are everywhere, pressing into every
available space. The plural \textbf{autobusy}
generalises: not just the number forty but all
the buses on all the routes through the city this
August.

\textbf{żłobiły koleiny} --- \textit{were cutting
ruts, were gouging grooves}. \textbf{Żłobić} ---
imperfective: to gouge, to cut, to hollow out a
channel. The word is related to \textit{żłób}
(a trough, a manger, a channel cut into stone or
wood) --- the action of repeated pressure carving
a path into a surface. \textbf{Koleiny} ---
\textit{ruts, wheel-tracks}: the grooves left by
repeated passage of heavy vehicles, the tracks
that become permanent features of a road surface.
\textbf{Żłobiły koleiny} together: the buses are
not merely driving over the asphalt but physically
reshaping it, cutting channels into it with their
weight.

\textbf{w rozmiękłym od upałów asfalcie} ---
\textit{in the asphalt softened by the heat}.
\textbf{Rozmiękły} --- passive participle of
\textit{rozmiękć} (to soften, to become soft):
the \textit{roz-} prefix marking a process of
dissolution, spreading softness through a material.
\textbf{Od upałów} --- \textit{from the heat,
because of the heat}: \textit{od} + genitive of
cause; \textit{upałów} is the genitive plural of
\textit{upał} (intense heat, heat wave) --- the
same root as \textit{upalny} in the chapter's
second sentence. The heat described as a quality
of August has now become an active agent: it has
softened the asphalt. \textbf{Asfalcie} ---
locative of \textit{asfalt}: in the asphalt, the
ruts being cut into it.

\medskip
The sentence completes the chapter's thermal
argument. The progression has been: the air is
hot and stuffy (\textit{upalny i duszny}); the
city is overheated (\textit{nagrzane miasto});
the mask is warm (\textit{rozgrzana guma}); the
bus is hot (\textit{nagrzana czterdziestka});
and now the road itself has softened and yields
to the weight passing over it. The heat is not
background weather but a physical force reshaping
the environment. \textbf{Żłobiły} --- the buses
are writing their passage into the city's surface,
leaving permanent marks. In a chapter about a
contamination that also left permanent marks ---
in bodies, in a generation, in a class that kept
vanishing --- the image of heavy things pressing
grooves into softened material carries an
undertow. The city is being marked. The children
in it are being marked. The ruts will remain long
after the summer ends.

%% ──────────────────────────────────────────────────────────
\poltext{Po zachodzie, kiedy słońce majestatycznie skryło
się za Zakładem Budowy Szybów, większość okien na osiedlu
przy ulicach Przedwiośnia i Pogodnej było pootwieranych
na oścież.}

\poltrans{After sunset, when the sun had majestically
hidden itself behind the Shaft Construction Works, most
of the windows on the estate on Przedwiośnie and Pogodna
streets were flung wide open.}

The day ends. The sentence moves from the industrial
skyline to the domestic detail in one arc, and the
street names it delivers are among the most quietly
pointed in the chapter.

\textbf{Po zachodzie} --- \textit{after sunset, after
the setting}. \textbf{Zachód} --- \textit{west; also
sunset, the setting of the sun} --- the direction and
the event sharing a word, as in many languages. The
genitive after \textit{po} marks time: \textit{after
the sunset}. The day established at the chapter's
opening --- August, hot, Silesian --- is now closing.

\textbf{kiedy słońce majestatycznie skryło się za
Zakładem Budowy Szybów} --- \textit{when the sun had
majestically hidden itself behind the Shaft Construction
Works}. \textbf{Majestatycznie} --- \textit{majestically,
with grandeur} --- a surprising adverb for a sunset
behind an industrial facility. The word belongs to
descriptions of mountain panoramas, cathedral silhouettes,
natural spectacle --- applied here to the sun disappearing
behind a factory, it confers ironic dignity on the
industrial skyline, treating it as the only grandeur
available. \textbf{Skryło się} --- perfective reflexive:
\textit{it hid itself}, the sun performing a deliberate
concealment. \textbf{Za Zakładem Budowy Szybów} ---
\textit{behind the Shaft Construction Works}: instrumental
of \textit{Zakład Budowy Szybów}. \textbf{Zakład} ---
works, plant, establishment. \textbf{Budowy} --- genitive
of \textit{budowa} (construction, building). \textbf{Szybów}
--- genitive plural of \textit{szyb}: a mine shaft, a
vertical tunnel sunk into the earth for extracting ore.
The Zakład Budowy Szybów was a specialised industrial
enterprise that constructed mine shafts --- another
institution of the Silesian extraction economy, neighbour
to the smelters and collieries that defined the region.
The sun sets behind the shaft-builders. This is the
local horizon.

\textbf{większość okien na osiedlu przy ulicach
Przedwiośnia i Pogodnej} --- \textit{most of the windows
on the estate on Przedwiośnie and Pogodna streets}.
\textbf{Większość} + genitive: \textit{the majority of}.
\textbf{Okien} --- genitive plural of \textit{okno}
(window). \textbf{Na osiedlu} --- \textit{on the estate}:
the familiar locative. \textbf{Przy ulicach} ---
\textit{on/at the streets}: \textit{przy} + locative,
indicating adjacency. And the street names:
\textbf{Przedwiośnia} --- genitive of \textit{Przedwiośnie}:
\textit{Early Spring, Pre-Spring}, the transitional
season before spring proper arrives, the thaw before
the warmth. \textbf{Pogodnej} --- genitive of
\textit{Pogodna}: \textit{Fair, Clear, Sunny} ---
from \textit{pogoda} (weather; also: good weather,
fair weather, sunshine). The streets of the estate
are named for optimistic meteorological and seasonal
states: Early Spring, Sunny. They were named, in the
way communist-era housing estates were named, for
aspiration and lightness --- and the children who
lived on them were being poisoned by the smelter
two streets over.

\textbf{było pootwieranych na oścież} --- \textit{were
flung wide open, were opened to their full extent}.
\textbf{Pootwieranych} --- this form rewards attention:
it is the genitive plural of the passive participle
\textit{pootwarty}, from the prefixed perfective
\textit{pootworzyć}: the \textit{po-} prefix applied
to \textit{otworzyć} (to open) signals a distributed
or thoroughgoing action --- not just opened but opened
comprehensively, many windows, all the way.
\textbf{Na oścież} --- \textit{wide open, to the full
extent}: an idiom from \textit{ość} (fish-bone,
spine), suggesting opening to the very frame, as far
as a thing will go. \textit{Otworzyć na oścież} ---
to fling wide open --- is the standard collocation.
The windows of the estate are not merely ajar; they
are thrown as wide as they will go, the whole building
breathing the evening air in after the heat of the day.

\medskip
The sentence closes the chapter's first day with
an image of collective opening: after the long hot
August, after the sun has set behind the shaft works,
the estate opens its windows to the outside. In 1975
on Przedwiośnie and Pogodna streets, this was simply
what you did when the evening finally brought relief
from the heat. The air that came in through those
windows --- the air of Silesia, neighbour to the
\textit{blajówa} and the Zakład Budowy Szybów ---
is the air the book is about. The windows are open.
Everything is let in.

%% ──────────────────────────────────────────────────────────
\poltext{Mieszkańcy przywykli do sinego, słodkawego,
gryzącego dymu, który nadciągał znad huty i zawisał
nisko, otulając domy smrodliwym kokonem. Najgorzej
było nocami.}

\poltrans{The residents had grown accustomed to the
blue-grey, sweetish, biting smoke that drifted in
from above the smelter and hung low, wrapping the
houses in a foul-smelling cocoon. It was worst at
night.}

The windows opened to let in the evening air --- and
this is what came through them. The chapter has been
building to this sentence since its first word. The
smoke is now named, described in three adjectives,
given a trajectory and a behaviour, and judged. The
two sentences together are the chapter's axis.

\textbf{Mieszkańcy przywykli} --- \textit{the
residents had grown accustomed, had got used to}.
\textbf{Przywyknąć} --- perfective: to become
accustomed, to habituate oneself. The perfective
marks a completed process: the habituation is done,
established, no longer in progress. \textbf{Mieszkańcy}
--- \textit{residents, inhabitants} --- from
\textit{mieszkać} (to live, to dwell). The subject
is collective and anonymous --- not the narrator's
family specifically but everyone on Przedwiośnie
and Pogodna, all the windows, all the people behind
them.

\textbf{do sinego, słodkawego, gryzącego dymu} ---
\textit{to the blue-grey, sweetish, biting smoke}.
Three adjectives, each adding a sensory dimension:

\textbf{Sinego} --- genitive of \textit{siny}:
a dark blue-grey, the colour of bruised flesh or
leaden sky, the colour of metal-smoke. \textit{Siny}
is a specifically morbid shade --- it describes
cyanosis, the bluish discolouration of skin starved
of oxygen, and the lips of the very cold or the
nearly dead. Applied to smoke, it carries that
register with it. The smoke is the colour of what
it does.

\textbf{Słodkawego} --- genitive of \textit{słodkawy}:
\textit{sweetish, faintly sweet} --- the \textit{-awy}
suffix indicating an approximate or impure quality
of the base adjective \textit{słodki} (sweet). Lead
and zinc smelting produces fumes with a characteristic
faintly sweet metallic odour --- the sweetness is
the chemistry of the process, the smell of metal
being reduced. \textit{Słodkawy} is not
pleasant-sweet but uncanny-sweet, the sweetness
of something that should not be sweet.

\textbf{Gryzącego} --- genitive of \textit{gryzący}:
\textit{biting, acrid, stinging} --- present active
participle of \textit{gryźć} (to bite, to gnaw):
the smoke that bites the throat and eyes. The three
adjectives move from visual (\textit{siny}) to
olfactory (\textit{słodkawy}) to physical sensation
(\textit{gryzący}): smoke that could be seen,
smelled, and felt.

\textbf{który nadciągał znad huty} --- \textit{that
drifted in from above the smelter}. \textbf{Nadciągać}
--- imperfective: to approach, to drift toward, to
roll in. \textbf{Znad} --- compound preposition:
\textit{z} (from) + \textit{nad} (above): from the
airspace over the smelter, descending from where
the stacks discharged it.

\textbf{i zawisał nisko} --- \textit{and hung low}.
\textbf{Zawisać} --- imperfective: to hang, to be
suspended. \textbf{Nisko} --- \textit{low}: not
dispersing upward and away but settling at the
level of windows, doors, and children playing
outside. Lead particulates are heavy; the physics
of the smoke's behaviour is in this word.

\textbf{otulając domy smrodliwym kokonem} ---
\textit{wrapping the houses in a foul-smelling
cocoon}. \textbf{Otulać} --- to wrap, to envelop,
to swaddle --- a word used for tenderly wrapping
a child in blankets. Here applied to smoke, the
tenderness is made sinister by context.
\textbf{Smrodliwym} --- \textit{foul-smelling,
stinking}: from \textit{smród} (stench).
\textbf{Kokon} --- \textit{cocoon}: the silkworm's
casing, the structure that encloses and transforms
what is inside it. The estate is wrapped as if
tenderly. What emerges from a cocoon is changed
by having been inside it.

\medskip

\textbf{Najgorzej było nocami.} --- \textit{It was
worst at night.} Five words, and they close a trap.
\textbf{Najgorzej} --- superlative of \textit{źle}
(badly): \textit{worst}. \textbf{Nocami} ---
instrumental plural of \textit{noc} (night):
\textit{at nights}, the instrumental of repeated
time. The windows were open; the smoke hung low;
it was worst at night. The logic is inexorable:
in the August heat the residents opened their
windows wide (\textit{na oścież}) for relief, and
the smoke that wrapped the estate came in. The
children slept inside the cocoon.

%% ──────────────────────────────────────────────────────────
\poltext{Wszyscy wiedzieli, że wtedy ściągano filtry
z kominów, by nadgonić normy produkcji ołowiu.}

\poltrans{Everyone knew that at those times the filters
were pulled from the chimneys in order to catch up
with the lead production quotas.}

The sentence everyone on the estate knew. It names
the mechanism — not weather, not accident, not
mystery — and it names the word the book has been
holding back since its title: \textbf{ołów}. Lead.

\textbf{Wszyscy wiedzieli} --- \textit{everyone
knew}. \textbf{Wszyscy} --- \textit{everyone, all
of them}: the same word as \textit{wszyscy są do
siebie podobni} in Chapter~1, Hanka's dismissal
of the men \textit{pod sklepem}. Here it means
something different: not \textit{they all look
alike} but \textit{the knowledge was universal,
shared, unspoken but common}. \textbf{Wiedzieli}
--- imperfective past: they knew, continuously,
habitually. This was not a revelation; it was
common knowledge, part of the estate's ambient
understanding of its own situation.

\textbf{że wtedy ściągano filtry z kominów} ---
\textit{that at those times the filters were pulled
from the chimneys}. \textbf{Wtedy} ---
\textit{then, at those times}: correlating with
\textit{nocami} from the previous sentence ---
at night, then, that was when. \textbf{Ściągano}
--- impersonal passive past of \textit{ściągać}
(to pull off, to remove, to take down): \textit{they
were pulled, someone pulled them}, the agent
unnamed and unnecessary. The filters did not
remove themselves; someone took them off. The
impersonal construction is the grammar of
institutional accountability that no one accepts.
\textbf{Filtry z kominów} --- \textit{filters from
the chimneys}: \textbf{filtr} is the industrial
filter fitted to a smokestack to capture
particulates before they are released into the
air. \textbf{Komin} --- chimney, smokestack ---
from the same root as Italian \textit{camino}.
The filters on the smelter's stacks captured lead
dust and zinc oxide; removing them meant
everything the combustion process produced went
directly into the air above the estate.

\textbf{by nadgonić normy produkcji ołowiu} ---
\textit{in order to catch up with the lead
production quotas}. \textbf{By} + infinitive:
the construction of purpose, \textit{in order to}.
\textbf{Nadgonić} --- perfective of \textit{nadganiać}:
to catch up, to make up ground, to close a gap
--- the \textit{nad-} prefix suggesting catching
up to something ahead of you. \textbf{Normy
produkcji} --- \textit{production quotas, production
norms}: \textbf{norma} (norm, quota, standard) is
the communist-era term for the production target
set by the central plan, the number a factory
was required to meet. \textbf{Ołowiu} --- genitive
of \textbf{ołów}: \textit{lead}. The word the book
is named for, arriving here for the first time
in the chapter text, in the genitive case, as the
object of an industrial process. Lead production
quotas. The plan required a certain tonnage of
lead; the filters slowed the process; removing
the filters at night, when inspectors were absent
and residents asleep, allowed the quota to be met.

\medskip
The sentence is the chapter's most explicit
statement of culpability, and it delivers it in
the flattest possible register: \textit{everyone
knew}. Not the authorities, not the workers, not
a whistleblower --- \textit{wszyscy}: the residents
of Przedwiośnie and Pogodna, the people whose
windows were open at night, knew that the worst
of the smoke came when the filters were removed
for the sake of the production plan. They knew,
and they lived with it, because what else does
one do? The communist production quota was not
negotiable. The estate was not moveable. The
windows were open because it was August and the
night air was the only relief from the heat.

\medskip
\noindent\textit{On} \textbf{ołów}: this is the
first appearance of the word in Chapter~2, and
only the second in the book (after the title).
The book is called \textit{Ołowiane dzieci} ---
\textit{Lead Children} --- and the subtitle
\textit{Zapomniana epidemia} --- \textit{The
Forgotten Epidemic} --- has been waiting for
this word to appear in the text. It arrives in
the genitive, subordinated to \textit{normy
produkcji}: not lead as a poison, not lead as
a substance that enters children's bodies, but
lead as a production commodity, a quota to be
met, a tonnage on a plan. The bureaucratic
framing is the point. The children are
\textit{ołowiane} --- leaden, made of lead ---
because someone needed to meet the norm.

%% ──────────────────────────────────────────────────────────
\poltext{Trzeba było wykonać plan; huta musiała czynem
uczcić kolejny zjazd partii i wnieść swój wkład w rozwój
kraju, który aspirował do miana wysokorozwiniętego dzięki
najcięższym gałęziom przemysłu: górnictwu i hutnictwu.}

\poltrans{The plan had to be fulfilled; the smelter was
obliged to honour the next party congress in deed and
contribute its share to the development of a country
that aspired to the status of highly developed thanks
to the heaviest branches of industry: mining and
metallurgy.}

The sentence delivers the ideological machinery behind
the filter removal. It is written in the language of
the system it describes --- the stilted, ceremonial
rhetoric of communist economic planning --- and the
gap between that language and the sleeping children
with their windows open is the sentence's entire
meaning.

\textbf{Trzeba było wykonać plan} --- \textit{the plan
had to be fulfilled, it was necessary to complete the
plan}. \textbf{Trzeba było} --- impersonal past of
\textit{trzeba} (it is necessary, one must): the
obligation stated without an agent. Not \textit{the
smelter managers decided} or \textit{the party
ordered} but the impersonal \textit{trzeba było} ---
it was necessary, as if the necessity were a law of
nature. \textbf{Wykonać plan} --- \textit{to fulfil
the plan, to complete the plan}: \textbf{wykonać}
(perfective of \textit{wykonywać}: to carry out, to
execute, to fulfil) + \textbf{plan} --- the central
plan, the production target handed down from the
planning authority. \textit{Wykonać plan} was the
fundamental obligation of every Soviet-bloc industrial
enterprise: the plan was not a suggestion but a
binding norm.

\textbf{huta musiała czynem uczcić kolejny zjazd
partii} --- \textit{the smelter was obliged to honour
the next party congress in deed}. \textbf{Musiała}
--- \textit{had to, was obliged to}: past tense of
\textit{musieć}, the modal of compulsion.
\textbf{Czynem} --- instrumental of \textit{czyn}
(deed, act, action): \textit{by means of a deed},
the instrumental of manner. \textbf{Uczcić} ---
perfective of \textit{czcić}: to honour, to
commemorate, to pay tribute to. \textbf{Kolejny
zjazd partii} --- \textit{the next party congress}:
\textbf{zjazd} (congress, convention, gathering)
of \textbf{partia} (the Party, the Polish United
Workers' Party, PZPR). Party congresses were
occasions for which industrial enterprises were
expected to demonstrate their loyalty and
productivity by overfulfilling their quotas ---
\textit{uczcić czynem}: to honour not with words
but with production. The phrase is saturated in
the rhetorical register of communist official
discourse, delivered here with perfect deadpan.

\textbf{i wnieść swój wkład w rozwój kraju} ---
\textit{and contribute its share to the development
of the country}. \textbf{Wnieść wkład} ---
\textit{to make a contribution, to bring one's
share}: a fixed phrase of official rhetoric.
\textbf{W rozwój kraju} --- \textit{to the
development of the country}: \textbf{rozwój}
(development, growth) + \textbf{kraju} (genitive
of \textit{kraj}, country). The language is
entirely that of the planning documents and party
speeches of the era.

\textbf{który aspirował do miana wysokorozwiniętego}
--- \textit{that aspired to the status of highly
developed}. \textbf{Aspirować do miana} ---
\textit{to aspire to the title/status of}:
\textbf{miano} (name, title, designation) is the
object of the aspiration. \textbf{Wysokorozwinięty}
--- \textit{highly developed}: a single compound
adjective, the Polish calque of the economic
category \textit{wysoko rozwinięty kraj}
(highly developed country), the status communist
Poland officially claimed to be approaching.
The aspiration is the key word: not \textit{był}
(was) but \textit{aspirował} (aspired to be).

\textbf{dzięki najcięższym gałęziom przemysłu:
górnictwu i hutnictwu} --- \textit{thanks to the
heaviest branches of industry: mining and
metallurgy}. \textbf{Dzięki} + dative: \textit{thanks
to, by means of}. \textbf{Najcięższym gałęziom}
--- \textit{the heaviest branches}: superlative
of \textit{ciężki} (heavy), dative plural agreeing
with \textbf{gałęziom} (branches, from
\textit{gałąź}). \textit{Ciężki przemysł} ---
\textit{heavy industry} --- was the ideological
cornerstone of Soviet-bloc economic policy: coal,
steel, metals, the industries that produced the
material base of socialist construction.
\textbf{Górnictwu i hutnictwu} --- \textit{mining
and metallurgy/smelting}: dative of \textit{górnictwo}
(mining, from \textit{góra}, mountain) and
\textit{hutnictwo} (metallurgy, smelting, from
\textit{huta}).

\medskip
The sentence is a sustained act of mimicry. Jędryk
reproduces the official language of the system ---
\textit{wykonać plan, uczcić czynem, wnieść wkład,
rozwój kraju, wysokorozwinięty} --- with such
fidelity that it reads like a quotation from a
party newspaper. The mimicry is the critique: the
same sentence that explains why the filters were
removed at night also explains, in the system's
own words, why that was not only permitted but
required. The plan had to be fulfilled. The party
congress had to be honoured. The country had to
develop. The filters came off. The children slept
with their windows open. The language of
development and honour and contribution does not
mention what it costs, or who pays.

\medskip
\noindent\textit{On} \textbf{gałąź}: the word for
a branch of industry is the same word as a branch
of a tree --- \textit{gałąź} (branch, bough). The
\textit{najcięższe gałęzie przemysłu} --- the
heaviest branches --- are literally branches heavy
enough to bring the tree down. The metaphor is
available in the language; whether Jędryk activates
it consciously or not, it is there.

%% ──────────────────────────────────────────────────────────
\poltext{Polska miała rosnąć w siłę, a ludzie żyć dostatniej.}

\poltrans{Poland was to grow in strength, and people were
to live more prosperously.}

The slogan. Ten words that were the stated premise of
the entire system — the promise that justified the
plan, the quotas, the filters removed at night, the
smoke in the cocoon. Jędryk does not attribute it to
anyone; he does not need to. Every Polish reader of a
certain age knows exactly where this sentence comes from.

\textbf{Polska miała rosnąć w siłę} --- \textit{Poland
was to grow in strength}. \textbf{Miała} + infinitive:
the construction of stated intention or obligation,
\textit{was to, was supposed to, was meant to} ---
not \textit{grew} (which would be a claim about reality)
but \textit{was to grow} (a promise, a programme, a
directive). \textbf{Rosnąć w siłę} --- \textit{to grow
in strength}: a fixed phrase, the standard formulation
of national strengthening in communist rhetoric.
\textbf{Siłę} --- accusative of \textit{siła}
(strength, force, power).

\textbf{a ludzie żyć dostatniej} --- \textit{and people
were to live more prosperously}. The second half of
the promise, with the same \textit{miała} understood
from the first clause: \textit{a ludzie} [mieli]
\textit{żyć dostatniej}. \textbf{A} --- the
coordinating conjunction of contrast-and-addition:
and, while, as well as. \textbf{Ludzie} ---
\textit{people, the people}: the collective subject.
\textbf{Żyć dostatniej} --- \textit{to live more
prosperously}: \textbf{dostatniej} is the comparative
of \textbf{dostatnio} (prosperously, in comfort,
with sufficiency), from \textbf{dostatek}
(sufficiency, plenty, prosperity). Not \textit{bogato}
(richly) or \textit{luksusowo} (luxuriously) but
\textit{dostatnio} --- comfortably, with enough:
the modest, realistic-sounding promise of the
welfare state.

\medskip
\noindent\textit{Context note:} \textbf{Polska miała
rosnąć w siłę, a ludzie żyć dostatniej} is the direct
echo of the Gierek-era slogan \textit{Polska rośnie
w siłę, a ludzie żyją dostatniej} --- \textit{Poland
is growing in strength, and people are living more
prosperously} --- the official formulation of Edward
Gierek's programme for the 1970s, when Poland borrowed
massively from the West to finance industrial
expansion and raise living standards. By shifting
from the present tense of the slogan (\textit{rośnie,
żyją}) to the modal \textit{miała\ldots{} żyć} ---
\textit{was to\ldots{} was to live} --- Jędryk
performs the entire historical verdict in a single
grammatical tense-shift. The slogan claimed it was
happening; the modal says it was supposed to happen.
The difference between those two is the 1980s,
the debt crisis, Solidarity, martial law, and a
generation of children with lead in their blood.
The filters came off so that Poland could grow in
strength. The people of Przedwiośnie and Pogodna
opened their windows so that they could live
more prosperously.

%% ──────────────────────────────────────────────────────────
\poltext{Z tego powodu nocą nad Szopienicami, Borkami,
Burowcem, Dąbrówką zalegał trujący dym. Aż zamykano okna.}

\poltrans{For this reason, at night above Szopienice,
Borki, Burowiec, Dąbrówka, poisonous smoke settled.
Until people closed their windows.}

The consequence clause. The slogan has been delivered;
now the slogan's price is named. The smoke is no longer
\textit{siny, słodkawy, gryzący} --- it is \textbf{trujący}:
poisonous. And the sentence that follows is three words
that reverse everything.

\textbf{Z tego powodu} --- \textit{for this reason,
because of this, on account of this}. \textbf{Z powodu}
+ genitive: the preposition of cause. \textbf{Tego}
--- genitive of \textit{to} (this): referring directly
back to the preceding sentence --- Poland was to grow
in strength, and for \textit{that reason}, the smoke
settled at night. The causal link is stated without
irony; the irony is structural.

\textbf{nocą} --- \textit{at night}: instrumental
singular of \textit{noc}, the same form as
\textit{nocami} (at nights, plural) two sentences
earlier. The smoke comes at night because that is
when the filters are off; \textit{nocami} established
the pattern, \textit{nocą} specifies this night,
these neighbourhoods, this settling.

\textbf{nad Szopienicami, Borkami, Burowcem, Dąbrówką}
--- \textit{above Szopienice, Borki, Burowiec,
Dąbrówka}: instrumental plural and singular forms of
place names after \textit{nad} (above, over).
Four neighbourhoods named in sequence, the smoke's
geography mapped in a list. \textbf{Szopienice} ---
the district where the smelter stood, the home of
the \textit{blajówa}. \textbf{Borki} --- a residential
district adjacent to the smelter. \textbf{Burowiec}
--- the narrator's estate, named in the chapter title,
here placed third in the smoke's itinerary.
\textbf{Dąbrówka} --- another district in the same
industrial constellation. The list performs the reach
of the contamination: not one street, not one estate,
but a chain of neighbourhoods, each one downwind of
the same source.

\textbf{zalegał trujący dym} --- \textit{poisonous
smoke settled, toxic fumes lay}. \textbf{Zalegać}
--- imperfective: to settle, to lie over, to fill
and remain --- the verb of something that spreads
across a surface and stays. \textbf{Trujący} ---
present active participle of \textit{truć} (to
poison, to intoxicate): \textit{poisoning, toxic,
that which poisons}. This is the chapter's first
use of the word of poison directly applied to the
smoke. Earlier it was \textit{siny, słodkawy,
gryzący} (blue-grey, sweetish, biting) ---
descriptive, sensory. Now it is \textbf{trujący}:
the participial adjective that names the function.
The smoke was always poisoning; only now does the
text say so.

\medskip

\textbf{Aż zamykano okna.} --- \textit{Until people
closed their windows. So bad that windows were
closed.} Three words, and they land as the
chapter's most devastating sentence so far.
\textbf{Aż} --- \textit{until, so much that,
to the point that}: a particle of extreme
consequence, expressing that a threshold was
reached. \textbf{Zamykano} --- impersonal passive
past of \textit{zamykać} (to close): \textit{they
were closed, one closed them}, no agent named.
\textbf{Okna} --- \textit{windows}.

The reversal is total. The previous passage
established that \textit{większość okien była
pootwieranych na oścież} --- most of the windows
were flung wide open --- after sunset, for relief
from the August heat. Now: \textit{aż zamykano
okna}. The smoke was bad enough that the windows
were closed. The same windows that were opened
for air had to be shut against the air. The
people chose between the August heat and the
poisonous smoke and sometimes the smoke won.

\textbf{Aż} is doing everything here. It is not
\textit{więc zamykano} (so they closed them) or
\textit{dlatego zamykano} (therefore they closed
them) but \textit{aż} --- \textit{to the point
that, so extreme that}. The closing of a window
is presented as an extreme outcome, a last resort,
the measure taken when endurance was exhausted.
In the August heat of a flat with no ventilation,
closing the windows was a significant sacrifice.
The smoke made it necessary.

\medskip
With these two sentences the chapter's argument
is complete in miniature: the slogan
(\textit{Polska miała rosnąć w siłę}), the
mechanism (\textit{z tego powodu}), the
consequence (\textit{zalegał trujący dym}), the
human response (\textit{aż zamykano okna}).
Production, poison, endurance, retreat. The
whole system in four sentences.
